---
output: github_document
---

```{r options, echo=FALSE}
library(knitr)
opts_chunk$set(warning=FALSE)
```

[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Build Status](https://travis-ci.org/adibender/pammtools.svg?branch=master)](https://travis-ci.org/adibender/pammtools)
[![Build Status](https://ci.appveyor.com/api/projects/status/github/adibender/pammtools?branch=master&svg=true)](https://ci.appveyor.com/project/adibender/pammtools/branch/master)
[![codecov.io](https://codecov.io/github/adibender/pammtools/coverage.svg?branch=master)](https://codecov.io/github/adibender/pammtools/branch/master)
[![MIT license](http://img.shields.io/badge/license-MIT-brightgreen.svg)](http://opensource.org/licenses/MIT)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.1048832.svg)](https://doi.org/10.5281/zenodo.1048832)

# `pammtools`: Piece-Wise Exponential Additive Mixed Modeling Tools


### Installation

Install from GitHub using

```{r eval = FALSE}
devtools::install_github("adibender/pammtools")
```

### Overview
`pammtools` offers various utilities and convenience functions that
facilitate working with Piece-wise exponential Additive Mixed Models (PAMMs).

To get started, see the [Articles](articles/index.html) section.

For a tutorial-like introduction to PAMMs see:

  - Andreas Bender, Andreas Groll, and Fabian Scheipl. 2018. “A Generalized Additive Model Approach to Time-to-Event Analysis.” Statistical Modelling. https://doi.org/10.1177/1471082X17748083.


A general framework for the representation and estimation of cumulative effects
(or exposure-lag-response associations) is described in:

- Andreas Bender, Fabian Scheipl, Wolfgang Hartl, Andrew G Day, Helmut Küchenhoff; "Penalized estimation of complex, non-linear exposure-lag-response associations", Biostatistics, , kxy003, https://doi.org/10.1093/biostatistics/kxy003
