---
title: "Baseline Hazard"
author: "Andreas Bender"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{baseline}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  fig.align = "center"
)
```


```{r, message = FALSE}
library(pam)
library(survival)
library(ggplot2)
theme_set(theme_bw())
library(magrittr)
library(dplyr)
library(mgcv)
```


Here we briefly demonstrate how to fit and visualize a simple 
baseline model using the **`pam`** package. 
We illustrate the procedure using censored survival data simulated from the Weibull 
distribution $T \sim WB(1.5, 2)$: 

```{r}
set.seed(300117)
tdf <- data.frame(time = rweibull(100, 1.5, 2)) %>% 
  mutate(
    time  = round(time, 3),
    status = time > rweibull(100, 1.5, 2))
head(tdf)
```

The below graph depicts the estimated cumulative hazard using the 
Nelson-Aalen estimator:.

```{r, fig.width = 4.5, fig.height = 4.5}
base.df <- basehaz(coxph(Surv(time, status)~1, data = tdf))
ggplot(base.df, aes(x = time, y = hazard)) + 
  geom_step() + 
  ylab(expression(Lambda(t))) + xlab("t") + 
  ggtitle("Nelson-Aalen estimate of the cumulative hazard")
```


## Data transformation 
First we need to bring the data in a suitable format 
 (see `vignette("data-transformation", package = "pam")`).

```{r}
# Use unique event times as interval break points
ped <- split_data(Surv(time, status)~., data = tdf, id = "id")
head(ped)
```

## Fit the baseline piece-wise exponential model (PEM)
```{r}
pem <- gam(ped_status ~ interval, data = ped, offset = offset, family = poisson())
```

Extract cumulative baseline estimate:

```{r}
int.df <- int_info(ped)
head(int.df)
int.df %<>% add_cumhazard(pem) %>% 
  left_join(base.df, by = c("tend" = "time"))
head(int.df)
```

Visualize the PEM estimate:

```{r, fig.width = 4.5, fig.height = 4.5}
ggplot(int.df, aes(x = tend)) + 
  geom_line(aes(y = cumhazard, col = "PEM")) + 
  geom_step(aes(y = hazard, col = "Nelson-Aalen"), lty = 2) + 
  scale_color_brewer(name = "Method", palette = "Set2") + 
  theme(legend.position = "bottom") + 
  ylab(expression(Lambda(t))) + xlab("t") + 
  ggtitle("Comparison of cumulative hazards using \n Cox-PH (Nelson-Aalen) vs. PEM")
all.equal(int.df$cumhazard, int.df$hazard)
```
<!-- cumulative hazard of PEM/PAM is NOT step function, it's piecewise linear!!! -->

Alternatively, we could use PAMs. This means estimating the baseline log-hazard rate semi-parametrically as a smooth, non-linear function evaluated at the end-points `tend` of the intervals defined for our model.
Note that the estimated log-hazard value at time-points `tend` gives the value of the log-hazard rate for the **entire previous interval** as PAMs use hazard rates that are **constant** in each interval -- that's where the *piecewise* in the name of the package comes from. Estimating the log hazard rate as a smooth function evaluated at `tend` instead of using an unpenalized estimator without such a smoothness assumption ensures that the hazard rate does not change too rapidly from interval to interval unless there is very strong evidence for such changes in the data.

```{r, fig.width = 4.5, fig.height = 4.5}
pam <- gam(ped_status ~ s(tend), data = ped, offset = offset, family = poisson())
summary(pam)
int.df$pamhaz <- predict(pam, newdata = int.df, type = "response")
int.df %<>% mutate(pamch = cumsum(pamhaz * intlen))
ggplot(int.df, aes(x = tend)) + 
  geom_line(aes(y = cumhazard, col = "PEM"), lty = 3) + 
  geom_step(aes(y = hazard, col = "Nelson-Aalen"), lty = 2) + 
  geom_line(aes(y = pamch, col = "PAM")) +
  scale_color_brewer(name = "Method", palette = "Set2", 
    breaks = c("Nelson-Aalen", "PEM", "PAM")) + 
  theme(legend.position = "bottom") + 
  ylab(expression(Lambda(t))) + xlab("t") + 
  ggtitle(paste0("Comparison of cumulative hazards using\n", 
    "Cox-PH (Nelson-Aalen) vs. PEM vs. PAM"))
```
