---
title: "Baseline Hazard"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{baseline}
 %\VignetteEngine{knitr::rmarkdown}
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  fig.align = "center",
  crop      = TRUE)
```


```{r, message = FALSE}
library(ggplot2)
theme_set(theme_bw())
library(dplyr)
library(survival)
library(mgcv)
library(pammtools)
Set1 <- RColorBrewer::brewer.pal(9, "Set1")
```


Here we briefly demonstrate how to fit and visualize a simple
baseline model using the **`pammtools`** package.
We illustrate the procedure using the `veteran` data from the
`survival` package:

```{r}
data("veteran", package = "survival")
veteran <- veteran %>%
  mutate(
    trt   = 1 * (trt == 2),
    prior = 1 * (prior != 10)) %>%
  filter(time < 400)
```

The below graph depicts the estimated cumulative hazard using the
Nelson-Aalen estimator:


```{r, fig.width = 4.5, fig.height = 4.5}
base_df <- basehaz(coxph(Surv(time, status)~1, data = veteran)) %>%
  rename(nelson_aalen = hazard)
ggplot(base_df, aes(x = time, y = nelson_aalen)) +
  geom_stephazard() +
  ylab(expression(hat(Lambda)(t))) + xlab("t") +
  ggtitle("Nelson-Aalen estimate of the cumulative hazard")
```


## Data transformation
First we need to bring the data in a suitable format
(see [vignette on data transformation](data-transformation.html)).

```{r}
# Use unique event times as interval break points
ped <- veteran %>% as_ped(Surv(time, status)~., id = "id")
head(ped[, 1:10])
```

## PEMs and PAMs

### Fit the baseline using piece-wise exponential model (PEM)
```{r}
pem <- glm(ped_status ~ interval, data = ped, offset = offset, family = poisson())
```

Extract cumulative baseline estimate:

```{r}
int_df <- int_info(ped)
head(int_df)
int_df <- int_df %>%
  add_cumu_hazard(pem, ci = FALSE) %>%
  left_join(base_df, by = c("tend" = "time"))
head(int_df)
```

Visualize the PEM estimate:

```{r, fig.width = 4.5, fig.height = 4.5}
ggplot(int_df, aes(x = tend)) +
  geom_stephazard(aes(y = nelson_aalen, col = "Nelson-Aalen")) +
  geom_hazard(aes(y = cumu_hazard, col = "PEM")) +
  scale_color_manual(name = "Method", values = c(Set1[1:2])) +
  theme(legend.position = "bottom") +
  ylab(expression(hat(Lambda)(t))) + xlab("t") +
  ggtitle("Comparison of cumulative hazards estimated by \n Cox-PH (Nelson-Aalen) vs. PEM")
all.equal(int_df$cumu_hazard, int_df$nelson_aalen)
```

### Fit the baseline using Piece-wise exponential additive model (PAM)

Alternatively, we could use PAMs. This means estimating the baseline log-hazard
rate semi-parametrically as a smooth, non-linear function evaluated at the
end-points `tend` of the intervals defined for our model.

Note that the estimated log-hazard value at time-points `tend` gives the value
of the log-hazard rate for the **entire previous interval** as PAMs estimate
hazard rates that are **constant** in each interval - that's where the
*piece-wise* in the name of the method comes from.

Estimating the log hazard rate as a smooth function evaluated at `tend` -
instead of using an unpenalized estimator without such a smoothness assumption -
ensures that the hazard rate does not change too rapidly from interval to
interval unless there is sufficient evidence for such changes in the data.

```{r}
pam <- gam(ped_status ~ s(tend), data = ped, offset = offset, family = poisson())
summary(pam)
```

### Graphical comparison

In the figure below we compare the previous baseline estimates with the
PAM estimates.

<details>
  <summary>Expand here for R-Code</summary>

```{r}
# add PAM predictions to previous data
int_df$pamhaz <- predict(pam, newdata = int_df, type = "response")
int_df <- int_df %>% mutate(pamch = cumsum(pamhaz * intlen))

gg_baseline <- ggplot(int_df, aes(x = tend)) +
  # geom_ribbon(aes(ymin=cumu_lower, ymax=cumu_upper), alpha=0.2) +
  geom_stephazard(aes(y = nelson_aalen, col = "Nelson-Aalen")) +
  geom_line(aes(y = cumu_hazard, col = "PEM")) +
  geom_line(aes(y = pamch,     col = "PAM")) +
  scale_color_manual(
    name   = "Method",
    values = c("PEM" = Set1[2], "PAM" = "black", "Nelson-Aalen" = Set1[1])) +
  theme(legend.position = "bottom") +
  ylab(expression(hat(Lambda)(t))) + xlab("t") +
  ggtitle(paste0("Comparison of cumulative hazards estimated by\n",
    "Cox-PH (Nelson-Aalen) vs. PEM vs. PAM"))
```

</details>

```{r, fig.width = 4.5, fig.height = 4.5, echo = FALSE}
gg_baseline
```

All models are in good agreement.
