---
title: "Bayesian Baseline PAMMs: mgcv, jagam, and brms"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: Remote.bib
link-citations: yes
vignette: >
  %\VignetteIndexEntry{Bayesian Baseline PAMMs: mgcv, jagam, and brms}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
# Collapse non-didactic chunks (package setup, plotting, table code) behind a
# <details> toggle so they don't distract from the modelling code. Set the
# chunk option `fold = "<label>"` to fold a chunk's source; output (plots,
# tables) is shown normally.
local({
  default_source <- knit_hooks$get("source")
  knit_hooks$set(source = function(x, options) {
    out <- default_source(x, options)
    if (is.null(options$fold)) {
      return(out)
    }
    summary <- if (is.character(options$fold)) options$fold else "Show code"
    paste0("<details>\n<summary>", summary, "</summary>\n\n", out, "\n\n</details>")
  })
})
```

```{r libraries, fold = "Show package setup"}
library(dplyr)
library(ggplot2)
library(mgcv)
library(pammtools)
library(survival)
library(tidyr)
library(tibble)

theme_set(theme_bw())
```

This vignette compares baseline cumulative hazard estimates from:

1. a default frequentist PAMM fit with `mgcv`,
2. a Bayesian PAMM fit via `mgcv::jagam` [@Wood2016jagam], which generates JAGS
   [@Plummer2003] code from the GAM specification and samples it with `rjags`,
3. a Bayesian PAMM fit via `brms` [@Burkner2017] (which compiles the model to
   Stan).

All models use the same spline basis for the baseline hazard, and we compare
them against the Nelson-Aalen estimate and against each other on a common time
grid.

> **Build note.** The JAGS and Stan fits run on an ~11,600-row piece-wise
> exponential data set and take 15--20 minutes combined -- too slow to evaluate
> on every build. The model-fitting and prediction code below is therefore shown
> but not executed (`eval = FALSE`); it is run once by
> [`vignettes/precompute_bayesian.R`](https://github.com/adibender/pammtools/blob/master/vignettes/precompute_bayesian.R),
> which saves the resulting `pred_df` to `bayesian_results.rds`. The plots and
> tables below are rendered live from that saved data frame. Running `rjags`
> additionally needs a JAGS installation, and `brms` needs a working Stan
> toolchain.

## Data preparation

```{r data-prep}
data("tumor")
tumor <- tumor %>%
  slice(1:200) %>%
  mutate(id = row_number())

base_df <- basehaz(coxph(Surv(days, status) ~ 1, data = tumor), centered = FALSE) %>%
  rename(nelson_aalen = hazard)

ped <- as_ped(tumor, Surv(days, status) ~ 1, id = "id")
```

## Model fits

### 1) Default PAMM (`mgcv`)

```{r fit-mgcv, eval = FALSE}
pam_mgcv <- mgcv::gam(
  formula = ped_status ~ s(tend, k = 20),
  data = ped,
  family = poisson(),
  offset = offset,
  method = "REML"
)
```

### 2) Bayesian PAMM via `jagam` + `rjags`

```{r fit-jagam, eval = FALSE}
set.seed(101)

jagam_file <- tempfile(fileext = ".jags")
pam_jagam_prep <- mgcv::jagam(
  formula = ped_status ~ s(tend, k = 20),
  data = ped,
  family = poisson(),
  offset = offset,
  file = jagam_file,
  sp.prior = "gamma",
  diagonalize = TRUE
)

library(rjags)
rjags::load.module("glm")
jagam_sampler <- rjags::jags.model(
  file = jagam_file,
  data = pam_jagam_prep$jags.data,
  inits = pam_jagam_prep$jags.ini, # data-driven starts from the penalised GAM
  n.chains = 2,
  quiet = FALSE
)

jagam_post <- rjags::jags.samples(
  model = jagam_sampler,
  variable.names = c("b", "lambda"),
  n.iter = 4000,
  thin = 10
)

pam_jagam <- mgcv::sim2jam(jagam_post, pam_jagam_prep$pregam)
```

### 3) Bayesian PAMM via `brms`

```{r fit-brms, eval = FALSE}
library(brms)
set.seed(101)

# The PED has many rows, so each Stan iteration is comparatively expensive;
# running the two chains on separate cores and starting from `init = 0`
# (a hazard of ~1 before the offset) gives a stable, reasonably quick fit.
pam_brms <- brms::brm(
  formula = ped_status ~ s(tend, k = 20) + offset(offset),
  data = ped,
  family = poisson(),
  chains = 2,
  cores = 2,
  iter = 3000,
  warmup = 750,
  init = 0,
  seed = 101,
  refresh = 0,
  silent = 1
)
summary(pam_brms)
brms::pp_check(pam_brms) +
  labs(title = "Posterior predictive check")
```

## Common prediction pipeline

All three models are now driven through the *same* `add_cumu_hazard()` call.
For `mgcv` and `jagam` this uses the analytic (coefficient-based) confidence
intervals as before. For `brms` we make the fit a first-class pammtools backend
by defining the two primitives of the unified inference interface --
`get_hazard()` (point hazard) and `sim_hazard()` (matrix of posterior hazard
draws). Once those exist, `add_cumu_hazard(ci_type = "sim")` builds the
cumulative hazard and its posterior credible interval from them, so no manual
`cumsum`/quantile bookkeeping is needed. This is exactly the mechanism the
[xgboost backend vignette](xgboost-backend.html) uses for a tree ensemble.

```{r prediction-setup}
conf_level <- 0.95

newdata_base <- make_newdata(ped, tend = sort(unique(ped$tend))) %>%
  arrange(tend)

method_labels <- c(
  mgcv = "PAMM (mgcv)",
  jagam = "Bayesian PAMM (jagam)",
  brms = "Bayesian PAMM (brms)"
)
```

```{r prediction-code, eval = FALSE}
# brms as a pammtools backend: the two primitives the unified inference
# interface needs. `sim_hazard()` returns one column per posterior draw; setting
# the offset to 0 yields the bare hazard rate (exp(eta) without the
# log-interval-length term), which `add_cumu_hazard()` then multiplies by the
# interval length itself.
sim_hazard.brmsfit <- function(object, newdata, nsim = NULL, ...) {
  newdata$offset <- 0
  ep <- brms::posterior_epred(
    object,
    newdata = newdata,
    summary = FALSE,
    re_formula = NA
  )
  t(as.matrix(ep)) # rows = time points, cols = posterior draws
}
get_hazard.brmsfit <- function(object, newdata, ...) {
  rowMeans(sim_hazard.brmsfit(object, newdata, ...))
}

get_cumu <- function(model, method_id, newdata, conf_level = 0.95) {
  # jam (jagam posterior) objects need to advertise the gam/glm/lm classes so
  # that the analytic CI machinery and design-matrix extraction dispatch.
  if (inherits(model, "jam")) {
    class(model) <- unique(c(class(model), "gam", "glm", "lm"))
  }
  # brms has no coefficient covariance -> simulation CIs from posterior draws;
  # mgcv/jagam keep their analytic (coefficient-based) CIs.
  ci_type <- if (inherits(model, "brmsfit")) "sim" else "default"
  pred <- add_cumu_hazard(
    newdata,
    model,
    ci = TRUE,
    ci_type = ci_type,
    se_mult = stats::qnorm((1 + conf_level) / 2),
    alpha = 1 - conf_level,
    # boundary = FALSE keeps all methods on the identical `tend` grid: the
    # default would prepend a t = 0 row for gam/jagam but not for brms, which
    # would misalign the curves and the downstream comparison join.
    boundary = FALSE
  )

  tibble::tibble(
    tend = pred$tend,
    method_id = method_id,
    method = unname(method_labels[method_id]),
    cumu_hazard = pred$cumu_hazard,
    cumu_lower = pred$cumu_lower,
    cumu_upper = pred$cumu_upper
  )
}

pred_df <- bind_rows(
  get_cumu(pam_mgcv, "mgcv", newdata_base, conf_level = conf_level),
  get_cumu(pam_jagam, "jagam", newdata_base, conf_level = conf_level),
  get_cumu(pam_brms, "brms", newdata_base, conf_level = conf_level)
)
```

```{r load-results, include = FALSE}
# Precomputed by vignettes/precompute_bayesian.R (see the build note above).
pred_df <- readRDS("bayesian_results.rds")
```

```{r prediction-curves, fold = "Reshape for comparison"}
# Guard against a stale `bayesian_results.rds`: the comparison below joins the
# precomputed `pred_df` to the live `newdata_base` grid by position, so the two
# grids must match. Fail loudly if the precompute is out of sync with the Rmd.
stopifnot(identical(
  sort(unique(pred_df$tend)),
  sort(unique(newdata_base$tend))
))

na_on_grid <- stats::approx(
  x = base_df$time,
  y = base_df$nelson_aalen,
  xout = newdata_base$tend,
  method = "constant",
  f = 0,
  rule = 2
)$y

pred_wide <- pred_df %>%
  select(tend, method_id, cumu_hazard) %>%
  pivot_wider(names_from = method_id, values_from = cumu_hazard) %>%
  mutate(nelson_aalen = na_on_grid)

method_ids <- c("mgcv", "jagam", "brms")
pairwise_combos <- utils::combn(method_ids, 2, simplify = FALSE)
```

## Visual comparison to Nelson-Aalen

```{r comparison-plot, fig.width = 7.2, fig.height = 4.8, fold = "Plotting code"}
ggplot(pred_df, aes(x = tend, y = cumu_hazard, color = method, fill = method, linetype = method)) +
  geom_ribbon(aes(ymin = cumu_lower, ymax = cumu_upper), alpha = 0.18, color = NA) +
  geom_line(linewidth = 0.75) +
  geom_stephazard(
    data = base_df,
    aes(x = time, y = nelson_aalen),
    inherit.aes = FALSE,
    color = "black",
    linetype = 2
  ) +
  labs(
    x = "time",
    y = expression(hat(Lambda)(t)),
    title = "Baseline cumulative hazard comparison",
    subtitle = "Dashed black curve: Nelson-Aalen reference"
  ) +
  theme(legend.position = "bottom")
```


## Numeric comparisons

```{r comparison-na-table, fold = "Table code"}
cmp_na_tbl <- bind_rows(lapply(method_ids, function(method_id) {
  diff <- pred_wide[[method_id]] - pred_wide$nelson_aalen
  tibble(
    method = method_labels[[method_id]],
    rmse_vs_nelson_aalen = sqrt(mean(diff^2)),
    max_abs_diff_vs_nelson_aalen = max(abs(diff))
  )
}))

knitr::kable(cmp_na_tbl, digits = 4)
```

```{r comparison-pairwise-table, fold = "Table code"}
cmp_pairwise_tbl <- bind_rows(lapply(pairwise_combos, function(ids) {
  lhs <- ids[[1]]
  rhs <- ids[[2]]
  diff <- pred_wide[[lhs]] - pred_wide[[rhs]]
  tibble(
    contrast = paste0(method_labels[[lhs]], " - ", method_labels[[rhs]]),
    rmse = sqrt(mean(diff^2)),
    max_abs_diff = max(abs(diff))
  )
}))
knitr::kable(cmp_pairwise_tbl, digits = 4)
```

## References
