---
title: "Bayesian Baseline PAMMs: `mgcv`, `jagam`, and `brms`"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Baseline PAMMs}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(
  fig.align = "center",
  cache = TRUE,
  message = FALSE,
  warning = FALSE
)
```

```{r libraries}
library(dplyr)
library(ggplot2)
library(mgcv)
library(pammtools)
library(survival)
library(tidyr)
# Bayesian back-ends. These are hard dependencies of this article: loading them
# here makes the build fail loudly (instead of silently skipping comparisons) if
# `rjags`/JAGS or `brms`/Stan are not available in the rendering environment.
library(rjags)
library(brms)

theme_set(theme_bw())
```

This vignette compares baseline cumulative hazard estimates from:

1. a default frequentist PAMM fit with `mgcv`,
2. a Bayesian PAMM fit via `mgcv::jagam` + `rjags`,
3. a Bayesian PAMM fit via `brms`.

All models use the same spline basis for the baseline hazard, and we compare
them against the Nelson-Aalen estimate and against each other on a common time
grid.

## Data preparation

```{r data-prep}
data("tumor")
tumor <- tumor %>%
  slice(1:200) %>%
  mutate(id = row_number())

base_df <- basehaz(coxph(Surv(days, status) ~ 1, data = tumor), centered = FALSE) %>%
  rename(nelson_aalen = hazard)

ped <- as_ped(tumor, Surv(days, status) ~ 1, id = "id")
```

## Model fits

### 1) Default PAMM (`mgcv`)

```{r fit-mgcv}
pam_mgcv <- mgcv::gam(
  formula = ped_status ~ s(tend, k = 20),
  data = ped,
  family = poisson(),
  offset = offset,
  method = "REML"
)
```

### 2) Bayesian PAMM via `jagam` + `rjags`

```{r fit-jagam}
set.seed(101)

jagam_file <- tempfile(fileext = ".jags")
pam_jagam_prep <- mgcv::jagam(
  formula = ped_status ~ s(tend, k = 20),
  data = ped,
  family = poisson(),
  offset = offset,
  file = jagam_file,
  sp.prior = "gamma",
  diagonalize = TRUE
)

rjags::load.module("glm")
jagam_sampler <- rjags::jags.model(
  file = jagam_file,
  data = pam_jagam_prep$jags.data,
  inits = pam_jagam_prep$jags.ini,
  n.chains = 2,
  quiet = FALSE
)

jagam_post <- rjags::jags.samples(
  model = jagam_sampler,
  variable.names = c("b", "lambda"),
  n.iter = 4000,
  thin = 10
)

pam_jagam <- mgcv::sim2jam(jagam_post, pam_jagam_prep$pregam)
```

### 3) Bayesian PAMM via `brms`

```{r fit-brms}
set.seed(101)

pam_brms <- brms::brm(
  formula = ped_status ~ s(tend, k = 20) + offset(offset),
  data = ped,
  family = poisson(),
  chains = 2,
  iter = 3000,
  warmup = 750,
  seed = 101,
  refresh = 0,
  silent = 1
)
summary(pam_brms)
brms::pp_check(pam_brms) +
  labs(title = "Posterior predictive check")
```

## Common prediction pipeline

```{r prediction-helpers}
conf_level <- 0.95

newdata_base <- make_newdata(ped, tend = sort(unique(ped$tend))) %>%
  arrange(tend)

# `mgcv`/`jagam` predictions go through `add_cumu_hazard()`, which reconstructs
# the interval lengths from the fitted object. The `brms` predictions, however,
# are produced directly via `posterior_epred()` and therefore need the model
# offset (the log interval length) to be present in the prediction data.
newdata_brms <- newdata_base %>%
  left_join(int_info(ped) %>% transmute(tend, offset = log(intlen)), by = "tend")

method_labels <- c(
  mgcv = "PAMM (mgcv)",
  jagam = "Bayesian PAMM (jagam)",
  brms = "Bayesian PAMM (brms)"
)
get_cumu_from_pamm <- function(model, method_id, newdata, conf_level = 0.95) {
  se_mult <- stats::qnorm((1 + conf_level) / 2)
  add_model <- model
  if (inherits(add_model, "jam")) {
    class(add_model) <- unique(c(class(add_model), "gam", "glm", "lm"))
  }
  pred <- add_cumu_hazard(
    newdata = newdata,
    object = add_model,
    ci = TRUE,
    se_mult = se_mult
  )

  tibble(
    tend = pred$tend,
    method_id = method_id,
    method = unname(method_labels[method_id]),
    cumu_hazard = pred$cumu_hazard,
    cumu_lower = pred$cumu_lower,
    cumu_upper = pred$cumu_upper
  )
}

get_cumu_from_brms <- function(model, method_id, newdata, conf_level = 0.95) {
  alpha <- 1 - conf_level
  epred <- brms::posterior_epred(
    model,
    newdata = newdata,
    summary = FALSE,
    re_formula = NA
  )
  if (length(dim(epred)) != 2L) {
    stop("Expected a 2D draws-by-time matrix from brms::posterior_epred().")
  }
  epred <- as.matrix(epred)
  cumu_draws <- t(apply(epred, 1, cumsum))
  bounds <- apply(
    cumu_draws,
    2,
    stats::quantile,
    probs = c(alpha / 2, 1 - alpha / 2)
  )

  tibble(
    tend = newdata$tend,
    method_id = method_id,
    method = unname(method_labels[method_id]),
    cumu_hazard = colMeans(cumu_draws),
    cumu_lower = bounds[1, ],
    cumu_upper = bounds[2, ]
  )
}
```

```{r prediction-curves}
pred_df <- bind_rows(
  get_cumu_from_pamm(pam_mgcv, "mgcv", newdata_base, conf_level = conf_level),
  get_cumu_from_pamm(pam_jagam, "jagam", newdata_base, conf_level = conf_level),
  get_cumu_from_brms(pam_brms, "brms", newdata_brms, conf_level = conf_level)
)

na_on_grid <- stats::approx(
  x = base_df$time,
  y = base_df$nelson_aalen,
  xout = newdata_base$tend,
  method = "constant",
  f = 0,
  rule = 2
)$y

pred_wide <- pred_df %>%
  select(tend, method_id, cumu_hazard) %>%
  pivot_wider(names_from = method_id, values_from = cumu_hazard) %>%
  mutate(nelson_aalen = na_on_grid)

method_ids <- c("mgcv", "jagam", "brms")
pairwise_combos <- utils::combn(method_ids, 2, simplify = FALSE)
```

## Visual comparison to Nelson-Aalen

```{r comparison-plot, fig.width = 7.2, fig.height = 4.8}
ggplot(pred_df, aes(x = tend, y = cumu_hazard, color = method, fill = method, linetype = method)) +
  geom_ribbon(aes(ymin = cumu_lower, ymax = cumu_upper), alpha = 0.18, color = NA) +
  geom_line(linewidth = 0.75) +
  geom_stephazard(
    data = base_df,
    aes(x = time, y = nelson_aalen),
    inherit.aes = FALSE,
    color = "black",
    linetype = 2
  ) +
  labs(
    x = "time",
    y = expression(hat(Lambda)(t)),
    title = "Baseline cumulative hazard comparison",
    subtitle = "Dashed black curve: Nelson-Aalen reference"
  ) +
  theme(legend.position = "bottom")
```


## Numeric comparisons

```{r comparison-na-table}
cmp_na_tbl <- bind_rows(lapply(method_ids, function(method_id) {
  diff <- pred_wide[[method_id]] - pred_wide$nelson_aalen
  tibble(
    method = method_labels[[method_id]],
    rmse_vs_nelson_aalen = sqrt(mean(diff^2)),
    max_abs_diff_vs_nelson_aalen = max(abs(diff))
  )
}))

knitr::kable(cmp_na_tbl, digits = 4)
```

```{r comparison-pairwise-table}
cmp_pairwise_tbl <- bind_rows(lapply(pairwise_combos, function(ids) {
  lhs <- ids[[1]]
  rhs <- ids[[2]]
  diff <- pred_wide[[lhs]] - pred_wide[[rhs]]
  tibble(
    contrast = paste0(method_labels[[lhs]], " - ", method_labels[[rhs]]),
    rmse = sqrt(mean(diff^2)),
    max_abs_diff = max(abs(diff))
  )
}))
knitr::kable(cmp_pairwise_tbl, digits = 4)
```
