---
title: "Vignette Title"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(
  fig.align  = "center",
  cache      = TRUE,
  message    = FALSE,
  fig.height = 5,
  fig.width  = 5
)
```


# Convenience functions

Survival analysis is accompanied by a set of standard task, 
aside of fitting the model, that need to be performed in most practical applications. 
These task include: 

- summary/visualization of fixed effects 
- summary/visualization of smooth effects
- extraction/visualization of the baseline hazard
- extraction/visualization of the hazard/cumulative hazard/survival probabilities
for specific covariate specifications
- comparison of hazard/cumulative hazard/survival probabilities for a set 
of covariate specifications
- comparing hazard/cumulative hazard/survival probabilities estimated using 
different model specifications or different subsets of the data 

This package contains some convenience functions to achieve these tasks. 
The overall philosophy of the package was to provide functions that return the 
underlying data used for visualization in a tidy format, such that everybody 
can use familiar tools for further processing. However, some convenience functions 
for visualization are also presented. 

For illustration we use the `mgus2` data from the survival package. 


## Summary/visualization of fixed effects 

```{r}
library(pam)
library(mgcv)
library(magrittr)
library(dplyr)
data("mgus2", package="survival")
ped <- split_data(Surv(futime, death)~age + hgb + sex, data=mgus2, id="id")
pam <- gam(status~s(tend) + sex + age + s(hgb), data=ped, family=poisson(), offset=offset)
```

The easiest way to obtain overview of the estimated coefficients for fixed effects 
is to call the `summary` function on the fitted model object, as is usual 
for other model objects in the R Language.

```{r options}
summary(pam)
```



Here, male subjects have increased risk by a factor of 
$\exp(\Sexpr{round(coef(pam)[2], 2)}) = \Sexpr{round(exp(coef(pam)[2]), 2)}.

Additionally, this package provides a function to visualize fixed coefficients 
in terms of tree/forest plots: 

```{r}
## yet to be implemented 
```


## Summary/visualization of smooth effects
A first quick overview over the smooth effects is provided in the bottom part of 
output of the `summary` call above. The interpretation is the same as as for 
`mgcv::gam` objects (see `?summary.gam` for details). Briefly, the `edf` column 
provides the number of effective degrees of freedom used to fit the spline 
(values larger than 1 indicate non-linearity) and the p-value refers to 
significance tests with $H_0: f_j(x_j) = 0$ (developed in Wood 2015). 

Usually however, we want to visualize the non-linear effect by plotting 
$f_j(x_j)$ over the range of $x_j$ values. One possibility using base R 
plotting functions is to call the default plot function for `mgcv::gam` objects: 

```{r}
plot(pam, select=2)
```

If you want to customize the graphic using other plotting paradigms you can 
extract the data used for plotting via: 

```{r}
hgb.term <- ped %>% get_terms(fit=pam, terms="hgb")
head(hgb.term)
```

Then we can use for example `ggplot2` to produce an equivalent visualization. 

```{r}
library(ggplot2)
ggplot(hgb.term, aes(x=x, y=eff)) + 
	geom_line() + 
	geom_ribbon(aes(ymin=ci.lower, ymax=ci.upper), alpha=0.3)
```

For convenience we also include a function that plots or selected smooth 
effects at once: 

```{r}
ped %>% gg_smooth(pam, terms=c("tend", "hgb"))
```


## Extraction/Visualization of the baseline hazard

In PAMs the baseline hazard \lambda_0(t) is estimated semi-parametrically as a 
(potentially) non-linear smooth function. Thus the model specification usually 
looks like 

$$
\log(\lambda(t|x)) = f_0(t) + \beta_0 + \ldots.
$$

Thus the term $f_0(t)$ represents the time-dependent part of the ($log$-)baseline-hazard. 

To create a data set with all interval cut-points we can use the function 
`ped_info` applied to an object of class `ped` (piece-wise exponential data). 
In addition to all time-points the function returns mean values for continuous 
variables and modus for categorical variables.

```{r}
ped.info  <- ped_info(ped)
head(ped.info)
tail(ped.info)
```

This data set can be used 
to conveniently obtain the hazard, cumulative hazard and survival functions evaluated 
at all time points in the data. 


```{r}
ped.info %<>% add_hazard(pam)
ped.info %>% select(interval, hazard, lower, upper) %>% head()
ped.info %>% select(interval, hazard, lower, upper) %>% tail()
```

Which in turn can be conveniently plotted using respective functions, e.g. 

```{r}
ggplot(ped.info, aes(x=tend)) + 
	geom_step(aes(y=hazard)) + 
	geom_stepribbon(aes(ymin=lower, ymax=upper), alpha=0.2) + 
	ylab(expression(hat(lambda)(t))) + xlab(expression(t))
```

or the cumulative hazard 

```{r}
gg.chaz <- ped.info %>% add_cumhazard(pam) %>% 
	ggplot(aes(x=tend, y=cumhazard, ymin=cumlower, ymax=cumupper))
gg.chaz + geom_step() + geom_stepribbon(alpha=0.2) + 
	ylab(expression(hat(Lambda(t)))) + xlab(expression(t))
```

This approach is very convenient, as the usual tidy `dplyr` workflow can 
be maintained throughout, for example, if we wanted to plot the cumulative 
hazard for different sexes:

```{r}
pinfo.sex <- ped %>% group_by(sex) %>% ped_info() %>% add_cumhazard(pam)
head(pinfo.sex)
ggplot(pinfo.sex, aes(x=tend, y=cumhazard)) + 
	geom_stepribbon(aes(ymin=cumlower, ymax=cumupper, fill=sex), alpha=0.3) + 
	geom_step(aes(col=sex))
```

