---
title: "Convenience functions for post-processing PAMs"
author: "Andreas Bender"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{convenience}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: pam.bib
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  fig.align  = "center",
  cache      = TRUE,
  message    = FALSE,
  fig.height = 5,
  fig.width  = 5
)
```


# Convenience functions

Survival analysis entails a set of standard tasks 
beyond model estimation which need to be performed in most real world applications. 
These task include: 

- summary/visualization of estimated effects 
- extraction/visualization of the estimated baseline hazard
- extraction/visualization of the predicted (cumulative) hazards or survival probabilities
for specific covariate specifications
- comparison of the predicted (cumulative) hazards or survival probabilities for a set 
of covariate specifications
- comparison of the predicted (cumulative) hazards or survival probabilities and estimated effects
resulting from different model specifications or different subsets of the data. 

This package contains some convenience functions to achieve these tasks. 
The overall philosophy of the package was to provide functions that return the 
underlying data used for visualization in a *tidy* format [@wickham2014], such that everybody 
can use familiar tools for further processing. In addition, some convenience functions 
for visualization are also included. 

We use the `mgus2` data [@kyle1993] from the `survival` package for illustration 
in the following.  


## Summary/visualization of fixed effects 

```{r model, cache =TRUE}
library(pam)
library(mgcv)
library(magrittr)
library(dplyr)
data("mgus2", package = "survival")
ped <- split_data(Surv(futime, death)~age + hgb + sex, data = mgus2, id = "id")
pam <- gam(ped_status ~ s(tend) + sex + age + s(hgb), data = ped, 
  family = poisson(), offset = offset)
```

The easiest way to obtain an overview of the estimated coefficients for fixed effects 
is to simply call `summary` on the fitted model object:

```{r options}
summary(pam)
```

We can see that the estimated mortality risk for male subjects is increased by a factor of 
about $\exp(`r round(coef(pam)[2], 2)`) = `r round(exp(coef(pam)[2]), 2)`$ compared to female subjects.

<!--
Additionally, this package provides a function to visualize fixed coefficients 
in terms of tree/forest plots: 

```{r}
## yet to be implemented 
```
-->

## Summary/visualization of smooth effects
A first quick overview over the smooth effects is provided in the bottom part of 
output of the `summary` call above. The interpretation is the same as as for 
`mgcv::gam` objects (see `?summary.gam` for details). Briefly, the `edf` column 
provides the number of effective degrees of freedom used to fit the spline 
(values larger than 1 indicate non-linearity) and the p-value refers to 
a significance test against the null hypothesis of "no effect" 
(i.e., $H_0: f_j(x_j) = 0$; test developed in @wood2013). 

Usually, however, we want to visualize the non-linear effect by plotting 
$f_j(x_j)$ over the range of $x_j$ values. One possibility using base R 
plotting functions is to call the default plot function for `mgcv::gam` objects.
Note that PAM effects are additive on the log-hazard scale, so we have to plot 
$\exp(f_j(x_j))$ (right panel below) in order to read hazard *ratios* off the figure:

```{r, fig.width=8}
layout(t(1:2))
plot(pam, select = 2, ylim = c(-1, 2))
abline(h = 0, col = "grey")
plot(pam, select = 2, trans = exp, #plot exp(f(x)) ...
  ylim = c(-1, 2), ylab = "exp(f(hgb))", log = "y")  #.. but use logarithmic axis for legibility 
abline(h = 1, col = "grey")
```

If you want to customize the figure using other graphics packages you can 
extract the data used for plotting via: 

```{r}
hgb.term <- ped %>% get_terms(fit = pam, terms = "hgb")
head(hgb.term)
```

We can then use `ggplot2` to produce an equivalent visualization: 

```{r}
library(ggplot2)
theme_set(theme_bw())
scale_fill_discrete <- scale_fill_brewer
scale_color_discrete <- scale_color_brewer
ggplot(hgb.term, aes(x = x, y = eff)) + 
	geom_line() + 
	geom_ribbon(aes(ymin = ci.lower, ymax = ci.upper), alpha = 0.3)
```

For convenience we also include a function that plots mutiple selected smooth 
effects in one figure: 

```{r}
ped %>% gg_smooth(pam, terms = c("tend", "hgb"))
```


## Extraction/Visualization of the baseline hazard

In PAMs the baseline hazard $\lambda_0(t)$ is estimated semi-parametrically as a 
(potentially) non-linear smooth function. Thus the model specification usually 
looks like 

$$
\log(\lambda(t|x)) = \beta_0 + f_0(t) + \ldots
$$

Thus the term $f_0(t)$ represents the time-varying part of the log-baseline-hazard. 

To create a data set with all interval cut-points we can use the function 
`ped_info` applied to an object of class `ped` (**p**iece-wise **e**xponential **d**ata). 
In addition to information about each interval, the data set returned by the function contains 
mean values of continuous variables and the modal values of categorical variables.

```{r}
ped.info  <- ped_info(ped)
head(ped.info)
tail(ped.info)
```

This data set can be used to conveniently obtain the hazard, cumulative hazard and survival functions evaluated at the joint mean (and mode) of the covariates for all time points in the data:

```{r}
ped.info %<>% add_hazard(pam)
ped.info %>% select(interval, hazard, lower, upper) %>% head()
ped.info %>% select(interval, hazard, lower, upper) %>% tail()
```

Which in turn can be conveniently plotted using respective functions, e.g. 

```{r}
ggplot(ped.info, aes(x = tend)) + 
	geom_step(aes(y = hazard)) + 
	geom_stepribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) + 
	ylab(expression(hat(lambda)(t))) + xlab(expression(t))
```

or the cumulative hazard 

```{r}
gg.chaz <- ped.info %>% add_cumhazard(pam) %>% 
	ggplot(aes(x = tend, y = cumhazard, ymin = cumlower, ymax = cumupper))
gg.chaz + geom_line() + geom_ribbon(alpha = 0.2) + 
	ylab(expression(hat(Lambda(t)))) + xlab(expression(t))
```

This approach is very convenient, as the usual tidy `dplyr` workflow can 
be maintained throughout, for example, if we wanted to plot the cumulative 
hazard for different sexes:

```{r}
pinfo.sex <- ped %>% group_by(sex) %>% ped_info() %>% add_cumhazard(pam)
head(pinfo.sex)
ggplot(pinfo.sex, aes(x = tend, y = cumhazard)) + 
	geom_ribbon(aes(ymin = cumlower, ymax = cumupper, fill = sex), alpha = 0.3) + 
	geom_line(aes(col = sex)) +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1")
```

## References
