---
title: "Workflow and convenience functions"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Workflow and convenience functions}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: Remote.bib
link-citations: yes
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  fig.align = "center",
  crop      = TRUE,
  cache     = TRUE)
```

```{r, message=FALSE}
library(dplyr)
library(mgcv)
library(pammtools)
library(ggplot2)
theme_set(theme_bw())
```

Survival analysis entails a set of standard tasks
beyond model estimation which need to be performed in most real world applications.
These task include:

- extraction/visualization of the predicted (cumulative) hazards or survival
- summary/visualization of estimated effects
probabilities for specific covariate specifications
- visualization of non-linear covariate effects
- ...

The **`pammtools`** package provides many convenience functions to facilitate
these tasks. The overall philosophy of the package is to provide functions that
return the underlying data used for visualization in a *tidy* format [@wickham2014],
such that everybody can use familiar tools for further processing. In addition,
some (high level) convenience functions for visualization are also included.

To illustrate the usual workflow and some of the convenience functions provided
by **`pammtools`**, we will use the `tumor` in the following sections.
For a more complete overview see @Bender2018f.


## Hazard, cumulative hazard and survival probabilities
Often one is interested in calculating the hazard, cumulative hazard or
survival probability given specific values for some covariates (while other
covariates are fixed at their mean values). For PAMMs, one usually also needs
to create a data set where all intervals occur, in order to be able to calculate
predicted values at different time points. **`pammtools`** provides a flexible
interface to create such data sets (`make_newdata`) and the functions

  - `add_hazard`
  - `add_cumu_hazard` and
  - `add_surv_prob`

to add respective predicted values (including confidence intervals) of the
respective quantities. We illustrate the workflow using the `tumor` data and
the model given below:

```{r}
ped <- tumor %>% as_ped(Surv(days, status)~ age + sex + complications, id = "id")
pam <- gam(ped_status ~ s(tend) + sex + age + complications, data = ped,
  family = poisson(), offset = offset)
```

In the following, we create a new data set that contains all intervals as well
as mean values for all other covariates. Note that within `make_newdata` you
can specify the desired covariate values by using any function applicable to
the data type of the respective column. The new data will contain one row for
each combination of the two variables (similar to `expand.grid`):

```{r}
ped_df <- ped %>% make_newdata(tend = unique(tend), age = seq_range(age, n = 3))
```

Calling the respective functions, we can add the (predicted) hazard to the newly
created data.

### Hazard

```{r}
ped_df <- ped_df %>% add_hazard(pam)
ped_df %>% select(interval, hazard, ci_lower, ci_upper) %>% head()
ped_df %>% select(interval, hazard, ci_lower, ci_upper) %>% tail()
```

Which in turn can be conveniently plotted using respective functions, e.g.

```{r, fig.width=4, fig.height=3}
ggplot(ped_df, aes(x = tend, group = age)) +
	geom_stephazard(aes(y = hazard, col = age)) +
	geom_stepribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = age), alpha = 0.2) +
	ylab(expression(hat(lambda)(t))) + xlab(expression(t))
```

### Hazard Ratio
The hazard ratio can be calculated by specifying the `reference`
argument:

```{r}
# the code below gives hazard reatio of a 30 year old relative to a 58 year old (cp)
# ped %>%
#   make_newdata(age = c(30)) %>%
#   add_hazard(pam, reference = list(age = c(58))) %>%
#   select(hazard, ci_lower, ci_upper)
# -> 30 year old has about half the risk of a 58 year old
```

This might seem a little convoluted, but can be easily generalized in case of time-varying effects, time-dependent covariates, cumulative effects, etc.


### Cumulative hazard
Analogously, we can obtain the cumulative hazard (piece-wise linear function).
Note that for the cumulative hazard and survival probability it is **important to
group the data** accordingly, such that the cumulative hazard is cumulated
for each value of the grouping variable. Otherwise it would be cumulated over the
whole data set (which is usually not intended, except if the specified variable
is a time-dependent covariate).

```{r, fig.width=4, fig.height=3}
ped_df %>%
  group_by(age) %>%
  add_cumu_hazard(pam) %>%
	ggplot(aes(x = tend, y = cumu_hazard, ymin = cumu_lower, ymax = cumu_upper, group = age)) +
    geom_hazard(aes(col = age)) + geom_ribbon(aes(fill = age), alpha = 0.2) +
	 ylab(expression(hat(Lambda)(t))) + xlab(expression(t))
```

### Survival probability

```{r, fig.width=4, fig.height=3}
ped_df %>% group_by(age) %>% add_surv_prob(pam) %>%
  ggplot(aes(x = tend, y = surv_prob, ymax = surv_lower, ymin = surv_upper, group = age)) +
  geom_line(aes(col = age)) + geom_ribbon(aes(fill = age), alpha = 0.2) +
  ylab(expression(hat(S)(t))) + xlab(expression(t))
```

This approach is very convenient, as it can be extended to arbitrary covariate
specifications:

```{r, fig.width=6, fig.height=3}
ped_df_sex <- ped %>%
  make_newdata(tend = unique(tend), age = seq_range(age, 3),
    complications = levels(complications)) %>%
  group_by(complications, age) %>%
  add_surv_prob(pam)
ggplot(ped_df_sex, aes(x = tend, y = surv_prob, group = age)) +
	geom_ribbon(aes(ymin = surv_lower, ymax = surv_upper, fill = age), alpha = 0.3) +
  geom_surv(aes(col = age)) +
  ylab(expression(hat(S)(t))) + xlab(expression(t)) +
  facet_wrap(~complications, labeller = label_both)
```


## Visualization of non-linear covariate effects
Data sets for the visualization of the (non-linear) effect of a continuous
covariates can be created in a similar fashion. For illustration, we add
a non-linear age effect to the previous model (see also high level
functions`?gg_smooth` and `?gg_tensor` and `gg_re` for `mgcv::plot.gam` like plots):

```{r}
ped <- tumor %>% as_ped(Surv(days, status)~ age + sex + complications +
  charlson_score, id = "id")
pam <- gam(ped_status ~ s(tend) + sex + s(age) + complications ,
  data = ped, family = poisson(), offset = offset)
```

Using `make_newdata` and `add_term` we obtain:
```{r, fig.width = 4, fig.height = 4}
term_charlson <- ped %>%
  make_newdata(age = seq_range(age, n = 25)) %>%
  add_term(pam, term = "age", reference = list(age = mean(.$age)))
ggplot(term_charlson, aes(x = age, y = fit)) +
  geom_line() +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.2)
```

Similarly, we could use high level function `gg_slice` to obtain
the estimated effect of the age variable, relative to the mean age and
age of 30, respectively:

```{r, fig.width = 4, fig.height = 4}
gg_slice(ped, pam, "age", age = seq_range(age, n = 25),
  reference = list(age = mean(ped$age)))
gg_slice(ped, pam, "age", age = seq_range(age, n = 25),
  reference = list(age = c(30)))
```



## References
