---
title: "Cumulative Effects"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cumulative Effects}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: Remote.bib
link-citations: yes
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  fig.align  = "center",
  fig.width  = 4,
  fig.height = 4,
  crop       = TRUE,
  cache      = TRUE)
## output hooks, e.g., to truncate rows of output
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines) == 1) {
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(if (abs(lines[1]) > 1) more else NULL,
            x[lines],
            if (length(x) > lines[abs(length(lines))]) more else NULL
           )
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })
```


This vignette provides a short reference for the estimation and interpretation
of cumulative effects using **`pammtools`**. For a more detailed overview see
@Bender2018f. In the [first section](cumulative-effects.html#introduction), a
short introduction to cumulative effects is provided. In the
[second section](cumulative-effects.html#example-sofa-score-and-extubation),
we present a worked example on a real data set, including necessary data
transformation, model estimation and visualization.

<details>
  <summary>R Setup</summary>

```{r echo = TRUE, message = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
theme_set(theme_bw())
library(mgcv)
library(pammtools)
```

</details>

## Introduction

Cumulative effects can be thought of as an extension to the modeling of
time-constant effects of time-dependent covariates
(TDCs; see vignette on [time-dependent covariates](./tdcovar.html)).
Standard modeling of TDCs assumes that only the
current (or one lagged) value of the covariate can affect the hazard at time $t$,
i.e.,
$$\lambda(t|z(t)) = \lambda_0(t)\exp(\beta z(t)).$$

Cumulative effects on the other hand allow the hazard at time $t$ to depend on
multiple past observations of the covariate, such that the effect at time $t$
is the sum of all weighted effects of past observations (partial effects).
to make this more concrete, consider the following quantities:

  - $t$: the follow-up time
  - $t_z$: the time at which the TDC $z$ was observed (exposure time)
  - $z(t_z)$: the value of covariate $z$ at exposure time $t_z$
  - $\mathbf{z} = \{z(t_{z}):t_z \leq t\}$: the exposure history of TDC $z$,
  i.e., all recorded observations of the TDC (at time t).

Given these definitions, one possible specification of a model with a
cumulative effect was suggested by @Sylvestre2009 and is given below:

$$
\lambda(t|\mathbf{z}) = \lambda_0(t)\exp\left(\sum_{t_z \leq t} h(t-t_z)z(t_z)\right)
$$

Here, past observations are weighted by a (potentially) non-linear function $h$
of the *latency* ($t-t_z$), i.e., the time that has passed since covariate
$z(t_z)$ was observed, relative to the time of interest $t$. The cumulative
effect is then the sum of all *partial effects* $h(t-t_z)z(t_z)$.

The above model also makes some simplifying assumptions, e.g.:

  - the partial effect of the TDC ($z$) is linear in the covariate values
  - the weighting function $h$ only depends on the latency, not on the specific
  combination of follow-up $t$ and *exposure time* $t_z$
  - the covariate has been observed in 1 unit steps w.r.t. time-scale $t$

A more flexible cumulative effect can be defined as follows:

$$
g(\mathbf{z}, t) = \int_{\mathcal{T}(t)}h(t, t_z, z(t_z))\mathrm{d}t_z,
$$

with

  - $g(\mathbf{z},t)$: the cumulative effect of covariate history $\mathbf{z}$
  on the hazard at time $t$ (numerically, the integral is usually approximated
  by a weighted sum)
  - $h(t, t_z, z(t_z))$: the contribution of covariate $z$ observed at
  time $t_z$ with value $z(t_z)$ to the cumulative effect at time $t$
  (*partial effect* in the following). In the framework of PAMMs, these partial
  effects ares estimated by penalized splines.
  - $\mathcal{T}(t)$: the integration limits (lag-lead window) that control
  how many past observations of $\mathbf{z}$ contribute to the cumulative effect
  at time $t$. The minimal requirement is $\mathcal{T}(t)=\{t_{z}: t_z \leq t\}$,
  i.e., only past observations of the TDC can contribute to the cumulative effect at time $t$.

A visual representation of exemplary lag-lead windows is depicted below:

<details>

<summary>
  R-Code lag-lead windows
</summary>

```{r}
p_ll1 <- gg_laglead(0:10, 0:9, ll_fun = function(t, tz) t >= tz)
my_ll_fun <- function(t, tz, tlag = 2, tlead = 5) {
  t >= tz + tlag & t < tz + tlag + tlead
}
p_ll2 <- gg_laglead(0:10, 0:9, ll_fun = function(t, tz) t == tz)
```
</details>

```{r, echo = FALSE, fig.width = 7, fig.height = 3.5, out.width="600px", fig.cap="Illustration of two different lag-lead windows."}
gridExtra::grid.arrange(p_ll1, p_ll2, nrow = 1L)
```

- The left panel represents the minimal requirement
$\mathcal{T}(t)=\{t_{z}: t_z \leq t\}$. For example, the
observation of the TDC at time $t_z=4$, starts to contribute to the cumulative
effect in interval $(4,5]$ and continues to contribute until the last interval
$(9,10]$, although with potentially different weights, as, in general,
$h(t=5, t_z = 4, z(4))$ can be different to $h(t=10, t_z=4, z(4))$.

- The right panel illustrates that a "standard" model for time-to-event data
with a constant effect follows as a special case of the general definition of
the cumulative effect with lag-lead window
$\mathcal{T}(t)=\{t_{z}: t_z = t\}$ and $h(t, t_z, z(t_z)) \equiv \beta z(t)$


## Example: SOFA-score and extubation
For an illustration of the functionality that **`pammtools`** provides to
work with cumulative effects, we use (a sample) of the
data analyzed in @Heyard2018 and provided in the package **`TBFmultinomial`**
(note that their analyses had a totally different goal, including modeling
of competing risks; here we simply use the data for illustrative purposes).

In the following analysis, we will model the (cause-specific) hazard for
the event `extubation`, considering all competing events as censoring events.
Covariates include `gender`, `type` (admission type), `SAPSadmission`
(SAPS score at admission) and the time-dependent covariate `SOFA` score
(lower scores indicate better health).

To use the data for our purposes, we first perform some preprocessing, which
produces two data sets:

  - `event_df`: Data in "standard" format (one row per subject), that only
  includes time-constant covariates
  - `tdc_df`: Data containing information on time-dependent covariates
  (here `SOFA`) and exposure time $t_z$

<details>
  <summary>R-Code data preprocessing</summary>

```{r, warning = FALSE, results = "hide"}
data(VAP_data, package = "TBFmultinomial")
# create unique IDs
VAP_data <- VAP_data %>%
  mutate(tmp_id =  paste(ID, day, sep = ".")) %>%
  group_by(ID) %>%
  mutate(csdup = cumsum(duplicated(tmp_id))) %>%
  ungroup() %>%
  mutate(ID = ifelse(csdup > 0, ID + 1000, ID))

# assume constant SOFA score between updates
VAP_complete <- VAP_data %>%
  group_by(ID) %>%
  mutate(time = max(day)) %>%
  ungroup() %>%
  complete(ID, day = full_seq(day, 1)) %>%
  fill(gender, type, SAPSadmission, SOFA, outcome, time, .direction = "down") %>%
  filter(day <= time)

event_df <- VAP_complete %>%
  select(ID, gender, type, SAPSadmission, outcome, time) %>%
  group_by(ID) %>%
  slice(n()) %>%
  mutate(outcome = 1 * (outcome == "extubated")) %>%
  ungroup()

tdc_df <- VAP_complete %>%
  select(ID, day, SOFA) %>%
  mutate(day = day - 1) %>% # assume that SOFA available at the beginning of the day
  filter(day <= 49)
```
</details>
&nbsp;

An overview of the preprocessed data is given below:

```{r}
event_df %>% head()
tdc_df %>% head()
```

```{r fig.width = 6, fig.height = 4, echo = FALSE, out.width = "600px", fig.cap = "SOFA profiles for individual patients. Each line represents a patient, each pixel represents a measurement."}
tdc_df %>% group_by(ID) %>%
  mutate(
    sort = 1000 * sum(!is.na(SOFA)) +
      rep(mean(SOFA, na.rm = TRUE), sum(!is.na(SOFA)))) %>%
  ungroup() %>%
  mutate(ID = factor(ID) %>%  reorder(sort, mean)) %>%
  ggplot(aes(x = day, fill = SOFA, y = ID)) +
  geom_raster() + scale_fill_viridis_c() +
  theme(axis.text.y = element_blank()) +
  scale_y_discrete("patients", breaks = NULL)
```

### Data transformation
For illustration, we fit a DLNM [@Gasparrini2017] with cumulative effect of
the SOFA score ($z$) as defined below:

$$
g(\mathbf{z}, t) = \int_{\mathcal{T}(t)}h(t-t_z, z(t_z))\mathrm{d}t_z,
$$

with $\mathcal{T}(t) = \{t_z: t_z \leq t \leq t_z + 5\}$, which means that only
the SOFA scores observed within the last five days can affect the hazard at time
$t$.

[Remember](https://adibender.github.io/pammtools/articles/data-transformation.html#data-with-time-dependent-covariates-with-cumulative-effects)
that the specification of the partial effect $h(t,t_z, z(t_z))$ as well
as the definition of the lag-lead window are important parts of the analysis
and need to be considered at the beginning of the analysis as this information
goes into the code for the data transformation:

```{r, message = FALSE}
ped <- as_ped(
  list(event_df, tdc_df),
  Surv(time, outcome) ~ . + cumulative(latency(day), SOFA, tz_var = "day",
      ll_fun = function(t, tz) t >= tz &  t <= tz + 5),
  cut = 0:49, # administrative censoring at t = 49
  id = "ID")
ped$day_latency <- ped$day_latency * ped$LL
```
Above, we

  - used the formula special `cumulative` to inform `as_ped` to perform data
  preprocessing in order to fit cumulative effects
  - wrapped the variable `day` ($t_z$) within the `latency` function to calculate
  $t - t_z$
  - customized the `ll_fun` argument to match the desired specification of
  $\mathcal{T}(t)$

Note that the data now contains 3 matrix columns (`day_latency`, `SOFA` and
`LL`; the latter stores the lag-lead matrix $\mathcal{T}(t)$):



```{r, output.lines = c(1:4, 10:13)}
str(ped, 1)
```

```{r, echo = FALSE, fig.width = 4, fig.height = 4, out.width = "400px", fig.cap = "Lag-lead window for the first 10 days of the follow-up."}
ll_df <- get_laglead(ped)
ll_df <- ll_df %>% filter(t <= 10, tz <= 10)
class(ll_df) <- c("LL_df", class(ll_df))
gg_laglead(ll_df)
```

### Model estimation
After successful data transformation with `as_ped`, the model can be estimated
directly using `mgcv::gam`. Including matrix columns into the model specification
will inform `gam` to estimate cumulative effects. In our case the the following
call estimates the DLNM with a cumulative effect (of the SOFA score) as specified
above:

```{r}
mod <- gam(ped_status ~ s(tend) + type + gender + SAPSadmission +
   te(day_latency, SOFA, by = LL),
  method = "REML", offset = offset, family = poisson(), data = ped)
summary(mod)
```

### Interpretation
Interpretation of the estimated cumulative effects is best performed by
visualization of either the partial effects or the cumulative effect.
**`pammtools`** provides a couple of convenience functions that facilitate this
process:

  - `gg_tensor`: This visualizes 2D effect surfaces as heat maps/contour plots and
  is based on the output of `mgcv::plot.gam`
  - `gg_partial` and `gg_partial_ll`: The former visualizes the estimated partial
  effect, e.g., $\hat h(t-t_z, z(t_z))$ for each combination of the (specified)
  input, the latter visualizes the partial effects as a heat-map within the
  specified lag-lead window (this makes it easier to see which partial effects
  actually contribute to the cumulative effect in a specific interval)
  - `gg_slice`: Plots 1D slices of 2D surfaces and can be used more generally
  to plot covariate effects conditional on pre-specified values of other covariates.
  - `gg_cumu_eff`: Either plots the estimated cumulative effect $\hat g(\mathbf{z}, t)$ for a given $\mathbf{z}$ or the (log-) hazard ratio  $\log\left(\frac{\hat g(\mathbf{z}_1, t)}{\hat g(\mathbf{z}_2, t)}\right)$

Note that for each of these plot functions there are respective functions to
retrieve the data used for plotting such that the plots can be generated and
customized manually (see especially examples in `make_newdata`).

#### 2D partial effect surface
The code below visualizes the estimated 2D partial effect $\hat h(t-t_z, z(t_z))$ for each
combination of latency and SOFA score (including confidence intervals).
The gray areas indicate combinations that did not occur in the data.

```{r fig.width = 9, fig.height = 4, warning = FALSE, out.width="650px"}
gg_tensor(mod, ci = TRUE) + ylab("SOFA") + xlab("latency (day)") +
  theme(legend.position = "bottom")
```

This illustrates that a low SOFA score substantially increases the log-hazard
if it was observed recently (latency $<$ 2 days), while partial effects of the
SOFA score observed further in the past (latency $>$ 3 days) go towards zero
(the event of interest is "extubation", therefore increased (log-)hazards
imply increased probabilities of extubation). Note that for this graphic,
the function $\hat h(t-t_z, z(t_z))$ is evaluated at values of $t-t_z$ (`day_latency`)
that are not present in the data.

Alternatively, the function `gg_partial` can be used to produce similar
visualizations and allows more control over the inputs. For example, the
following code produces the partial effect surface plot evaluated only at
latencies 0 through 5 and calculates the partial effects relative to a patient
with SOFA score 10 (everything else being equal):
```{r fig.width = 4, fig.height = 4, out.width = "400px"}
gg_partial(ped, mod, term = "SOFA", day_latency = 0:5,
  SOFA = seq_range(SOFA, n = 20), LL = c(1), reference = list(SOFA = 10))
```
See also `gg_partial_ll` that shows the lag-lead window and partial effects
that contribute to the cumulative effect in each interval. Due to the definition
of the partial effect, however, these are constant in this case, as no
time-variation was specified.

#### 1D conditional effects (slices through the 2D surface):

```{r, fig.width = 8, fig.height = 4, out.width = "650px"}
p_slice_latency <- gg_slice(ped, mod, term = "SOFA",
  day_latency = unique(day_latency), SOFA = c(0, 5, 10))
p_slice_sofa <- gg_slice(ped, mod, term = "SOFA",
  day_latency = c(0, 1, 4), SOFA = unique(SOFA))
gridExtra::grid.arrange(p_slice_latency, p_slice_sofa, nrow = 1)
```

These plots contain essentially the same information as the 2D surface, but
focus on isolating the effects of the individual variables. This can be
especially useful for three-dimensional partial effects which are hard to
visualize otherwise.

#### Cumulative effects

Since it can be difficult to assess how the estimated partial effects actually
affect estimated hazard rates, **`pammtools`** provides additional functions to
visualize and compare estimated cumulative effects on the level of the
(log-)hazard rates for given TDCs $\mathbf{z}$.

Below, we visualize the cumulative effect in each interval of the follow-up.
Since the effect depends on a TDC $\mathbf{z}$, its values must be provided in
the call to `gg_cumu_eff` as `z1` argument (either length 1 for time-constant
values or length equal to maximum number of times $z(t_z)$ was observed).
If `z2` is additionally specified, the log-hazard ratio is calculated
as $\log\left(\frac{\hat g(\mathbf{z}_1, t)}{\hat g(\mathbf{z}_2, t)}\right)$.
Below, two such log-hazard ratios are visualized for two different comparisons
of SOFA score profiles:

  - On the left hand side, a patient with SOFA scores continuously decreasing
  throughout the follow-up is compared to a patient with constant SOFA score
  of 10
  - On the right hand side, a patient with SOFA scores increasing throughout the
  follow-up is compared to a patient with SOFA score 0 on the first 10 days,
  score 20 on the following 20 days and score 5 on the last 20 days of the
  follow-up.

```{r fig.width = 8, fig.height = 4, out.width = "650px"}
p_cumu1 <- gg_cumu_eff(ped, mod, term = "SOFA",
  z1 = seq(20, 0, length.out = 50), z2 = 10) +
  geom_point(size = rel(.5)) + geom_vline(xintercept = c(5, 25), lty = 2)
p_cumu2 <- gg_cumu_eff(ped, mod, term = "SOFA",
    z1 = seq(0, 20, length.out = 50),
    z2 = c(rep(0, 10), rep(20, 20), rep(5, 20))) +
  geom_point(size = rel(.5)) + geom_vline(xintercept = c(10, 30), lty = 2)
gridExtra::grid.arrange(p_cumu1, p_cumu2, nrow = 1)
```

## References
