---
title: "Data Transformation"
author: "Andreas Bender"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{data-transformation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=TRUE, message=FALSE}
library(pam)
library(magrittr)
library(dplyr)
```


In this vignette we provide details on transforming data into a format suitable 
to fit piece-wise exponential (additive) models (PAM). Three main cases need to be 
distinguished

  1. Data without time-dependent covariates 

  2. Data with time-dependent covariates 

  3. Data with time-dependent covariates that should be modeled as cumulative 
  effects


## Data without time-dependent covariates 

In this simple case the data transformation is relatively straight forward and 
handeled by the `split_data` function. This function internally calls 
`survival::survSplit`, thus all arguments available for the `survSplit` function 
are also available to the `split_data` function. 


### Using defaults

As an example data set we first consider Freireichs (citation) Leukimia data set 
available from the `bpcp` package: 

```{r}
data("leuk2", package="bpcp")
leuk2 %<>%
  mutate(id=row_number()) %>% 
  select(id,everything(), -pair)
head(leuk2)
```

Each row contains information on the survival time (`time`), an event indication
(`status`) and the `treatment` information (`6-MP` vs. `placebo`) as the only 
covariate. 

To bring the data into piece-wise exponential data (PED) format, we need to 

  - define $J+1$ interval break points $t_{\min}=\kappa_0 < \kappa_1 < \cdots < \kappa_J=t_{\max}$ 
  
- create "pseudo" observations for each interval $j=1,\ldots, J$  which subject 
  $i, i=1,\ldots,n$ visited. 

Using the `pam` package this is easily achieved with the `split_data` function
as follows: 

```{r}
library(pam)
ped <- split_data(Surv(time, status)~., data=leuk2, id="id")
dplyr::filter(ped, id %in% 1:2)
```

When no cut points are specified, the default is to use the unique event times. 
As can be seen from the above output, the function creates an `id` variable, 
indicating the subjects $i=1,\ldots,n$, with one row per `id` for each interval 
the subject "visited". Thus subject `1` visited one interval, subject `2` visited 
9 intervals. Additionally to the optional `id` variable the function also 
creates 

  - `tstart`: the start time of the interval
  - `tend`: the end time of the interval
  - `interval`: a factor variable denoting the interval
  - `intmid`: the interval mid-point
  - `intlen`: the interval length (usefull to have to calculate cumulative hazards
  for example)
  - `offset`: the log of `intlen` (must be included when fitting the model)
  
Additionally the time-constant covariates (here only treatment) are repeated 
$n_i$ number of times, where $n_i$ is the number of interval subject $i$ visited. 


### Custom cut points 
Per default the `split_data` function uses all unique event times as cut points 
(which is a sensible default as for example the (cumulative) baseline hazard 
estimates in Cox-PH functions are updated also only at event times). 

In some cases, however, one might want to reduce the number of cut points, 
to reduce the size of the resulting data object and/or for gain of computational 
efficiency  for example). To do so the `cut` argument has to be specified. 
Following the above example, we can use only 4 cut points
$\kappa_0 = 0, \kappa_1=10, \kappa_2=20, \kappa_3 = 35$, resulting in $J=3$ 
intervals: 

```{r}
ped2 <- split_data(Surv(time, status)~., data=leuk2, cut=c(0, 10, 20, 35), id="id")
ped2 %>% filter(id %in% 1:2)
```
Note taht subject with $id=2$ now also has only one row in the data set. The fact 
that subject 2 was in the risk set longer than subject one is however still 
reflected in the data set through the `offset` variable. 


<!-- ## Data with time-dependent covariates 
```{r}
data("pbc2", package = "JM")
pbc2.time <- select(pbc2, id, years, status2)
pbc2.tv <- select(pbc2, year, albumin)
pbc.ped <- split_data(Surv(years, status2)~., data=pbc2.time, id="id")
tv.df <- get_intervals(brks=sort(unique(pbc2$years)), pbc2.tv$year) %>% 
  bind_cols(pbc2.tv)
```
 -->