---
title: "Data Transformation"
author: "Andreas Bender"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{data-transformation}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: Remote.bib
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  fig.align  = "center",
  fig.width  = 4,
  fig.height = 4,
  crop       = TRUE)
```


```{r echo = TRUE, message = FALSE}
library(dplyr)
library(tidyr)
library(pammtools)
```


In this vignette we provide details on transforming data into a format suitable
to fit piece-wise exponential (additive) models (PAM). Three main cases need to be
distinguished

  1. [Data without time-dependent covariates](data-transformation.html#data-without-time-dependent-covariates)

  2. [Data with time-dependent covariates](data-transformation.html#data-with-time-dependent-covariates)

  3. [Data with time-dependent covariates that should be modeled as cumulative effects](data-transformation.html#data-with-time-dependent-covariates-with-cumulative-effects)


## Data without time-dependent covariates

In this simple case the data transformation is relatively straight forward and
handled by the `as_ped` (**as** **p**iece-wise **e**xponential **d**ata) function.
This function internally calls `survival::survSplit`, thus all arguments available
for the `survSplit` function are also available to the `as_ped` function.


### Using defaults

As an example data set we first consider Veterans’ Administration lung cancer
study [@Kalbfleisch1980] available from the `survival` package:

```{r}
data("veteran", package = "survival")
veteran <- veteran %>%
  mutate(
    id    = row_number(),
    trt   = 1*(trt == 2),
    prior = 1*(prior != 0)) %>%
  select(id,everything())
head(veteran)
```

Each row contains information on the survival time (`time`), an event indicator
(`status`) and the `treatment` information (`6-MP` vs. `placebo`) as the only
covariate.

To transform the data into piece-wise exponential data (`ped`) format, we need to

  - define $J+1$ interval break points $t_{\min} = \kappa_0 < \kappa_1 < \cdots < \kappa_J = t_{\max}$

- create *pseudo*-observations for each interval $j = 1,\ldots, J$ in which subject
  $i, i = 1,\ldots,n$ was under risk.

Using the `pammtools` package this is easily achieved with the `as_ped` function
as follows:

```{r}
ped <- as_ped(Surv(time, status)~., data = veteran, id = "id")
ped %>% filter(id %in% c(42, 85, 112)) %>% select(-diagtime, -prior)
```

When no cut points are specified, the default is to use the unique event times.
As can be seen from the above output, the function creates an `id` variable,
indicating the subjects $i = 1,\ldots,n$, with one row per `id` for each interval
the subject "visited". Thus,

- subject `42` was under risk during 5 intervals,
- subject `85` was under risk during one interval,
- and subject `112` in 33 intervals.

In addition to the optional `id` variable the function also creates

  - `tstart`: the beginning of each interval
  - `tend`: the end of each interval
  - `interval`: a factor variable denoting the interval
  - `offset`:  the log of the duration during which the subject was under risk in
  that interval

Additionally the time-constant covariates (here: `trt`, `celltype`, `karno`, etc.)
are repeated $n_i$ number of times, where $n_i$ is the number of intervals during
which subject $i$ was in the risk set.


### Custom cut points
Per default the `as_ped` function uses all unique event times as cut points
(which is a sensible default as for example the (cumulative) baseline hazard
estimates in Cox-PH functions are updated also only at event times).

In some cases, however, one might want to reduce the number of cut points,
to reduce the size of the resulting data object and/or faster estimation times.
To do so the `cut` argument has to be specified.
Following the above example, we can use only 4 cut points
$\kappa_0 = 0, \kappa_1 = 50, \kappa_2 = 200, \kappa_3 = 400$,
resulting in $J = 3$ intervals:

```{r}
ped2 <- as_ped(Surv(time, status)~., data = veteran, cut = c(0, 50, 200, 400),
  id = "id")
ped2 %>% filter(id %in% c(42, 85, 112)) %>% select(-diagtime, -prior)
```
Note that now subjects `42` and `85` have only one row in the data set.
The fact that subject `42` was in the risk set longer than subject `85` is,
however, still accounted for by the `offset` variable. Note that the `offset`s
for subject `85` and subject `112` in interval `(50,200]` are only zero
because subject `85` had an observed event time of `1`, thus `log(1-0)=0` and
subject `112` had an event at time `51`, thus `log(51-50)=0`:

```{r}
veteran %>% slice(c(85, 112))
```

## Data with time-dependent covariates
In case of data with time-dependent covariates (that should not be
modeled as cumulative effects), we must split the follow-up at
desired cut-points and *additionally* at the time-points at which
the TDC changes its value.

We assume that the data is provided in two data sets, one that
contains time-to-event data and time-constant covariates and one data set
that contains information of time-dependent covariates.

For illustration, we use the `pbc` data from the `survival` package.

## Data with time-dependent covariates (with cumulative effects)

Here we demonstrate data transformation of data with TDCs
($\mathbf{z}=\{z(t_{e,q}),q=1,\ldots,Q\}$, that subsequently
will be modeled as cumulative effects defined as

$$
g(\mathbf{z}, t) = \int_{\mathcal{T}(t)} h(t, t_e, z(t_e))\mathrm{d}t_e
$$
Here

  - the three-variate function $h(t,t_e,z(t_e))$ defines the  so-called
  *partial effects* of the TDC $z(t_e)$ observed at exposure time
  $t_e$ on the hazard at time $t$ [@Bender2018]. The general partial effect definition
  given above is only one possibility, other common choices are<br>&nbsp;
      - $h(t-t_e)z(t_e)$ (This is the WCE model by @Sylvestre2009) <br>&nbsp;
      - $h(t-t_e, z(t_e))$ (This is the DLNM model by @Gasparrini2017)<br>&nbsp;

  - the cumulative effect $g(\mathbf{z}, t)$ at follow-up time $t$ is the
  integral (or sum) of the partial effects over exposure times $t_e$ contained
  within $\mathcal{T}(t)$<br>&nbsp;

  - the so called lag-lead window (or window of effectiveness) is denoted by
  $\mathcal{T}(t)$. This represents the set of exposure times at which exposures 
  can affect the hazard rate at time $t$. The most common definition is
  $\mathcal{T}(t) = \{t_{e,q}: t \geq t_{e,q}, q=1,\ldots, Q\}$,
  which means that all exposures that were observed prior to $t$ or at $t$ are
  eligible.

 As before we use the function `as_ped` to transform the data and additionally
use the  formula special `func` (for **fun**ctional **c**ovariate) to
specify the partial effect structure on the right-hand side of the formula.

Let $t$ (`time`) the follow up time, $\mathbf{z}_1$ (`z1`) and $\mathbf{z}_2$ (`z2`)
two TDCs observed at different exposure time grids $t_e$ (`te1`) and $s_e$ (`te2`)
with lag-lead windows $\mathcal{T}_1(t) = \{t_{e,q}: t \geq t_{e,q}\}$
and $\mathcal{T}_2(t) = \{s_{e,k}: t \geq s_{e,k} + 2\}$ (defined by
`ll2 <- function(t, te) { t >= te + 2}`).

The table below gives a selection of possible partial effects and the usage of
`func` to create matrices needed to estimate different types of cumulative
effects of $\mathbf{z}_1$ and $\mathbf{z}_2$:

|                                   partial effect                                   |                                `as_ped` specification                                |
|------------------------------------------------------------------------------------|--------------------------------------------------------------------------------------|
| $\int_{\mathcal{T}_1}h(t-t_e, z_1(t_e))$                                            | `func(latency(te1), z1, te_var= "te1")`                                              |
| $\int_{\mathcal{T}_1}h(t, t-t_e, z_1(t_e))$                                        | `func(time, latency(te1), z1, te_var="te1")`                                         |
| $\int_{\mathcal{T}_1}h(t, t_e, z_1(t_e))$                                          | `func(time, te1, z1, te_var="te1")`                                                  |
| $\int_{\mathcal{T}_1}h(t, t_e, z_1(t_e)) + \int_{\mathcal{T}_2}h(t-s_e, z_2(s_e))$ | `func(time, te1, z1, te_var="te1") + func(latency(te2), z2, te_var="te2", ll_fun=ll2)` |
| ...                                                                                | ...                                                                                     |


Note that

  - the variable representing follow-up time $t$ in `func` (here `time`) must
  match the time variable specified on the left-hand side of the `formula`

  - the variable representing exposure time $t_e$ (here `te1` and `te2`) must be
  wrapped within `latency()` to tell `as_ped` to calculate $t-t_e$

  - by default, $\mathcal{T}(t)$ is defined as `function(t,te) {t >= te}`,
  thus for $\mathcal{T}_1$ it is not necessary to explicitly specify the lag-lead window.
  In case you want to use a custom lag-lead window, provide the respective
  function to the `ll_fun` argument in `func` (see `ll2` in the examples above)

  - `func` makes no difference between partial effects such as $h(t-t_e, z(t_e))$
  and $h(t-t_e)z(t_e)$ as the necessary data transformation is the same in both
  cases. Later, however, when fitting the model, a distinction must be made

  - more than one $z$ variable can be provided to `func`, which can be convenient
  if multiple covariates should have the same time components and the
  same lag-lead window

  - multiple `func` terms can be specified, having different exposure times
  `t_e`, `s_e` and/or different lag-lead windows for different covariates
  $\mathbf{z}_1$, $\mathbf{z}_2$

  - To tell `func` which of the variables specified is the exposure time $t_e$,
  the `te_var` argument must be specified within each `func` term. The
  follow-up time component $t$ will be automatically recognized via the left-hand
  side of the `formula`.


### One time-dependent covariate

For illustration we use the ICU patients
data sets `patient` and `daily` provided in the **`pammtools`** package, with
follow-up time `survhosp` ($t$), exposure time `Study_Day` ($t_e$) and TDCs
`caloriesPercentage` $\mathbf{z}_1$ and `proteinGproKG` ($\mathbf{z}_2$):

```{r}
head(patient)
head(daily)
```

Below we illustrate the usage of `as_ped` and `func` to obtain covariate
matrices for the latency matrix of follow up time and exposure time and
a matrix of the TDC (`caloriesPercentage`). The follow up will be split
in 30 intervals with interval breaks points `0:30`. By default,
the exposure time provided to `te_var` will be used as a suffix for the names of the new
matrix columns to indicate the exposure they refer to,
however, you can override the suffix by specifying the
`suffix` argument to `func`.

```{r}
ped <- as_ped(
  data    = list(patient, daily),
  formula = Surv(survhosp, PatientDied) ~ . |
    func(latency(Study_Day), caloriesPercentage, te_var="Study_Day"),
  cut     = 0:30,
  id = "CombinedID")
str(ped, 1)
```

As you can see, the data is transformed to the PED format as before and
two additional matrix columns are added, `Study_Day_latency` and
`caloriesPercentage`.<br>&nbsp;


### Two time-dependent covariates observed on the same exposure time grid

Using the same data, we show a slightly more complex example with two
functional covariate terms with partial effects
$h(t, t-t_e, z_1(t_e))$ and $h(t-t_e, z_2(t_e))$.
As in the case of ICU patient data, both TDCs were observed on the same exposure
time grid, we want to use the same lag-lead window and the latency term
$t-t_e$ occurs in both partial effects, we only need to specify one `func` term.

```{r}
ped <- as_ped(
  data = list(patient, daily),
  formula = Surv(survhosp, PatientDied) ~ . |
    func(survhosp, latency(Study_Day), caloriesPercentage, proteinGproKG,
      te_var = "Study_Day"),
  cut = 0:30,
  id = "CombinedID")
str(ped, 1)
```

### Multiple TDCs observed at different exposure time grids

To illustrate data transformation when TDCs $\mathbf{z}_1$ and $\mathbf{z}_2$
were observed on different exposure time grids ($t_e$ and $s_e$) and
are assumed to have different lag_lead windows $\mathcal{T}_1(t)$ and
$\mathcal{T}_2(t)$, we use simulated data `simdf_elra` contained in
**`pammtools`** (see example in `sim_pexp` for data generation).

```{r}
simdf_elra
simdf_elra %>% slice(1) %>% select(id, te1) %>% unnest()
simdf_elra %>% slice(1) %>% select(id, te2) %>% unnest()
```

Note that `te1` has a maximum length of 10, while `te2` has a maximum length
of `11` and while `te1` was observed during the follow up, `te2` was partially
observed before the beginning of the follow up.

If we wanted to estimate the following cumulative effects

  - $g(\mathbf{z}_1, t) = \int_{\mathcal{T}_1(t)} h(t, t-t_e, z_1(s_e))$ and
  - $g(\mathbf{z}_2, t) = \int_{\mathcal{T}_2(t)} h(t-t_e, z_2(s_e))$

with $\mathcal{T}_1(t)  = \{t_e: t \geq t_e + 2\}$ and
$\mathcal{T}_2(t)  = \{s_e: t \geq s_e\}$ the data transformation function must
reflect the different exposure times as well as different lag-lead window
specifications as illustrated below:

```{r}
ped_sim <- as_ped(
  data = simdf_elra,
  formula = Surv(time, status)~.|func(time, latency(te1), z.te1, te_var="te1") +
    func(latency(te2), z.te2, te_var = "te2"),
  cut = 0:10,
  id = "id")
str(ped_sim,1)
```

Note how the matrix covariates associated with `te1` have 10 columns while
matrix covariates associated with `te2` have 11 columns. Note also that
the lag-lead matrices differ
```{r}
ped_sim %>% select(id, LL_te1) %>% head(16)
ped_sim %>% select(id, LL_te2) %>% head(16)
```

## References
