---
title: "Data Transformation"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Transformation}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: Remote.bib
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  fig.align  = "center",
  fig.width  = 4,
  fig.height = 4,
  crop       = TRUE)
```


```{r echo = TRUE, message = FALSE}
library(dplyr)
library(tidyr)
library(pammtools)
```


In this vignette we provide details on transforming data into a format suitable
to fit piece-wise exponential (additive) models (PAM). Three main cases need to be
distinguished

  1. [Data without time-dependent covariates](data-transformation.html#standard-time-to-event-data)

  2. [Left-truncated data](data-transformation.html#left-truncated-data)

  2. [Data with time-dependent covariates](data-transformation.html#data-with-time-dependent-covariates)

  3. [Data with time-dependent covariates that should be modeled as cumulative effects](data-transformation.html#data-with-time-dependent-covariates-with-cumulative-effects)


## Standard time-to-event data

In the case of "standard" time-to-event the data, the transformation is relatively straight forward and handled by the `as_ped` (**as** **p**iece-wise **e**xponential **d**ata) function. This function internally calls `survival::survSplit`, thus all arguments available for the `survSplit` function are also available to the `as_ped` function.


### Using defaults

As an example data set we first consider a data set of a stomach area tumor
available from the **`pammtools`** package. Patients were operated on to remove the tumor. Some patients experienced complications during the operation:

```{r}
data("tumor")
tumor <- tumor %>%
  mutate(id = row_number()) %>%
  select(id, everything())
head(tumor)
```

Each row contains information on the survival time (`days`), an event indicator
(`status`) as well as information about comorbidities (`charlson_score`), `age`, `sex`, and covariates that might be relevant like whether or not complications occurred (`complications`) and whether the cancer metastasized (`metastases`).

To transform the data into piece-wise exponential data (`ped`) format, we need to

  - define $J+1$ interval break points $t_{\min} = \kappa_0 < \kappa_1 < \cdots < \kappa_J = t_{\max}$

- create *pseudo*-observations for each interval $j = 1,\ldots, J$ in which subject
  $i, i = 1,\ldots,n$ was under risk.

Using the `pammtools` package this is easily achieved with the `as_ped` function
as follows:

```{r}
ped <- as_ped(Surv(days, status) ~ ., data = tumor, id = "id")
ped %>% filter(id %in% c(132, 141, 145)) %>%
  select(id, tstart, tend, interval, offset, ped_status, age, sex, complications)
```

When no cut points are specified, the default is to use the unique event times.
As can be seen from the above output, the function creates an `id` variable,
indicating the subjects $i = 1,\ldots,n$, with one row per interval that
the subject "visited". Thus,

- subject `132` was at risk during 4 intervals,
- subject `141` was at risk during 12 intervals,
- and subject `145` in 7 intervals.

In addition to the optional `id` variable the function also creates

  - `tstart`: the beginning of each interval
  - `tend`: the end of each interval
  - `interval`: a factor variable denoting the interval
  - `offset`:  the log of the duration during which the subject was under risk in
  that interval

Additionally the time-constant covariates (here: `age`, `sex`, `complications`,)
are repeated $n_i$ number of times, where $n_i$ is the number of intervals during
which subject $i$ was in the risk set.


### Custom cut points
Per default, the `as_ped` function uses all unique event times as cut points
(which is a sensible default as for example the (cumulative) baseline hazard
estimates in Cox-PH functions are also updated at event times). Also, PAMMs represent the baseline hazard via splines whose basis functions are evaluated at
the cut points. Using unique events times automatically increases the density
of cut points in relevant areas of the follow-up.

In some cases, however, one might want to reduce the number of cut points,
to reduce the size of the resulting data and/or faster estimation times.
To do so the `cut` argument has to be specified.
Following the above example, we can use only 4 cut points
$\kappa_0 = 0, \kappa_1 = 3, \kappa_2 = 7, \kappa_3 = 15$,
resulting in $J = 3$ intervals:

```{r}
ped2 <- as_ped(Surv(days, status) ~ ., data = tumor, cut = c(0, 3, 7, 15), id = "id")
ped2 %>%
  filter(id %in% c(132, 141, 145)) %>%
  select(id, tstart, tend, interval, offset, ped_status, age, sex, complications)
```
Note that now subjects `141` and `145` have both three rows in the data set.
The fact that subject `141` was in the risk set of the last interval for a longer time than subject `145` is, however, still accounted for by the `offset` variable.
Note that the `offset` for subject `145` is only `0`, because the last interval for this subject starts at $t=7$ and the subject experienced and event at $t=8$, thus $\log(8-7) = 0$

```{r}
tumor %>% slice(c(85, 112))
```


## Left-truncated Data
In left truncated data, subjects enter the risk set at some time-point unequal to zero. Such data usually ships in the so called start-stop notation, where for each subject the start time specifies the left-truncation time (can also be 0) and the
stop time the time-to-event, which can be right-censored.

For an illustration, we use a data set on infant mortality and maternal death
available in the **`eha`** package (see `?eha::infants`) for details.

```{r}
data(infants, package = "eha")
head(infants)
```

As children were included in the study after death of their mother, their survival
time is left-truncated at the entry time (variable `enter`). When transforming such data to the PED format, we need to create a interval cut-point at each entry time.
In general, each interval has to be constructed such that only subjects already at risk have an entry in the PED data. The previously introduced function `as_ped` can be also used for this setting, however, the LHS of the formula must be in start-stop format:

```{r}
infants_ped <- as_ped(Surv(enter, exit, event) ~ mother + age, data = infants)
infants_ped %>% filter(id == 4)
```

For analysis of such data refer to the [Left-truncation vignette](left-truncation.html)

## Data with time-dependent covariates
In case of data with time-dependent covariates (that should not be
modeled as cumulative effects), the follow-up is usually split at
desired cut-points and *additionally* at the time-points at which
the TDC changes its value.

We assume that the data is provided in two data sets, one that
contains time-to-event data and time-constant covariates and one data set
that contains information of time-dependent covariates.

For illustration, we use the `pbc` data from the `survival` package.
Before the actual data transformation we perform the data preprocessing
in `vignette("timedep", package="survival")`:

```{r}
data("pbc", package = "survival") # loads pbc and pbcseq
pbc <- pbc %>% filter(id <= 312) %>%
  mutate(status = 1L * (status == 2)) %>%
  select(id:sex, bili, protime)
```

We want to transform the data such that we could fit a concurrent effect of
the time-dependent covariates `bili` and `protime`, therefore, the RHS of
the formula passed to `as_ped` needs to contain an additional component,
separated by `|` and all variables for which should be transformed to the
concurrent format specified within the `concurrent` special. The data sets
are provided within a list. In `concurrent`, we also have to specify the name of
the variable which stores information on the time points at which the TDCs are
updated.

```{r}
ped_pbc <- as_ped(
  data    = list(pbc, pbcseq),
  formula = Surv(time, status) ~ . + concurrent(bili, protime, tz_var = "day"),
  id      = "id")
```

Multiple data sets can be provided within the list and multiple `concurrent`
terms with different `tz_var` arguments can be specified with different
lags. This will increase the data set, as an (additional) split point will be generated for each (lagged) time-point at which the TDC was observed:

```{r, eval = FALSE}
pbc_ped <- as_ped(data = list(pbc, pbcseq),
  formula = Surv(time, status) ~ . + concurrent(bili, tz_var = "day") +
    concurrent(protime, tz_var = "day", lag = 10),
    id = "id")
```

If different `tz_var` arguments are provided, the union of both will be used as
cut points in addition to usual cut points.

**Warning**: For time-points on the follow up that have no observation of the
TDC, the last observation will be carried forward until the next update
is available.


## Data with time-dependent covariates (with cumulative effects)

Here we demonstrate data transformation of data with TDCs
($\mathbf{z}=\{z(t_{z,q}),q=1,\ldots,Q\}$, that subsequently
will be modeled as cumulative effects defined as

$$
g(\mathbf{z}, t) = \int_{\mathcal{T}(t)} h(t, t_z, z(t_z))\mathrm{d}t_z
$$
Here

  - the three-variate function $h(t,t_z,z(t_z))$ defines the  so-called
  *partial effects* of the TDC $z(t_z)$ observed at exposure time
  $t_z$ on the hazard at time $t$ [@Bender2018]. The general partial effect definition
  given above is only one possibility, other common choices are<br>&nbsp;
      - $h(t-t_z)z(t_z)$ (This is the WCE model by @Sylvestre2009) <br>&nbsp;
      - $h(t-t_z, z(t_z))$ (This is the DLNM model by @Gasparrini2017)<br>&nbsp;

  - the cumulative effect $g(\mathbf{z}, t)$ at follow-up time $t$ is the
  integral (or sum) of the partial effects over exposure times $t_z$ contained
  within $\mathcal{T}(t)$<br>&nbsp;

  - the so called lag-lead window (or window of effectiveness) is denoted by
  $\mathcal{T}(t)$. This represents the set of exposure times at which exposures
  can affect the hazard rate at time $t$. The most common definition is
  $\mathcal{T}(t) = \{t_{z,q}: t \geq t_{z,q}, q=1,\ldots, Q\}$,
  which means that all exposures that were observed prior to $t$ or at $t$ are
  eligible.

As before we use the function `as_ped` to transform the data and additionally
use the  formula special `cumulative` (for **fun**ctional **c**ovariate) to
specify the partial effect structure on the right-hand side of the formula.

Let $t$ (`time`) the follow up time, $\mathbf{z}_1$ (`z1`) and $\mathbf{z}_2$ (`z2`)
two TDCs observed at different exposure time grids $t_z$ (`tz1`) and $s_z$ (`tz2`)
with lag-lead windows $\mathcal{T}_1(t) = \{t_{z,q}: t \geq t_{z,q}\}$
and $\mathcal{T}_2(t) = \{s_{z,k}: t \geq s_{z,k} + 2\}$ (defined by
`ll2 <- function(t, tz) { t >= tz + 2}`).

The table below gives a selection of possible partial effects and the usage of
`cumulative` to create matrices needed to estimate different types of cumulative
effects of $\mathbf{z}_1$ and $\mathbf{z}_2$:

|                                   partial effect                                   |                                `as_ped` specification                                |
|------------------------------------------------------------------------------------|--------------------------------------------------------------------------------------|
| $\int_{\mathcal{T}_1}h(t-t_z, z_1(t_z))$                                            | `cumulative(latency(tz1), z1, tz_var= "tz1")`                                              |
| $\int_{\mathcal{T}_1}h(t, t-t_z, z_1(t_z))$                                        | `cumulative(time, latency(tz1), z1, tz_var="tz1")`                                         |
| $\int_{\mathcal{T}_1}h(t, t_z, z_1(t_z))$                                          | `cumulative(time, tz1, z1, tz_var="tz1")`                                                  |
| $\int_{\mathcal{T}_1}h(t, t_z, z_1(t_z)) + \int_{\mathcal{T}_2}h(t-s_z, z_2(s_z))$ | `cumulative(time, tz1, z1, tz_var="tz1") + cumulative(latency(tz2), z2, tz_var="tz2", ll_fun=ll2)` |
| ...                                                                                | ...                                                                                     |


Note that

  - the variable representing follow-up time $t$ in `cumulative` (here `time`) must
  match the time variable specified on the left-hand side of the `formula`

  - the variable representing exposure time $t_z$ (here `tz1` and `tz2`) must be
  wrapped within `latency()` to tell `as_ped` to calculate $t-t_z$

  - by default, $\mathcal{T}(t)$ is defined as `function(t, tz) {t >= tz}`,
  thus for $\mathcal{T}_1$ it is not necessary to explicitly specify the lag-lead window.
  In case you want to use a custom lag-lead window, provide the respective
  function to the `ll_fun` argument in `cumulative` (see `ll2` in the examples above)

  - `cumulative` makes no difference between partial effects such as $h(t-t_z, z(t_z))$
  and $h(t-t_z)z(t_z)$ as the necessary data transformation is the same in both
  cases. Later, however, when fitting the model, a distinction must be made

  - more than one $z$ variable can be provided to `cumulative`, which can be convenient
  if multiple covariates should have the same time components and the
  same lag-lead window

  - multiple `cumulative` terms can be specified, having different exposure times
  `t_z`, `s_z` and/or different lag-lead windows for different covariates
  $\mathbf{z}_1$, $\mathbf{z}_2$

  - To tell `cumulative` which of the variables specified is the exposure time $t_z$,
  the `tz_var` argument must be specified within each `cumulative` term. The
  follow-up time component $t$ will be automatically recognized via the left-hand
  side of the `formula`.


### One time-dependent covariate

For illustration we use the ICU patients
data sets `patient` and `daily` provided in the **`pammtools`** package, with
follow-up time `survhosp` ($t$), exposure time `Study_Day` ($t_z$) and TDCs
`caloriesPercentage` $\mathbf{z}_1$ and `proteinGproKG` ($\mathbf{z}_2$):

```{r}
head(patient)
head(daily)
```

Below we illustrate the usage of `as_ped` and `cumulative` to obtain covariate
matrices for the latency matrix of follow up time and exposure time and
a matrix of the TDC (`caloriesPercentage`). The follow up will be split
in 30 intervals with interval breaks points `0:30`. By default,
the exposure time provided to `tz_var` will be used as a suffix for the names of the new
matrix columns to indicate the exposure they refer to,
however, you can override the suffix by specifying the
`suffix` argument to `cumulative`.

```{r}
ped <- as_ped(
  data    = list(patient, daily),
  formula = Surv(survhosp, PatientDied) ~ . +
    cumulative(latency(Study_Day), caloriesPercentage, tz_var = "Study_Day"),
  cut     = 0:30,
  id = "CombinedID")
str(ped, 1)
```

As you can see, the data is transformed to the PED format as before and
two additional matrix columns are added, `Study_Day_latency` and
`caloriesPercentage`.<br>&nbsp;


### Two time-dependent covariates observed on the same exposure time grid

Using the same data, we show a slightly more complex example with two
functional covariate terms with partial effects
$h(t, t-t_z, z_1(t_z))$ and $h(t-t_z, z_2(t_z))$.
As in the case of ICU patient data, both TDCs were observed on the same exposure
time grid, we want to use the same lag-lead window and the latency term
$t-t_z$ occurs in both partial effects, we only need to specify one `cumulative` term.

```{r}
ped <- as_ped(
  data = list(patient, daily),
  formula = Surv(survhosp, PatientDied) ~ . +
    cumulative(survhosp, latency(Study_Day), caloriesPercentage, proteinGproKG,
      tz_var = "Study_Day"),
  cut = 0:30,
  id = "CombinedID")
str(ped, 1)
```

### Multiple TDCs observed at different exposure time grids

To illustrate data transformation when TDCs $\mathbf{z}_1$ and $\mathbf{z}_2$
were observed on different exposure time grids ($t_z$ and $s_z$) and
are assumed to have different lag_lead windows $\mathcal{T}_1(t)$ and
$\mathcal{T}_2(t)$, we use simulated data `simdf_elra` contained in
**`pammtools`** (see example in `sim_pexp` for data generation).

```{r}
simdf_elra
simdf_elra %>% slice(1) %>% select(id, tz1) %>% unnest()
simdf_elra %>% slice(1) %>% select(id, tz2) %>% unnest()
```

Note that `tz1` has a maximum length of 10, while `tz2` has a maximum length
of `11` and while `tz1` was observed during the follow up, `tz2` was partially
observed before the beginning of the follow up.

If we wanted to estimate the following cumulative effects

  - $g(\mathbf{z}_1, t) = \int_{\mathcal{T}_1(t)} h(t, t-t_z, z_1(s_z))$ and
  - $g(\mathbf{z}_2, t) = \int_{\mathcal{T}_2(t)} h(t-t_z, z_2(s_z))$

with $\mathcal{T}_1(t)  = \{t_z: t \geq t_z + 2\}$ and
$\mathcal{T}_2(t)  = \{s_z: t \geq s_z\}$ the data transformation function must
reflect the different exposure times as well as different lag-lead window
specifications as illustrated below:

```{r}
ped_sim <- as_ped(
  data = simdf_elra,
  formula = Surv(time, status) ~ . +
    cumulative(time, latency(tz1), z.tz1, tz_var = "tz1") +
    cumulative(latency(tz2), z.tz2, tz_var = "tz2"),
  cut = 0:10,
  id = "id")
str(ped_sim, 1)
```

Note how the matrix covariates associated with `tz1` have 10 columns while
matrix covariates associated with `tz2` have 11 columns. Note also that
the lag-lead matrices differ
```{r, fig.width = 8, out.width = "600px"}
gg_laglead(ped_sim)
```

## References
