---
title: "Data Transformation"
author: "Andreas Bender"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{data-transformation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: pam.bib
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  fig.align  = "center",
  fig.width  = 4,
  fig.height = 4,
  crop       = TRUE)
```


```{r echo = TRUE, message = FALSE}
library(magrittr)
library(dplyr)
library(pam)
```


In this vignette we provide details on transforming data into a format suitable 
to fit piece-wise exponential (additive) models (PAM). Three main cases need to be 
distinguished

  1. Data without time-dependent covariates 

  2. Data with time-dependent covariates 

  3. Data with time-dependent covariates that should be modeled as cumulative 
  effects


## Data without time-dependent covariates 

In this simple case the data transformation is relatively straight forward and 
handeled by the `split_data` function. This function internally calls 
`survival::survSplit`, thus all arguments available for the `survSplit` function 
are also available to the `split_data` function. 


### Using defaults

As an example data set we first consider Veterans’ Administration lung cancer 
study [@Kalbfleisch1980] available from the `survival` package:

```{r}
data("veteran", package = "survival")
veteran %<>%
  mutate(
    id = row_number(),
    trt = 1*(trt == 2),
    prior = 1*(prior != 0)) %>% 
  select(id,everything())
head(veteran)
```

Each row contains information on the survival time (`time`), an event indicator
(`status`) and the `treatment` information (`6-MP` vs. `placebo`) as the only 
covariate. 

To transform the data into piece-wise exponential data (`ped`) format, we need to 

  - define $J+1$ interval break points $t_{\min} = \kappa_0 < \kappa_1 < \cdots < \kappa_J = t_{\max}$ 
  
- create *pseudo*-observations for each interval $j = 1,\ldots, J$ in which subject 
  $i, i = 1,\ldots,n$ was under risk. 

Using the `pam` package this is easily achieved with the `split_data` function
as follows: 

```{r}
ped <- split_data(Surv(time, status)~., data = veteran, id = "id")
ped %>% filter(id %in% c(42, 85, 112)) %>% select(-diagtime, -prior)
```

When no cut points are specified, the default is to use the unique event times. 
As can be seen from the above output, the function creates an `id` variable, 
indicating the subjects $i = 1,\ldots,n$, with one row per `id` for each interval 
the subject "visited". Thus,  

- subject `42` was under risk during 5 intervals, 
- subject `85` was under risk during one interval, 
- and subject `112` in 33 intervals. 

In addition to the optional `id` variable the function also creates 

  - `tstart`: the beginning of each interval
  - `tend`: the end of each interval
  - `interval`: a factor variable denoting the interval
  - `offset`:  the log of the duration during which the subject was under risk in 
  that interval
  
Additionally the time-constant covariates (here: `trt`, `celltype`, `karno`, etc.) 
are repeated $n_i$ number of times, where $n_i$ is the number of intervals during 
which subject $i$ was in the risk set.


### Custom cut points 
Per default the `split_data` function uses all unique event times as cut points 
(which is a sensible default as for example the (cumulative) baseline hazard 
estimates in Cox-PH functions are updated also only at event times). 

In some cases, however, one might want to reduce the number of cut points, 
to reduce the size of the resulting data object and/or faster estimation times. 
To do so the `cut` argument has to be specified. 
Following the above example, we can use only 4 cut points
$\kappa_0 = 0, \kappa_1 = 50, \kappa_2 = 200, \kappa_3 = 400$, 
resulting in $J = 3$ intervals:

```{r}
ped2 <- split_data(Surv(time, status)~., data = veteran, cut = c(0, 50, 200, 400), 
  id = "id")
ped2 %>% filter(id %in% c(42, 85, 112)) %>% select(-diagtime, -prior)
```
Note that now subjects `42` and `85` have only one row in the data set. 
The fact that subjects `42` was in the risk set longer than subject `85` is,
however, still accounted for by the `offset` variable. Note that the `offset`s
for subject `85` and subject `112` in interval `(50,200]` are only zero 
because subject `85` had an observed event time of `1`, thus `log(1-0)=0` and 
subject `112` had an event at time `51`, thus `log(51-50)=0`: 

```{r}
veteran %>% slice(c(85, 112))
```
 
## References
