---
title: "Example MGUS"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{splines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(
  fig.align = "center",
  cache=TRUE, 
  message=FALSE, 
  fig.height=5, 
  fig.width=5
)
```

# Spline terms 
## Example MGUS data 

In this vignette we show the usage of spline smooth terms using example data 
presented in the respective vignette in the `survival` package 
(see `vignette("splines", package="survival")`)

The example presented in the vignette goes as follows, using the standard 
base R workflow and `termplot` for visualization: 

```{r}
library(pam)
library(mgcv)
library(survival)
data("mgus", package="survival")
mfit <- coxph(Surv(futime, death)~sex + pspline(age, df=4), data=mgus)
mfit
termplot(mfit, term=2, se=TRUE, col.term=1, col.se=1)
```

The equivalent fit using PAMs requieres one additional step of transforming the 
data into suitable format, afterwards we use the `mgcv::gam` function to fit 
the model: 

```{r}
mgus.ped <- split_data(Surv(futime, death)~sex + age, data=mgus, id="id")
head(mgus.ped)
pamfit <- gam(status ~ s(tend) + sex + s(age, bs="ps"), data = mgus.ped, method="REML", 
	family=poisson(), offset=offset)
summary(pamfit)
```

For visualization of the smooth effects we can use the default `mgcv::plot.gam` 
function: 

```{r, fig.width = 7, fig.height=3.5}
layout(matrix(1:2, nrow=1))
termplot(mfit, term=2, se=TRUE, col.term=1, col.se=1)
plot(pamfit, select=2, se=TRUE, ylim=c(-4, 3.5))
```

In this example the PAM approach estimates an linear effect of age, which is 
consistet with the estimation using `coxph`, as there is not much non-linearity.

Another example from the same vignette shows that estimated non-linear effect 
are very similar, when the effect of the covariate is in fact non-linear: 

```{r}
fit <- coxph(Surv(futime, death) ~ age + pspline(hgb, 4), mgus2)
mgus2.ped <- split_data(Surv(futime, death)~age + hgb, data=mgus2, id="id")
pamfit2 <- gam(status~s(tend) + age + s(hgb), data=mgus2.ped, 
	family=poisson(), offset=offset)
```
  
```{r, fig.width=7, fig.height=3.5}
layout(matrix(1:2, nrow=1))
termplot(fit, term=2, se=TRUE, col.term=1, col.se=1)
plot(pamfit2, select=2, se=TRUE, ylim=c(-0.5, 2))
```
