---
title: "Frailty and random effects"
author: "Andreas Bender"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{frailty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: pam.bib
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  fig.align = "center",
  fig.width = 4, 
  fig.height = 4,
  crop=TRUE)
```

```{r message=FALSE}
library(magrittr)
library(dplyr)
library(survival)
library(coxme)
library(pam)
library(mgcv)
```

Whenever subjects belonging to a cluster or group could have correlated outcomes,
our models must account for such dependency structures. 
We can do so by including random effects in the model. 
In the context of survival analysis such effects are called "frailty" terms. 

## NCCTG Lung Cancer Data

For illustration we consider the NCTG Lung Cancer Data [@Loprinzi1994] that 
is contained in the `survival` package [@Thernau2015]. A description
of the data is provided in `?lung`.

The data contain survival times (in days) of advanced lung cancer patients 
from different institutions: 

```{r}
lung %<>% mutate(inst = as.factor(inst))
head(lung)

## 228 patients from 18 institutions
nrow(lung)
lung %>% group_by(inst) %>% summarize(n=n())
```


## Cox model with gaussian frailty 

The penalized Cox model with frailty terms has been described in @Thernau2003
and implemented in the `coxme` package for Gaussian frailties. 
Below we apply the `coxme` function to the `lung` data:

```{r}
cme <- coxme(Surv(time, status) ~ ph.ecog + (1|inst), data=lung)
coef(cme)
summary(cme$frail$inst)
```

- The effect of `ph.ecog` was estimated to be about `0.5` on the log-hazard scale. 
- The estimated random intercepts ("frailties"") for the different institution are all relatively close to zero and have small standard deviation: `sd(cme$frail$inst)=``r round(sd(cme$frail$inst), 3)`,

## PAM with gaussian frailty

For an equivalent model using PAMs we first transform the data set and 
then call `mgcv:::gam` for the fit. To include a gaussian random effect 
we set the `bs` argument in the respective penalized `s()` term to `bs='re'` (see description 
in `?smooth.terms`): 

```{r}
ped <- split_data(Surv(time, status)~., data=lung, id="id")
pam <- gam(ped_status ~ s(tend) + ph.ecog + s(inst, bs="re"), 
	data=ped, family=poisson(), offset=offset)
```

Comparison of the estimates of the two models shows that they are in close 
agreement (we use `tidy_re` from the `pam` package to extract random effects):

```{r}
cbind(pam = coef(pam)[2], cox=coef(cme))
re <- tidy_re(pam)
plot(cme$frail$inst, re$fit, las=1, xlab="Frailty (cox)", ylab="Frailty (PAM)")
abline(0, 1)
```

The `pam` package also provides a convenience function for a quantile-quantile plot of the estimated frailties to check the Gaussian assumption: 

```{r}
gg_re(pam)
```




## References 