---
title: "Interval-Censored Data"
author: "pammtools"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: Remote.bib
vignette: >
  %\VignetteIndexEntry{Interval-Censored Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  fig.align  = "center",
  cache      = FALSE,
  message    = FALSE,
  warning    = FALSE,
  fig.height = 4,
  fig.width  = 6
)
```

```{r, message = FALSE}
library(pammtools)
library(survival)
library(dplyr)
library(ggplot2)
```

## Background

In many studies the event time is not observed exactly but only known to lie in
an interval $(L_i, R_i]$ -- for example because subjects are examined at a
sequence of *inspection times* and we only learn that the event happened between
the last clean and the first positive visit. This is **interval censoring**.

A piecewise-exponential additive model (PAMM) assumes a piecewise-constant
hazard, so the conditional distribution of the event time within $(L_i, R_i]$ is
known in closed form. `pammtools` exploits this with a **multiple imputation
(MI)** strategy [@pan2000]: exact event times are
repeatedly drawn from $p(T \mid L_i < T \le R_i, x_i, \theta)$, each completed
data set is transformed and re-fit with the *standard* right-censored pipeline,
and the fits are pooled for inference following Rubin's rules [@rubin1987].
Naive midpoint imputation, by contrast, typically yields reasonable point
estimates but invalid (too narrow) confidence intervals. Related model-based
imputation schemes for the interval-censored hazard are discussed by
@bebchuk2000; the `icenReg` package [@andersonbergman2017] provides
non-parametric and parametric alternatives that are useful as a cross-check.

Interval-censored data are specified through the usual **survival** interface,
`Surv(L, R, type = "interval2")`, where `L == R` denotes an exact time,
`R == Inf` a right-censored observation, `L == 0` a left-censored one, and
`0 < L < R < Inf` a genuine interval.

## Simulating interval-censored data

`add_inspections()` turns exact simulated times (here from `sim_pexp()`) into
interval-censored panel data by drawing (per-subject, by default) inspection times. The true
event time is retained in `true_time` so we can check our work.

Here a binary covariate `trt` acts through a *time-varying* effect that is
**absent at baseline and grows over follow-up**. The control group (`trt == 0`)
has a *constant* hazard; the other group starts at the same hazard but its
log-hazard then rises linearly in time -- the log-hazard ratio is `0.15 * t`,
zero at `t = 0` -- so the two hazards start together and fan apart, an
increasingly harmful effect and a clear departure from proportional hazards.
Because factors enter the (numeric) linear predictor directly, the effect is
written with an explicit indicator `(trt == 1)`.

The inspection `rate` controls how much information the interval censoring
destroys: a dense grid pins each event down tightly and leaves little to impute,
a sparse one brackets events only loosely. We use a moderate schedule
(`rate = 1.5`, mean inter-inspection gap $2/3$) -- dense enough that these simple
hazards are recovered well, yet still leaving each event bracketed in an interval
that straddles several analysis cut-points, so the gap between proper MI and
naive midpoint intervals stays appreciable.

```{r}
set.seed(3)
df  <- data.frame(trt = rbinom(1000, 1, 0.5) |> as.factor())
sdf <- sim_pexp(~ -2.2 + 0.15 * t * (trt == 1), df,
                cut = seq(0, 10, by = 0.1))
icd <- add_inspections(sdf, rate = 1.5, max_time = 10)
icd %>% select(id, trt, true_time, L, R) %>% head()
```

## Fitting a PAMM by multiple imputation

`pamm_ic()` runs the whole MI workflow: an initial midpoint fit, then `m`
rounds of (proper) imputation and re-fitting on a fixed interval grid.

We give each group its own smooth log-baseline-hazard via `s(tend, by = trt)`
(a factor `by`-variable yields one centred smooth per level) plus the parametric
`trt` main effect for the overall level difference between groups.

```{r}
fit <- pamm_ic(
  Surv(L, R, type = "interval2") ~ trt,
  data          = icd,
  model_formula = ped_status ~ trt + s(tend, by = trt),
  cut           = seq(0, 10, by = 0.5),
  m             = 20
)
fit
```

`print()` gives a compact overview; `summary()` reports the **pooled** inference:
parametric coefficients and covariances are combined with Rubin's rules (so the
reported standard errors include the between-imputation variability), term
p-values use the median-p rule [@bolt2022], and the fraction of missing
information induced by the interval censoring is reported as well.

```{r}
summary(fit)
```

The `m` imputation fits are stored in `fit$fits` in a slimmed form (the
per-observation slots are dropped, so memory does not grow with the data set
size), and pooled summary diagnostics live in `fit$pooled`. Because smooth bases
can differ across imputed data sets, `fit$pooled` is not a single `gam` object
for direct `predict()` or `plot()` calls. The pooled `add_*()` methods below
operate on the imputation fits directly: they combine the per-imputation
posterior draws so the intervals reflect both within- and between-imputation
uncertainty.

The **fraction of missing information** (FMI) is a Rubin-style diagnostic for
how much of an estimate's uncertainty is due to the unobserved exact event times.
For a scalar quantity $Q$, write $W$ for the average within-imputation variance
and $B$ for the between-imputation variance of the point estimates. The relative
increase in variance is $(1 + 1/m)B/W$, and the FMI is the corresponding fraction
of the total uncertainty attributable to the missing information (with the usual
small-sample correction for finite `m`). Values near zero mean that interval
censoring adds little uncertainty for that quantity; values near one mean the
imputed event times dominate the uncertainty. `summary()` reports this for the
parametric coefficients, and for smooth terms it reports five-number summaries
over the training PED grid. Below we compute the same diagnostic directly for the
hazard rate, point-by-point on the plotting grid.

FMI is also useful for choosing the number of imputations. Small values of `m`
such as 5 or 10 are fine while developing the model, but reported analyses should
usually start around `m = 20`. If the reported FMI values are mostly below 0.10,
that is often enough; if they are around 0.10--0.30, use roughly 30--50
imputations; if they are around 0.30--0.50, use roughly 50--100; and if they are
larger than 0.50, use at least 100 and check sensitivity to a larger `m` or a
different random seed. The classical relative-efficiency approximation,
$(1 + \lambda/m)^{-1}$ for FMI $\lambda$, explains why moderate `m` can already
be efficient on average, but plotted smooths and interval endpoints can still
show Monte Carlo noise. Keep this separate from `nsim`: `m` controls imputation
uncertainty, whereas `nsim` controls the simulation noise in posterior-draw
intervals for plotted quantities of interest.

We compare two PAMM estimates per group. The **MI** estimate pools the imputation
fits, so its intervals carry both the within-fit and the between-imputation
uncertainty. The **midpoint** estimate is the initialiser fit (`fit$init_fit`):
a single PAMM fit to the interval midpoints, which treats those imputed times as
if observed. We request **simulation-based** intervals (`ci_type = "sim"`) for the
midpoint fit so that both estimates use the same (proper) interval machinery as
the pooled `pamm_ic` methods -- the pooled methods always simulate from each fit's
posterior, so this is a like-for-like comparison that differs only by the
between-imputation variance. (The default `ci_type = "default"` instead cumulates
the pointwise hazard limits, which yields a noticeably more conservative band.)

```{r}
ped <- as_ped(icd, Surv(L, R, type = "interval2") ~ trt, cut = seq(0, 10, by = 0.5))
nd  <- make_newdata(ped, tend = unique(tend), trt = unique(trt)) %>% group_by(trt)

# midpoint = initialiser fit (single completed data set, no imputation uncertainty)
set.seed(1)   # the simulation-based intervals draw from each fit's posterior
surv_mid   <- add_surv_prob(nd, fit$init_fit, ci_type = "sim", nsim = 1000)
surv_mi    <- add_surv_prob(nd, fit, nsim = 1000)
hazard_mid <- add_hazard(nd, fit$init_fit, ci_type = "sim", nsim = 1000)
hazard_mi  <- add_hazard(nd, fit, nsim = 1000)
```

```{r, echo = FALSE}
qoi_fmi <- function(qhat, uhat) {
  m <- ncol(qhat)
  if (m <= 1) {
    return(rep(NA_real_, nrow(qhat)))
  }

  W <- rowMeans(uhat)
  B <- apply(qhat, 1, stats::var)

  riv <- rep(NA_real_, length(W))
  ok <- is.finite(W) & W > 0 & is.finite(B)
  riv[ok] <- (1 + 1 / m) * B[ok] / W[ok]
  riv[is.finite(W) & W <= 0 & is.finite(B) & B == 0] <- 0
  riv[is.finite(W) & W <= 0 & is.finite(B) & B > 0] <- Inf

  fmi <- rep(NA_real_, length(riv))
  fmi[is.finite(riv) & riv == 0] <- 0
  finite <- is.finite(riv) & riv > 0
  dfbr <- (m - 1) * (1 + 1 / riv[finite])^2
  fmi[finite] <- (riv[finite] + 2 / (dfbr + 3)) / (riv[finite] + 1)
  fmi[is.infinite(riv) & riv > 0] <- 1

  pmin(pmax(fmi, 0), 1)
}

hazard_fmi <- function(object, newdata) {
  hazard_pieces <- vapply(object$fits, function(f) {
    X <- mgcv::predict.gam(f, newdata = newdata, type = "lpmatrix")
    eta <- as.numeric(X %*% stats::coef(f))
    hazard <- exp(eta)
    var_eta <- pmax(rowSums((X %*% f[["Vp"]]) * X), 0)
    cbind(hazard = hazard, variance = hazard^2 * var_eta)
  }, matrix(0, nrow(newdata), 2))

  out <- as.data.frame(newdata)
  out$fmi <- qoi_fmi(
    hazard_pieces[, "hazard", ],
    hazard_pieces[, "variance", ]
  )
  out
}

haz_fmi <- hazard_fmi(fit, nd) %>%
  mutate(trt = paste0("trt = ", trt), panel = "Hazard FMI")
```

As a non-parametric cross-check we add the **Turnbull** estimator -- the NPMLE of
the survival function for interval-censored data -- computed per group.
`survival::survfit()` runs the Turnbull EM algorithm when handed an `interval2`
response and returns pointwise confidence limits.

```{r}
tb    <- survfit(Surv(L, R, type = "interval2") ~ trt, data = icd)
tb_df <- data.frame(
  tend  = tb$time,
  surv  = tb$surv,
  lower = tb$lower,
  upper = tb$upper,
  trt   = factor(rep(c(0, 1), tb$strata))   # strata are ordered trt = 0, then 1
)
```

All three recover the diverging survival curves well -- the groups start together
and the increasingly harmful effect opens a growing gap. The midpoint intervals
are the tightest -- they ignore the imputation uncertainty and are therefore
over-confident; the MI intervals are wider (the censoring here induces a sizeable
fraction of missing information, see `summary(fit)` above), and the
non-parametric Turnbull intervals are wider and more jagged still.

```{r, fig.width = 9, fig.height = 3.6}
prep <- function(d, method) {
  transmute(d, tend, trt = factor(trt),
    surv = surv_prob, lower = surv_lower, upper = surv_upper, method = method)
}
surv_all <- bind_rows(
  prep(surv_mid, "PAMM (midpoint)"),
  prep(surv_mi,  "PAMM (MI)"),
  transmute(tb_df, tend, trt, surv, lower, upper, method = "Turnbull"))
surv_all$method <- factor(surv_all$method,
  levels = c("PAMM (midpoint)", "PAMM (MI)", "Turnbull"))

ggplot(surv_all, aes(x = tend, y = surv, color = trt, fill = trt)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, color = NA) +
  geom_line() +
  facet_grid(~ method) +
  scale_color_viridis_d(end = 0.8, aesthetics = c("colour", "fill")) +
  labs(x = "time", y = "S(t)", color = "trt", fill = "trt",
    title = "Interval-censored S(t): midpoint vs. multiple imputation vs. Turnbull")
```

The over-confidence of the midpoint fit is easy to quantify. The ratio below is
`MI interval width / midpoint interval width`, so values above one mean that
proper MI gives wider intervals:

```{r}
# MI-CI-width / midpoint-CI-width
summary((surv_mi$surv_upper  - surv_mi$surv_lower) / 
        (surv_mid$surv_upper - surv_mid$surv_lower))
```

The effect is more visible on the **hazard** scale, which depends directly on
exact event timing which interval censoring leaves more uncertain. Overlaying
the two methods within each group shows the MI hazard bands enveloping
the (over-confident) midpoint-based bands, while the point estimates mostly agree:

```{r, fig.width = 8, fig.height = 7.2}
haz_panels <- c("Hazard rate", "Hazard FMI", "Observed censoring")
haz_all <- bind_rows(
  mutate(hazard_mid, method = "PAMM (midpoint)"),
  mutate(hazard_mi,  method = "PAMM (MI)")) %>%
  mutate(trt = paste0("trt = ", trt),
    method = factor(method, levels = c("PAMM (midpoint)", "PAMM (MI)")),
    panel = factor("Hazard rate", levels = haz_panels))

# true conditional hazards from the data-generating log-hazard
tgrid <- seq(0, 10, by = 0.1)
haz_true <- bind_rows(
  data.frame(tend = tgrid, trt = "trt = 0", hazard = exp(-2.2)),
  data.frame(tend = tgrid, trt = "trt = 1", hazard = exp(-2.2 + 0.15 * tgrid))) %>%
  mutate(panel = factor("Hazard rate", levels = haz_panels))

haz_fmi <- haz_fmi %>%
  mutate(panel = factor(panel, levels = haz_panels))

censor_panel <- icd %>%
  mutate(
    trt = paste0("trt = ", trt),
    panel = factor("Observed censoring", levels = haz_panels),
    kind = case_when(
      L == R ~ "Exact",
      is.infinite(R) ~ "Right-censored",
      TRUE ~ "Interval"
    ),
    xmin = pmax(L, 0),
    xmax = pmin(ifelse(is.infinite(R), 10, R), 10)
  ) %>%
  group_by(trt) %>%
  arrange(xmin, xmax, .by_group = TRUE) %>%
  mutate(subject = row_number()) %>%
  ungroup()

ggplot() +
  geom_stepribbon(data = haz_all,
    aes(x = tend, ymin = ci_lower, ymax = ci_upper, fill = method),
    alpha = 0.3, color = NA) +
  geom_step(data = haz_all,
    aes(x = tend, y = hazard, color = method)) +
  geom_line(data = haz_true, aes(x = tend, y = hazard), inherit.aes = FALSE,
    color = "gold", alpha = 0.5, linewidth = 0.9, linetype = 2) +
  geom_segment(data = filter(censor_panel, kind != "Exact"),
    aes(x = xmin, xend = xmax, y = subject, yend = subject),
    inherit.aes = FALSE, color = "grey35", alpha = 0.22, linewidth = 0.25) +
  geom_point(data = filter(censor_panel, kind == "Exact"),
    aes(x = xmin, y = subject),
    inherit.aes = FALSE, color = "grey10", alpha = 0.5, size = 0.45) +
  geom_step(data = haz_fmi, aes(x = tend, y = fmi),
    inherit.aes = FALSE, color = "black", linewidth = 0.8) +
  facet_grid(panel ~ trt, scales = "free_y") +
  scale_color_viridis_d(end = 0.5, aesthetics = c("colour", "fill"), option = "C") +
  labs(x = "time", y = NULL, color = NULL, fill = NULL,
    title = "Conditional hazard rate and hazard-scale FMI\n(true rates in gold)")
```

The true rates (gold) are tracked closely across the whole range -- the constant
`trt = 0` hazard and the rising `trt = 1` hazard are both recovered well. The
midpoint and MI point estimates nearly coincide, consistent with midpoint giving
reasonable point estimates but over-confident intervals; the MI bands are visibly
wider throughout. The middle row shows *where* interval censoring bites on the
hazard scale: high FMI means that most uncertainty in the local hazard estimate
comes from not observing exact event times. The pattern is time-local, rather
than a single model-wide number, because the event-time ambiguity matters most
where the interval-censored observations leave the hazard shape weakly pinned
down. The bottom row shows the observed censoring pattern itself: each horizontal
line is one subject's observed interval (right-censored observations are drawn up
to the analysis horizon), and exact observations would appear as points.

The hazard interval-width ratio uses the same format as the survival calculation:

```{r}
# MI-CI-width / midpoint-CI-width
summary((hazard_mi$ci_upper  - hazard_mi$ci_lower) /
        (hazard_mid$ci_upper - hazard_mid$ci_lower))
```

So the imputation uncertainty inflates the hazard intervals more than the
survival intervals: $S(t)$ aggregates the hazard over time and the group-level
survival stays fairly well identified, whereas the hazard itself is tied directly
to exact event times, which interval censoring leaves uncertain (cf.\ the
large fraction of missing information reported by `summary()` above).  
The same machinery is available for `add_cumu_hazard()`.

## Sparse inspection schedules: iterated MI

By default all `m` imputations are drawn from the single midpoint-initialised
fit ("one-step" MI). When the inspection intervals are *narrow* relative to
the time scale this is unbiased, but with *wide* gaps (mean gap of the order
of a third of the follow-up) the initialiser's midpoint bias leaks into the
imputations: early-time survival is over-estimated, and -- because the pooled
standard errors remain honest -- coverage *deteriorates as $n$ grows*.

The remedy is `iter`: with `iter = k`, each imputation chain alternates
imputation and re-fitting on its own completed data $k$ times, so later
imputations come from fits whose dependence on the initialiser is
progressively attenuated (a chained-MI scheme in the spirit of poor man's
data augmentation [@weitanner1990] and the `maxit` iterations of `mice`
[@vanbuuren2011]).

```{r, eval = FALSE}
fit_sparse <- pamm_ic(
  Surv(L, R, type = "interval2") ~ trt,
  data          = icd,
  model_formula = ped_status ~ trt + s(tend, by = trt),
  cut           = seq(0, 10, by = 0.5),
  m             = 20,
  iter          = 3   # recommended for sparse panels
)
```

**Recommendation** (from the package's Monte Carlo benchmark, 500 replications
per cell): keep the default `iter = 1` for densely inspected data; use
`iter = 3` for sparse panels -- it removes most of the early-time survival bias
(`iter = 5` essentially all of it, with bias shrinking roughly geometrically)
at a fit cost roughly linear in `iter`. Two caveats from the same benchmark:

* the *hazard* near sharp peaks remains attenuated under sparse inspection no
  matter how many iterations are used -- with wide intervals that peak
  information is genuinely lost, not just imputed badly; and
* with flexible time-varying-effect terms and small samples, iterating can
  occasionally amplify a weakly identified chain into a degenerate fit with
  vacuous intervals. `pamm_ic()` flags such chains with a warning (and in
  `$unstable_chains`); if it fires, inspect the affected fits and consider a
  less flexible `model_formula` or fewer iterations.

## Competing risks

With $K$ competing causes, `pamm_ic_cr()` imputes the event time from a fitted
**cause-specific** PAMM while retaining observed event causes. (It accepts the
same `iter` argument as `pamm_ic()`, with the same sparse-panel
recommendation.) This is the usual
competing-risks interval-censoring setup: the cause is known at detection, but
the event time is only known to lie in $(L, R]$. If an event cause is explicitly
missing (`NA`), `pamm_ic_cr()` can also sample the cause from the fitted
cause-specific hazards at the imputed time. Each completed data set is then
re-fit as a cause-specific PAMM, and cumulative incidence functions are obtained
from those hazards with `add_cif()`. Note that this targets cause-specific
hazards (and CIFs derived from them), and so differs from @delord2016, who
develop MI for interval-censored competing risks in the **subdistribution-hazard**
(Fine--Gray) framework; the scheme here is the cause-specific analogue. Because
we impute from a cause-specific model and analyse with one, the imputation and
analysis models are congenial.

### How event times are imputed when cause is known

For a subject known to fail from cause $k$ in $(L, R]$, the correct conditional
density of the event time is the cause-$k$ *sub-density*, restricted to the
interval,
$$
p(t \mid L < T \le R,\ \text{cause}=k) \;\propto\; f_k(t) = h_k(t)\,S(t),
\qquad
S(t) = \exp\!\Big(-\textstyle\int_0^t h_\bullet(u)\,du\Big),
$$
where $h_\bullet = \sum_j h_j$ is the **all-cause** hazard. The all-cause hazard
is unavoidable here: to fail from cause $k$ *at* $t$, the subject must first
survive *all* causes up to $t$ (hence the all-cause survival $S$) and then fail
from cause $k$ (hence $h_k$). Drawing instead from a "cause-$k$-only" law
$h_k(t)\exp(-\int h_k)$ would be the classic net/latent-time error -- it ignores
that a competing cause may remove the subject before $t$, biasing imputed times
too late.

The all-cause *conditional* CDF inverts in closed form on the piecewise grid
(it is just the single-event sampler applied to $S = \exp(-H_\bullet)$), whereas
$f_k(t) = h_k(t)S(t)$ has no equally clean inverse. So `pamm_ic_cr()` uses the
all-cause conditional only as an easy-to-invert **proposal** and then **accepts**
the draw with probability $h_k(t)/h_\bullet(t) \le 1$. The accepted draws are
proportional to (proposal) $\times$ (acceptance) $= h_\bullet(t)S(t)\cdot
h_k(t)/h_\bullet(t) \propto h_k(t)S(t) = f_k(t)$ -- exactly the target; the
$h_\bullet$ cancels. The cause itself is not re-sampled in this known-cause
case. For an **unknown** event cause (`NA` in the cause column), the same
machinery is read the other way: the time is drawn from the all-cause
conditional and the cause is then sampled with probability
$h_k(t)/h_\bullet(t)$, the same `hazard / total_hazard` ratio used to build CIF
increments in `add_cif()`.

Two consequences are worth noting:

* If the cause-specific hazards are proportional ($h_k/h_\bullet$ constant over
  time) the acceptance is uniform and the all-cause conditional already *equals*
  the cause-$k$ conditional -- the rejection step is a no-op. It only matters
  when the cause mix shifts over follow-up (e.g. one cause dominates early,
  another late), which is exactly when cause-$k$ times need to be re-weighted.
* For a **rare** cause $h_k/h_\bullet$ is small everywhere, so acceptance is low;
  the sampler caps the number of tries and errors if it cannot accept a valid
  proposal. For strongly imbalanced causes a direct inversion of the cause-$k$
  sub-cumulative would be more robust -- a possible refinement.

### Example

To make the estimand visible, we simulate a small competing-risks data set with
two genuinely different causes: cause 1 dominates early and decreases over time,
whereas cause 2 rises later and is more common under treatment. The target
estimands are the cause-specific CIFs
$F_k(t \mid x) = P(T \le t, J = k \mid x)$.

There is no direct `survival::survfit()` analogue of Turnbull's estimator for
interval-censored competing-risk CIFs. Since this is simulated data, however, we
can use the latent exact event times to compute an **oracle** nonparametric
Aalen-Johansen estimate and check whether the interval-censored MI fit recovers
the same CIFs.

```{r, echo = FALSE}
sim_cr_piecewise <- function(data, cut, max_time = max(cut)) {
  n <- nrow(data)
  time <- rep(max_time, n)
  cause <- rep(0L, n)
  trt_num <- as.integer(as.character(data$trt))

  for (i in seq_len(n)) {
    for (j in seq_len(length(cut) - 1L)) {
      t0 <- cut[j]
      t1 <- cut[j + 1L]
      tm <- (t0 + t1) / 2
      h1 <- exp(-2.45 - 0.10 * tm + 0.05 * trt_num[i])
      h2 <- exp(-4.40 + 0.24 * tm + 0.75 * trt_num[i])
      htot <- h1 + h2
      p_event <- 1 - exp(-htot * (t1 - t0))

      if (runif(1) <= p_event) {
        time[i] <- t0 - log(1 - runif(1) * p_event) / htot
        cause[i] <- sample(c(1L, 2L), 1L, prob = c(h1, h2))
        break
      }
    }
  }

  data.frame(data, time = time, cause = cause, status = as.integer(cause > 0))
}

aj_cif_df <- function(fit) {
  sm <- summary(fit)
  states <- setdiff(colnames(sm$pstate), "(s0)")
  bind_rows(lapply(states, function(state) {
    data.frame(
      tend = sm$time,
      cif = sm$pstate[, state],
      cause = state,
      trt = sub("^trt=", "", as.character(sm$strata)),
      method = "Aalen-Johansen (true times)"
    )
  }))
}
```

```{r, fig.width = 8, fig.height = 3.8}
set.seed(4)
cr_cut <- seq(0, 10, by = 0.1)
cr_exact <- data.frame(trt = rbinom(700, 1, 0.5) |> as.factor()) |>
  sim_cr_piecewise(cut = cr_cut, max_time = 10)
icd_cr <- add_inspections(cr_exact, rate = 1.3, max_time = 10)
icd_cr$cause <- factor(icd_cr$cause)

fit_cr <- pamm_ic_cr(
  Surv(L, R, type = "interval2") ~ trt,
  cause = "cause",
  model_formula = ped_status ~ cause * trt + s(tend, by = cause, k = 6),
  cut = seq(0, 10, by = 0.5),
  data = icd_cr,
  m = 20
)
summary(fit_cr)

ped_cr <- as_ped(
  transform(icd_cr, time = pmin(true_time, 10), status = cause),
  Surv(time, status) ~ trt,
  cut = seq(0, 10, by = 0.5)
)
nd_cr <- make_newdata(
  ped_cr,
  tend = unique(tend),
  cause = levels(icd_cr$cause)[-1],
  trt = unique(trt)
) |>
  group_by(cause, trt)
cif_cr <- add_cif(nd_cr, fit_cr, nsim = 150)

aj_cr <- survfit(
  Surv(time, factor(cause, levels = c(0, 1, 2))) ~ trt,
  data = cr_exact
) |>
  aj_cif_df()
cif_lines <- bind_rows(
  transmute(cif_cr, tend, trt, cause, cif, method = "PAMM (MI)"),
  aj_cr
)
cif_lines$method <- factor(
  cif_lines$method,
  levels = c("PAMM (MI)", "Aalen-Johansen (true times)")
)

ggplot(cif_lines, aes(x = tend, y = cif, color = cause)) +
  geom_ribbon(data = cif_cr,
    aes(ymin = cif_lower, ymax = cif_upper, fill = cause),
    alpha = 0.18, color = NA) +
  geom_line(aes(linetype = method), linewidth = 0.8) +
  scale_color_viridis_d(end = 0.5, aesthetics = c("colour", "fill"), option = "C") +
  scale_linetype_manual(values = c("solid", "22")) +
  facet_grid(~ trt) +
  labs(x = "time", y = "CIF", color = "cause", fill = "cause", linetype = NULL,
    title = "Interval-censored competing-risk CIFs\nwith oracle Aalen-Johansen benchmark")
```

The dashed Aalen-Johansen curves use the simulated exact event times and are
therefore an oracle reference, not an estimator available in a real
interval-censored data set. They are useful here because they target the same
CIFs as `add_cif()`. The MI curves track the oracle well: cause 1 contributes
more early failures, while cause 2 accumulates later and more strongly in the
treated group. This example is deliberately small so the vignette remains quick
to render; for an analysis use a larger `m` and more posterior draws in
`add_cif()`.


## Notes and limitations

* The interval cut-points are fixed once and shared across all imputations; this
  is required for the imputations to be poolable.
* The imputation-model coefficients are drawn from their posterior before
  each imputation ("proper" MI); this is what makes the pooled intervals
  calibrated.
* Recurrent-event and general multi-state interval censoring are not yet
  supported: imputing one transition time changes the at-risk window for the
  next, so the imputations must be drawn jointly along each subject's path -- a
  planned extension.

## References
