---
title: "Multi-State Modelling"
author: "Johannes Piller, Andreas Bender"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-State Modelling}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, echo = FALSE, warning = FALSE}
library(knitr)
opts_chunk$set(
  fig.align  = "center",
  cache      = FALSE,
  message    = FALSE,
  fig.height = 5,
  fig.width  = 5,
  dpi        = 300
)
library(ggplot2)
theme_set(theme_bw())
```

## Introduction

In this vignette, we show examples of how to estimate and visualize transition probabilities for multi-state data. 
To validate that the PAMM baseline hazards are correctly specified, we compare the 
cumulative hazards transition probabilities, and their simulated  [confidence intervals](https://adibender.github.io/simpamm/confidence-intervals.html) estimated by `pammtools` against those from the Aalen-Johannsen estimator as implemented in `mstate`. 
Since we fit a baseline model stratified by treatment effect without additional covariates, both approaches should yield identical results up to 
numerical approximation from the piecewise-constant hazard assumption in PAMMs.

## Data transformation
The data transformation required to fit PAMMs to multi-state data extends the standard 
single-event PED transformation (see the [data transformation vignette](https://adibender.github.io/pammtools/articles/data-transformation.html) for details) to each transition type. The follow-up of each subject 
is split at all observed transition times across the entire dataset, and a row is added 
for every interval-transition combination the subject is at risk for. Two key 
differences arise compared to the single-event case: first, delayed entry into the risk 
set is handled automatically, since subjects are only at risk for transitions out of a 
state after they have entered it; second, competing events are treated as censoring for 
all other transitions within the same interval. The result is a stacked long-format 
dataset with one row per subject, interval, and transition, which can be passed 
directly to a Poisson regression model. In `pammtools`, this transformation is handled 
by `as_ped()` with the additional `transition` argument specifying the column that 
identifies the transition type.

# `prothr` Data - Time to abnormal prothrombin levels in liver cirrhosis

For illustration, we use the `prothr` data set from the **`mstate`** package. 
- 488 liver cirrhosis patients
- endpoints: `status` = 0/1 for censoring, event
- events of interest: `status` = 1 and `to` = 2/3 for abnormal prothrombin level (patient can transition between normal and abnormal prothrombin level back and forth), death, `tstart`, `tend` as time-to-event
- variable of interest  `treat`: A patient's treatment (Placebo, Prednisone)

```{r, echo = TRUE, warning = FALSE}
library(survival)
library(mgcv)
library(ggplot2)
library(pammtools)
library(purrr)
library(mstate)
library(checkmate)
library(dplyr)

data(prothr, package = "mstate")
prothr |> filter(id == 46) |> knitr::kable() # example patients
```
In general, one has to follow three steps to derive transition probabilities from multi-state survival data. 
- First, we need to transform the survival data `prothr` into piecewise exponential data. 
- Second, we need to estimate the log hazard structure using PAM objects. 
- Third, we need to post-process the data to include all relevant objects of interest in our data set. 

## Transformation to PED format

Transforming the survival data `prothr` into piecewise exponential data, we can use 
```{r ms-prothr-ped, echo = TRUE}
# not necessary, prothr already contains all possible transitions
# my.prothr <- prothr |> add_counterfactual_transitions() # add possible transitions

data("prothr", package = "mstate")
prothr <- prothr |> 
  mutate(transition = as.factor(paste0(from, "->", to))
         , treat = as.factor(treat)) |>
  rename(tstart = Tstart, tstop = Tstop) |>
  filter(tstart != tstop) |>
  select(-trans)

ped <- as_ped(
  data       = prothr,
  formula    = Surv(tstart, tstop, status)~ .,
  transition = "transition",
  id         = "id",
  timescale  = "calendar",
  tdc_specials="concurrent"
)
```
where `add_counterfactual_transitions` is a helper function, which adds all possible transitions at each point in time.
Since the `prothr` data involves an illness-death model where patients can transition 
back and forth between normal and abnormal prothrombin levels, hazards naturally depend 
on overall disease duration rather than time spent in the current state. We therefore 
use the calendar time scale (`timescale = "calendar"`), where time runs continuously 
from study entry and is not reset after each transition.

## Baseline Multi-State Model

Estimating the log hazard structure using PAM objects, we can use
```{r ms-prothr-pam, echo = TRUE}
pam <- bam(ped_status ~ s(tend, by=transition, k = 20) + transition * treat
           , data = ped
           , family = poisson()
           , offset = offset
           , method = "fREML"
           , discrete = TRUE)
summary(pam)
```


## Post-processing, i.e. plotting the transition probabilities
Transition probabilities are 
derived from the cumulative hazards estimated by the model. 
Specifically, for each time point $t$, the increment of the cumulative hazard
matrix $d\mathbf{H}(t)$ is computed from the transition-specific cumulative hazards.
The diagonal elements are set to minus the sum of all outgoing transition hazards,
ensuring rows sum to zero. Transition probabilities are then obtained as the successive
product of matrices $(\mathbf{I} + d\mathbf{H}(u))$ over all observed transition times
$u$ between $s$ and $t$:

$$\mathbf{P}(s, t) = \prod_{u \in [s,t)} \left( \mathbf{I} + d\mathbf{H}(u)\right)$$

This is the continuous-time analogue of the Aalen-Johannsen estimator, and reduces
to it when the cumulative hazards are estimated non-parametrically. In `pammtools`,
the cumulative hazards are instead estimated via Poisson regression, allowing
covariate effects to propagate directly into the transition probabilities. The
function `add_trans_prob()` implements this product-limit calculation internally,
so the post-processing step below is identical regardless of model complexity.

```{r ms-prothr-ndf, echo = TRUE, warning = FALSE}
ndf <- make_newdata(ped, 
        tend  = unique(tend)[unique(tend) < 3000], 
        treat  = unique(treat), 
        transition = unique(transition)
        ) 

ndf <- ndf  |>
  group_by(treat, transition) |>  # important!
  arrange(treat, transition, tend) |>
  add_trans_prob(pam, ci=TRUE)

ndf <- ndf  |>
  group_by(treat, transition) |>  # important!
  add_cumu_hazard(pam, overwrite = T)
```
`make_newdata` creates a data set containing all covariates and all their combinations from the PAM object. The convenience function `add_trans_prob` crates a new column `trans_prob`, which can be visualized using the `ggplot2` functionality.

```{r prothr-viz, echo = TRUE, fig.width = 8, fig.height = 4, out.width = "600px"}
# visualization
ggplot(ndf, aes(x=tend)) + 
  geom_line(aes(y=trans_prob, col=treat)) +
  geom_ribbon(aes(ymin = trans_lower, ymax = trans_upper, fill=treat), alpha = .3) +
  scale_color_manual(values = c("firebrick2"
                                , "steelblue")
                     )+
  scale_fill_manual(values = c("firebrick2"
                                , "steelblue")
                     )+
  facet_wrap(~transition) +
  scale_x_continuous(limits = c(0, 3000), oob = scales::oob_keep) +
  scale_y_continuous(limits = c(0, 1),    oob = scales::oob_keep) +
  labs(y = "Transition Probability", x = "time", color = "Treatment", fill= "Treatment")

```

## Comparison of the results with Aalen-Johannsen estimator

Comparing the ``pammtools`` results with the ``mstate`` results, we want to validate that the baselines are indeed correct. The following code shows the comparison between the ``mstate``.

First, we compare the cumulative hazards
```{r, echo = FALSE, warning = FALSE}
# pammtools
ndf <- ndf |>
  mutate(package = "pammtools")
```
```{r prothr-prep-mstate, eval=TRUE, echo = FALSE, warning = FALSE}
library(mstate)

# --- setup ------------------------------------------------------------------
data(prothr, package = "mstate")
tmat        <- attr(prothr, "trans")
treat_levels <- c("Placebo", "Prednisone")
trans_labels <- c("1->2", "1->3", "2->1", "2->3")
trans_recode <- c("1->2" = "0->1", "1->3" = "0->2", "2->1" = "1->0", "2->3" = "1->2")
alpha        <- 0.05

transitions_to_extract <- list(
  list(from = 1, to = 2, label = "1->2"),
  list(from = 1, to = 3, label = "1->3"),
  list(from = 2, to = 1, label = "2->1"),
  list(from = 2, to = 3, label = "2->3")
)

# --- helpers ----------------------------------------------------------------
fit_msfit <- function(treat_label) {
  pr <- subset(prothr, treat == treat_label)
  attr(pr, "trans") <- tmat
  msfit(coxph(Surv(Tstart, Tstop, status) ~ strata(trans), data = pr), trans = tmat)
}

extract_probtrans <- function(pt, treat_label) {
  z <- qnorm(1 - alpha / 2)
  map_dfr(transitions_to_extract, function(tr) {
    state_col <- paste0("pstate", tr$to)
    se_col    <- paste0("se",     tr$to)
    prob      <- pt[[tr$from]][[state_col]]
    se        <- pt[[tr$from]][[se_col]]
    data.frame(
      tend        = pt[[tr$from]]$time,
      trans_prob  = prob,
      trans_lower = prob - z * se,
      trans_upper = prob + z * se,
      transition  = tr$label
    )
  }) |>
    mutate(treat = treat_label, package = "mstate")
}

# --- fit --------------------------------------------------------------------
msf <- setNames(map(treat_levels, fit_msfit), treat_levels)
pt  <- setNames(map(msf, ~ probtrans(.x, predt = 0)), treat_levels)

# --- cumulative hazards -----------------------------------------------------
long_mstate <- map_dfr(treat_levels, ~ {
  msf[[.x]]$Haz |>
    mutate(transition = trans_labels[trans], treat = .x)
}) |>
  rename(tend = time, cumu_hazard = Haz) |>
  mutate(package = "mstate") |>
  select(tend, treat, transition, cumu_hazard, package)

long_haz_df <- ndf |>
  mutate(package = "pammtools") |>
  select(tend, treat, transition, cumu_hazard, package) |>
  bind_rows(long_mstate) |>
  mutate(transition = trans_recode[transition])

# --- transition probabilities -----------------------------------------------
pt_df_mstate <- map_dfr(
  treat_levels,
  ~ extract_probtrans(pt[[.x]], .x)
)

pt_df <- ndf |>
  mutate(package = "pammtools") |>
  select(tend, treat, transition, trans_prob, trans_lower, trans_upper, package) |>
  bind_rows(pt_df_mstate)
```


```{r, eval=TRUE, echo = TRUE, fig.width = 8, fig.height = 4, out.width = "600px"}
comparison_nelaal <- ggplot(long_haz_df, aes(x=tend, y=cumu_hazard, col=treat, linetype = package)) + 
  geom_line() +
  facet_wrap(~transition, ncol = 4, labeller = label_both) +
  scale_color_manual(values = c("firebrick2"
                                , "steelblue")
                     )+
  # scale_linetype_manual(values = c("solid", "dashed", "solid", "dashed")) +18:
  ylab("Cumulative Hazards") +
  xlab("time in days") +
  scale_y_continuous(limits = c(0, 6), oob = scales::oob_keep) +
  scale_linetype_manual(values=c("dotted", "solid")) +
  theme_bw()

comparison_nelaal
```

Second, we compare transition probability estimates. 

```{r, eval=TRUE, echo = FALSE, fig.width = 8, fig.height = 4, out.width = "600px"}
comparison_aaljoh <- ggplot(pt_df, aes(x=tend, y=trans_prob, col=treat, linetype = package)) + 
  geom_line() +
  # geom_ribbon(aes(ymin = trans_lower, ymax = trans_upper, fill=treat), alpha = .3) +
  facet_wrap(~transition, ncol = 4, labeller = label_both) +
  scale_color_manual(values = c("firebrick2"
                                , "steelblue")
                     )+
  # scale_linetype_manual(values = c("solid", "dashed", "solid", "dashed")) +18:
  ylab("Transition Probabilities") +
  xlab("time in days") +
  scale_y_continuous(limits = c(0, 1),    oob = scales::oob_keep) +
  scale_x_continuous(limits = c(0, 3000), oob = scales::oob_keep) +
  scale_linetype_manual(values=c("dotted", "solid")) +
  theme_bw()

comparison_aaljoh
```

Third, we compare the width of the confidence bands
```{r, eval=TRUE, echo = FALSE, fig.width = 8, fig.height = 4, out.width = "600px"}
comparison_aaljoh_ci <- ggplot(pt_df, aes(x=tend, y=trans_prob, col=interaction(package, treat), linetype = package)) + 
  geom_line(linetype = "solid") +
  geom_line(aes(y=trans_lower), linetype = "dotted") +
  geom_line(aes(y=trans_upper), linetype = "dotted") +
  geom_ribbon(data = subset(pt_df, package == "mstate"), aes(ymin = trans_lower, ymax = trans_upper), fill = "grey", alpha = 0.3) +
  facet_wrap(~treat+transition, ncol = 4, labeller = label_both) +
  ylab("Transition Probabilities") +
  xlab("time in days") +
  scale_y_continuous(limits = c(0, 1),    oob = scales::oob_keep) +
  scale_x_continuous(limits = c(0, 3000), oob = scales::oob_keep) +
  scale_color_manual(values = c("mstate.Placebo" = "grey"
                              ,"mstate.Prednisone" = "grey"
                              , "pammtools.Placebo" = "firebrick2"
                              , "pammtools.Prednisone" = "steelblue")
                   )+
  theme_bw()+
  guides(color = "none", fill = "none", linetype = "none")

comparison_aaljoh_ci
```

<!-- Zaehlmarke VGWort -->
<img src="https://vg09.met.vgwort.de/na/f993b9c06b5249adb509a4df30d807a6" width="1" height="1" alt="">
