---
title: "Multi-State Modelling"
author: "Johannes Piller"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-State Modelling}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, echo = FALSE, warning = FALSE}
library(knitr)
opts_chunk$set(
  fig.align  = "center",
  cache      = FALSE,
  message    = FALSE,
  fig.height = 5,
  fig.width  = 5,
  dpi        = 300
)

```

## Introduction

In this vignette, we show examples of how to estimate and visualize transition probabilities for multi-state data. 
To validate that the PAMM baseline hazards are correctly specified, we compare the 
cumulative hazards transition probabilities, and their simulated  [confidence intervals](https://adibender.github.io/simpamm/confidence-intervals.html) estimated by `pammtools` against those from the Aalen-Johannsen estimator as implemented in `mstate`. 
Since we fit a baseline model stratified by treatment effect without additional covariates, both approaches should yield identical results up to 
numerical approximation from the piecewise-constant hazard assumption in PAMMs.

## Example multi-state analysis: Time to abnormal prothrombin levels in liver cirrhosis using the `prothr` data

For illustration, we use the `prothr` data set from the **`mstate`** package. 
- 488 liver cirrhosis patients
- endpoints: `status` = 0/1 for censoring, event
- events of interest: `status` = 1 and `to` = 2/3 for abnormal prothrombin level (patient can transition between normal and abnormal prothrombin level back and forth), death, `tstart`, `tend` as time-to-event
- variable of interest  `treat`: A patient's treatment (Placebo, Prednisone)

<details>

<summary>
  R-Code Setup
</summary>
```{r, echo = TRUE, warning = FALSE}
library(survival)
library(mgcv)
library(ggplot2)
library(pammtools)
library(purrr)
library(mstate)
library(checkmate)
library(dplyr)

theme_set(theme_bw())

data(prothr, package = "mstate")
```
</details>

```{r, eval = TRUE, echo = FALSE, warning = FALSE}
prothr |> filter(id == 46) |> knitr::kable() # example patients
```

In general, one has to follow three steps to derive transition probabilities from multi-state survival data. 
- First, we need to transform the survival data `prothr` into piecewise exponential data. 
- Second, we need to estimate the log hazard structure using PAM objects. 
- Third, we need to post-process the data to include all relevant objects of interest in our data set. 

## Data transformation: From raw `prothr` data to piecewise exponential data

The data transformation required to fit PAMMs to multi-state data extends the standard 
single-event piecewise exponential data (PED) transformation (see the [data transformation vignette](https://adibender.github.io/pammtools/articles/data-transformation.html) for details) to each transition type. The follow-up of each subject 
is split at all observed transition times across the entire dataset, and a row is added 
for every interval-transition combination the subject is at risk for. Two key 
differences arise compared to the single-event case: first, delayed entry into the risk 
set is handled automatically, since subjects are only at risk for transitions out of a 
state after they have entered it; second, competing events are treated as censoring for 
all other transitions within the same interval. The result is a stacked long-format 
dataset with one row per subject, interval, and transition, which can be passed 
directly to a Poisson regression model. In `pammtools`, this transformation is handled 
by `as_ped()` with the additional `transition` argument specifying the column that 
identifies the transition type.

```{r ms-prothr-ped, echo = TRUE}
data("prothr", package = "mstate")
prothr <- prothr |> 
  mutate(transition = as.factor(paste0(from, "->", to))
         , treat = as.factor(treat)) |>
  rename(tstart = Tstart, tstop = Tstop) |>
  filter(tstart != tstop) |>
  select(-trans)

ped <- as_ped(
  data       = prothr,
  formula    = Surv(tstart, tstop, status)~ .,
  transition = "transition",
  id         = "id",
  timescale  = "calendar",
  tdc_specials="concurrent"
)

```

<details>

<summary>
  Transformed PED for `id == 46`
</summary>
```{r, echo = TRUE, warning = FALSE}
ped |> filter(id == 46) |> knitr::kable()
```
</details>

Since the `prothr` data involves an illness-death model where patients can transition 
back and forth between normal and abnormal prothrombin levels, hazards naturally depend 
on overall disease duration rather than time spent in the current state. We therefore 
use the calendar time scale (`timescale = "calendar"`), where time runs continuously 
from study entry and is not reset after each transition.

## Model estimation: Fitting a baseline multi-state model

Estimating the log hazard structure using PAM objects, we can use
```{r ms-prothr-pam, echo = TRUE}
pam <- bam(ped_status ~ s(tend, by=transition, k = 30) + transition * treat
           , data = ped
           , family = poisson()
           , offset = offset
           , method = "fREML"
           , discrete = TRUE)
summary(pam)
```
As with single-event survival data, transforming to PED format recasts the survival 
problem as a Poisson regression, which is why we specify `family = poisson()` with 
`offset = offset` (the offset accounts for the varying interval lengths introduced 
by the PED transformation). 

Choosing `k = 30` sets the number of basis functions for the smooth baseline hazard 
`s(tend, by = transition)`, allowing sufficient flexibility to capture complex 
transition-specific hazard shapes. The actual smoothness is controlled by the 
penalty term estimated via `fREML`, so the choice of `k` only sets an upper bound 
on flexibility.

The two computational arguments `method = "fREML"` and `discrete = TRUE` are chosen 
for efficiency. `fREML` (fast REML) is the recommended smoothing parameter estimation 
method for `bam()` on large datasets, as it is considerably faster than standard REML 
while producing nearly identical estimates. `discrete = TRUE` enables discretization 
of covariates before fitting, which further reduces computation time and memory usage 
by exploiting the fact that `tend` takes only a limited number of unique values by 
construction of the PED transformation.

## Post-processing: Calculating and visualizing transition probabilities
Transition probabilities are derived from the cumulative hazards estimated by the model. 
Specifically, for each time point $t$, the increment of the cumulative hazard
matrix $d\mathbf{H}(t)$ is computed from the transition-specific cumulative hazards.
The diagonal elements are set to minus the sum of all outgoing transition hazards,
ensuring rows sum to zero. Transition probabilities are then obtained as the successive
product of matrices $(\mathbf{I} + d\mathbf{H}(u))$ over all observed transition times
$u$ between $s$ and $t$:

$$\mathbf{P}(s, t) = \prod_{u \in [s,t)} \left( \mathbf{I} + d\mathbf{H}(u)\right)$$

This is the continuous-time analogue of the Aalen-Johannsen estimator, and reduces
to it when the cumulative hazards are estimated non-parametrically. 
In `pammtools`, the cumulative hazards are instead estimated via Poisson regression, allowing
covariate effects to propagate directly into the transition probabilities. The
function `add_trans_prob()` implements this product-limit calculation internally,
so the post-processing step below is identical regardless of model complexity.

In general, note that transition probabilities summarize *net movement* and don't distinguish paths.
This is especially relevant when interpreting transition probabilities for 
multi-state models with back-transitions (as in the `prothr` example).

```{r ms-prothr-ndf, echo = TRUE, warning = FALSE}
ndf <- make_newdata(ped, 
        tend  = unique(tend)[unique(tend) < 3000], 
        treat  = unique(treat), 
        transition = unique(transition)
        ) 

ndf <- ndf  |>
  group_by(treat, transition) |>  # important!
  arrange(treat, transition, tend) |>
  add_trans_prob(pam, ci=TRUE)
```
`make_newdata` creates a data set containing all covariates and all their combinations from the `PAM` object, i.e. for each event time up to 3000 days include rows for both treatment groups (`Prednisone`, `Placebo`) and each possible transition between the states (`normal prothrombin (1)`, `abnormal prothrombin (2)`, and `death (3)`). The convenience function `add_trans_prob` crates a new column `trans_prob` (with confidence interval bounds `trans_lower` and `trans_upper`), which can be visualized using `ggplot2`.

```{r prothr-viz, echo = TRUE, fig.width = 8, fig.height = 4, out.width = "600px"}
# visualization
ggplot(ndf, aes(x=tend)) + 
  geom_line(aes(y=trans_prob, col=treat)) +
  geom_ribbon(aes(ymin = trans_lower, ymax = trans_upper, fill=treat), alpha = .3) +
  scale_color_manual(values = c("firebrick2"
                                , "steelblue")
                     )+
  scale_fill_manual(values = c("firebrick2"
                                , "steelblue")
                     )+
  facet_wrap(~transition) +
  scale_x_continuous(limits = c(0, 3000), oob = scales::oob_keep) +
  scale_y_continuous(limits = c(0, 1),    oob = scales::oob_keep) +
  labs(y = "Transition Probability", x = "time", color = "Treatment", fill= "Treatment")

```

## Model Comparisons: Comparison of ``pammtools`` results with the ``mstate`` Aalen-Johannsen estimator
To validate that the PAMM baseline hazards are correctly specified, we compare
cumulative hazards and transition probabilities estimated by `pammtools` against
the non-parametric Aalen-Johannsen estimator from `mstate`. Since we fit a
baseline model without covariates, both approaches should yield identical results
up to numerical approximation from the piecewise-constant hazard assumption in PAMMs.

First, we compare the cumulative hazards, which form the basis for all downstream
quantities, i.e. here the transition probabilities.
```{r, echo = FALSE, warning = FALSE}
# pammtools
ndf <- ndf  |>
  group_by(treat, transition) |>  # important!
  add_cumu_hazard(pam, overwrite = TRUE) |>
  mutate(package = "pammtools")
```

<details>

<summary>
  R-Code for comparing ``mstate`` models
</summary>

```{r prothr-prep-mstate, eval=TRUE, echo = TRUE, warning = FALSE}
# --- setup ------------------------------------------------------------------
data(prothr, package = "mstate")
tmat        <- attr(prothr, "trans")
treat_levels <- c("Placebo", "Prednisone")
trans_labels <- c("1->2", "1->3", "2->1", "2->3")
trans_recode <- c("1->2" = "0->1", "1->3" = "0->2", "2->1" = "1->0", "2->3" = "1->2")
alpha        <- 0.05

transitions_to_extract <- list(
  list(from = 1, to = 2, label = "1->2"),
  list(from = 1, to = 3, label = "1->3"),
  list(from = 2, to = 1, label = "2->1"),
  list(from = 2, to = 3, label = "2->3")
)

# --- helpers ----------------------------------------------------------------
fit_msfit <- function(treat_label) {
  pr <- subset(prothr, treat == treat_label)
  attr(pr, "trans") <- tmat
  msfit(coxph(Surv(Tstart, Tstop, status) ~ strata(trans), data = pr), trans = tmat)
}

extract_probtrans <- function(pt, treat_label) {
  z <- qnorm(1 - alpha / 2)
  map_dfr(transitions_to_extract, function(tr) {
    state_col <- paste0("pstate", tr$to)
    se_col    <- paste0("se",     tr$to)
    prob      <- pt[[tr$from]][[state_col]]
    se        <- pt[[tr$from]][[se_col]]
    data.frame(
      tend        = pt[[tr$from]]$time,
      trans_prob  = prob,
      trans_lower = prob - z * se,
      trans_upper = prob + z * se,
      transition  = tr$label
    )
  }) |>
    mutate(treat = treat_label, package = "mstate")
}

# --- fit --------------------------------------------------------------------
msf <- setNames(map(treat_levels, fit_msfit), treat_levels)
pt  <- setNames(map(msf, ~ probtrans(.x, predt = 0)), treat_levels)

# --- cumulative hazards -----------------------------------------------------
long_mstate <- map_dfr(treat_levels, ~ {
  msf[[.x]]$Haz |>
    mutate(transition = trans_labels[trans], treat = .x)
}) |>
  rename(tend = time, cumu_hazard = Haz) |>
  mutate(package = "mstate") |>
  select(tend, treat, transition, cumu_hazard, package)

long_haz_df <- ndf |>
  mutate(package = "pammtools") |>
  select(tend, treat, transition, cumu_hazard, package) |>
  bind_rows(long_mstate) |>
  filter(tend < 3000) |>
  mutate(transition = trans_recode[transition])

# --- transition probabilities -----------------------------------------------
pt_df_mstate <- map_dfr(
  treat_levels,
  ~ extract_probtrans(pt[[.x]], .x)
)

pt_df <- ndf |>
  mutate(package = "pammtools") |>
  select(tend, treat, transition, trans_prob, trans_lower, trans_upper, package) |>
  bind_rows(pt_df_mstate) |>
  filter(tend < 3000)

# --- plotting ---------------------------------------------------------------
p_comparison_cumu_haz <- ggplot(
  long_haz_df, 
  aes(x=tend, y=cumu_hazard, 
      col=treat, 
      linetype = package)
  ) + 
  geom_line() +
  facet_wrap(~transition, ncol = 4, labeller = label_both) +
  scale_color_manual(values = c("firebrick2"
                                , "steelblue")
                     ) +
  ylab("Cumulative Hazards") +
  xlab("time in days") +
  scale_y_continuous(limits = c(0, 6), oob = scales::oob_keep) +
  scale_linetype_manual(values=c("dotted", "solid")) +
  theme_bw()

p_comparison_trans_prob <- ggplot(
  pt_df,
  aes(x = tend, y = trans_prob,
      col      = interaction(package, treat),
      fill     = interaction(package, treat))
) +
  # point estimates
  geom_line(linewidth = 0.7) +
  # pammtools CIs as dotted lines
  geom_line(
    data = subset(pt_df, package == "pammtools"),
    aes(y = trans_lower), linetype = "dotted", linewidth = 0.4
  ) +
  geom_line(
    data = subset(pt_df, package == "pammtools"),
    aes(y = trans_upper), linetype = "dotted", linewidth = 0.4
  ) +
  # mstate CIs as shaded ribbon
  geom_ribbon(
    data = subset(pt_df, package == "mstate"),
    aes(ymin = trans_lower, ymax = trans_upper),
    alpha = 0.2, color = NA
  ) +
  facet_grid(rows = vars(treat), cols = vars(transition), labeller = label_both) +
  scale_color_manual(
    values = c(
      "mstate.Placebo"       = "grey50",
      "mstate.Prednisone"    = "grey50",
      "pammtools.Placebo"    = "firebrick2",
      "pammtools.Prednisone" = "steelblue"
    )
  ) +
  scale_fill_manual(
    values = c(
      "mstate.Placebo"       = "grey70",
      "mstate.Prednisone"    = "grey70",
      "pammtools.Placebo"    = "firebrick2",
      "pammtools.Prednisone" = "steelblue"
    )
  ) +
  scale_y_continuous(limits = c(0, 1), oob = scales::oob_keep) +
  labs(y = "Transition Probability", x = "Time in days") +
  theme_bw() +
  guides(color = "none", fill = "none")
```
</details>


```{r, eval=TRUE, echo = FALSE, fig.width = 8, fig.height = 4, out.width = "100%", fig.alt = "Line plot comparing pammtools and mstate cumulative hazards over time for four transitions, stratified by treatment group."}
p_comparison_cumu_haz
```

Second, we compare transition probability point estimates and confidence bands. 
Confidence bands from `mstate` (shaded ribbon) and `pammtools`
(dotted lines) are overlaid to allow a direct assessment of both location and
uncertainty.

```{r, eval=TRUE, echo = FALSE, fig.width = 8, fig.height = 4, out.width = "100%", fig.alt = "Line plot with confidence intervals comparing pammtools and mstate transition probabilities over time by treatment group and transition."}
p_comparison_trans_prob
```

<!-- Zaehlmarke VGWort -->
<img src="https://vg09.met.vgwort.de/na/f993b9c06b5249adb509a4df30d807a6" width="1" height="1" alt="">
