---
title: "Multi-State"
author: "Johannes Piller, Andreas Bender"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{multi-state}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  fig.align  = "center",
  cache      = TRUE,
  message    = FALSE,
  fig.height = 5,
  fig.width  = 5,
  dpi        = 300
)
library(ggplot2)
theme_set(theme_bw())
```

## Introduction

In this vignette, we show examples of how to estimate and visualize transition probabilities for multi-state data. First, we show that a baseline PAMMs model returns equal probabilities as the empirical transition probabilities stemming from the Aalen-Johannsen estimator. Second, we show that the estimated baseline model in a multi-state setting with backtransitions using PAMMs is equal to a Cox proportional hazard model. Third, we illustrate the application of PAMMs in the analysis of the non-linear effect of hemoglobin on the probabilities to transition to a plasma cell malignancy (PCM) and / or death.

## Data transformation
The data transformation required to fit PAMMs to multi-state data is similar to the transformation in the single event case (see the [data transformation vignette](https://adibender.github.io/pammtools/articles/data-transformation.html) for details). In fact, internally the standard transformation is applied to each event type using `as_ped`, however, some choices have to be made

- left-truncation becomes relevant, because, so far, settings considered only transitions from the initial status to a single event of interest 
- specify the start and end of an event
- return the data as a list (one element for each transition) or a stacked data set (with an additional column (covariate), indicating the transition)

# `prothr` Data - Time to abnormal prothrombin levels in liver cirrhosis

For illustration, we use the `prothr` data set from the **`mstate`** package. 
- 488 liver cirrhosis patients
- endpoints: `status` = 0/1 for censoring, event
- events of interest: `status` = 1 and `to` = 2/3 for abnormal prothrombin level (patient can transition between normal and abnormal prothrombin level back and forth), death, `tstart`, `tend` as time-to-event
- variable of interest  `treat`: A patient's treatment (Placebo, Prednisone)

```{r, echo = TRUE}
library(survival)
library(pammtools)
library(purrr)

data(prothr, package = "mstate")
prothr |> filter(id == 46) |> knitr::kable() # example patients
```

```{r, echo = FALSE}
data("prothr", package = "mstate")
prothr <- prothr %>%
  rename(tstart = Tstart, tstop = Tstop) %>%
  filter(tstart != tstop)
```

In general, one has to follow three steps to derive transition probabilities from multi-state survival data. 
- First, we need to transform the survival data `prothr` into piecewise exponential data. 
- Second, we need to estimate the log hazard structure using PAM objects. 
- Third, we need to post-process the data to include all relevant objects of interest in our data set. 

## Transformation to PED format

Transforming the survival data `prothr` into piecewise exponential data, we can use 
```{r lib-ms-pammtools, echo = FALSE}
# load function as it is not yet included in the package
# devtools::load_all("C:/Users/ra63liw/Documents/98_git/pammtools-multi-state/pammtools")
# source("C:/Users/ra63liw/Documents/98_git/pammtools-multi-state/pammtools/R/add-functions.R")
# source("C:/Users/ra63liw/Documents/98_git/pammtools-multi-state/pammtools/tmp/add_transition_probabilities.R")
library(checkmate)
library(tidyverse)
```
```{r, echo = TRUE, dependson=c("lib-ms-pammtools")}
my.prothr <- prothr |> filter(status == 1) |> add_counterfactual_transitions() # add possible transitions
ped <- as_ped_multistate(
  data       = my.prothr,
  formula    = Surv(tstart, tstop, status)~ .,
  transition = "transition",
  id         = "id",
  timescale  = "calendar",
  tdc_specials="concurrent"
)
```
where `add_counterfactual_transitions` is a helper function, which adds all possible transitions at each point in time.

## Baseline Multi-State Model

Estimating the log hazard structure using PAM objects, we can use
```{r echo = TRUE}
pam <- pamm(ped_status ~ s(tend, by=transition) + transition * treat, data = ped)
summary(pam)
```


## Post-processing, i.e. plotting the transition probabilities

Post-processing the data to include all relevant objects of interest in our data set, we can use 
```{r, echo = TRUE, warning = FALSE, dependson=c("lib-ms-pammtools")}
ndf <- make_newdata(ped, tend  = unique(tend), treat  = unique(treat), transition = unique(transition)) 
ndf <- ndf  |>
  group_by(treat, transition) |>  # important!
  # arrange(treat, transition, tend) |>
  # add_trans_ci(pam) |>
  add_trans_prob(pam)
```
where `make_newdata` creates a data set containing all covariates and all their combinations from the PAM object. The convenience function `add_trans_prob` crates a new column `trans_prob`, which can be visualized.

```{r prothr-viz, echo = TRUE, fig.width = 8, fig.height = 4, out.width = "600px"}
# visualization
ggplot(ndf, aes(x=tend)) + 
  geom_line(aes(y=trans_prob, col=treat)) +
  # geom_ribbon(aes(ymin = trans_lower, ymax = trans_upper, fill=treat), alpha = .3) +
  facet_wrap(~transition) +
  labs(y = "Transition Probability", x = "time", color = "Treatment", fill= "Treatment")

```

## Comparison of the results with Aalen-Johannsen estimator
```{r, echo = FALSE, warning = FALSE, dependson=c("lib-ms-pammtools")}
library(mstate)
library(msm)
library(mvna)

# code from mstate documentation
data(prothr)
tmat <- attr(prothr, "trans")
pr0 <- subset(prothr, treat=="Placebo")
attr(pr0, "trans") <- tmat
pr1 <- subset(prothr, treat=="Prednisone")
attr(pr1, "trans") <- tmat
c0 <- coxph(Surv(Tstart, Tstop, status) ~ strata(trans), data=pr0)
c1 <- coxph(Surv(Tstart, Tstop, status) ~ strata(trans), data=pr1)


msf0 <- msfit(c0, trans=tmat)
msf1 <- msfit(c1, trans=tmat)

# Comparison as in Figure 2 of Titman (2015)
# Aalen-Johansen
pt0 <- probtrans(msf0, predt=0)[[2]] # changed predt from 1000 to 0
pt1 <- probtrans(msf1, predt=0)[[2]] # changed predt from 1000 to 0
par(mfrow=c(1,2))
plot(pt0$time, pt0$pstate1, type="s", lwd=2, xlim=c(0,4000), ylim=c(0,1),
     xlab="Time since randomisation (days)", ylab="Probability")
lines(pt1$time, pt1$pstate1, type="s", lwd=2, lty=3)
legend("topright", c("Placebo", "Prednisone"), lwd=2, lty=1:2, bty="n")
title(main="Aalen-Johansen")
# Landmark Aalen-Johansen
LMpt0 <- LMAJ(msdata=pr0, s=0, from=2) # changed predt from 1000 to 0
LMpt1 <- LMAJ(msdata=pr1, s=0, from=2) # changed predt from 1000 to 0
plot(LMpt0$time, LMpt0$pstate1, type="s", lwd=2, xlim=c(0,4000), ylim=c(0,1),
     xlab="Time since randomisation (days)", ylab="Probability")
lines(LMpt1$time, LMpt1$pstate1, type="s", lwd=2, lty=3)
legend("topright", c("Placebo", "Prednisone"), lwd=2, lty=1:2, bty="n")
title(main="Landmark Aalen-Johansen")
par(mfrow=c(1,1))

# plot hazards
mstate_dat0 <- msf0$Haz %>% mutate(transition = case_when(
  trans == 1 ~ "1->2",
  trans == 2 ~ "1->3",
  trans == 3 ~ "2->1",
  trans == 4 ~ "2->3",
  .default = "cens"
  )
  , treat = "Placebo")
mstate_dat1 <- msf1$Haz %>% mutate(transition = case_when(
  trans == 1 ~ "1->2",
  trans == 2 ~ "1->3",
  trans == 3 ~ "2->1",
  trans == 4 ~ "2->3",
  .default = "cens"
  )
, treat = "Prednisone")

mstate_dat <- rbind(mstate_dat0, mstate_dat1)


ggplot(data = mstate_dat, aes(x=time, y=Haz)) + 
  geom_line(aes(col=treat)) +
  facet_wrap(~transition, ncol = 4, labeller = label_both)

# data analysis / preparation
head(prothr)
table(prothr$from, prothr$to)
# transitions 1->2, 1->3, 2->1, 2->3 possible.
# classical multi-state setup with recurrent events. 

# try analysis with etm
# transform days in fractions of year
data <- prothr %>% mutate(Tstart = Tstart / 365.25, 
                          Tstop = Tstop /365.25,
                          from = from - 1,
                          to = to - 1)
head(data)
dim(data)


# build transition matrix for mvna
tra <- matrix(FALSE, ncol = 3, nrow = 3)
dimnames(tra) <- list(c("0", "1", "2"), c("0", "1", "2"))
tra[1, 2:3] <- TRUE
tra[2, c(1,3)] <- TRUE

# print transition matrix
tra

my.data <- prothr %>% 
  mutate(from = from -1, to = to-1) %>%
  rename(entry = Tstart, exit = Tstop) %>% 
  arrange(id, entry, exit) %>% 
  filter(!(entry == exit)) %>%
  distinct()
dim(my.data)
head(my.data)

my.data %>% filter(tstart== tstop)

table(my.data$from, my.data$to)

# calculate cause specific hazards
my.nelaal <- mvna(my.data, c("0", "1", "2"), tra, "cens")
if (require(lattice)){
  xyplot(my.nelaal
         , strip=strip.custom(bg="white")
         , ylab="Cumulative Hazard"
         , lwd=2
         # , xlim=c(0,5)
         # , ylim=c(0,10)
         )
}
# differentiate between placebo and not
my.data.placebo <- my.data %>% filter(treat=="Placebo")
table(my.data.placebo$from, my.data.placebo$to)

my.nelaal <- mvna(my.data.placebo, c("0", "1", "2"), tra, "cens")
if (require(lattice)){
  xyplot(my.nelaal
         , strip=strip.custom(bg="white")
         , ylab="Cumulative Hazard"
         , lwd=2
         # , xlim=c(0,5)
         # , ylim=c(0,10)
  )
}


# estimate transition probabilites
etm.prothr <- etm(my.data, c("0", "1", "2"), tra, "cens", s = 0)

par(mfrow=c(2,2))
plot(etm.prothr, tr.choice = "0 1", conf.int = TRUE,
     lwd = 2, legend = FALSE, ylim = c(0, 1),
     xlim = c(0, 4000), xlab = "Days",
     ci.fun = "cloglog")

plot(etm.prothr, tr.choice = "0 2", conf.int = TRUE,
     lwd = 2, legend = FALSE, ylim = c(0, 1),
     xlim = c(0, 4000), xlab = "Days",
     ci.fun = "cloglog")

plot(etm.prothr, tr.choice = "1 0", conf.int = TRUE,
     lwd = 2, legend = FALSE, ylim = c(0, 1),
     xlim = c(0, 4000), xlab = "Days",
     ci.fun = "cloglog")

plot(etm.prothr, tr.choice = "1 2", conf.int = TRUE,
     lwd = 2, legend = FALSE, ylim = c(0, 1),
     xlim = c(0, 4000), xlab = "Days",
     ci.fun = "cloglog")
par(mfrow=c(1,1))
```
