---
title: "Multi-State Modelling"
author: "Johannes Piller, Andreas Bender"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{multi-state}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  fig.align  = "center",
  cache      = TRUE,
  message    = FALSE,
  fig.height = 5,
  fig.width  = 5,
  dpi        = 300
)
library(ggplot2)
theme_set(theme_bw())
```

## Introduction

In this vignette, we show examples of how to estimate and visualize transition probabilities for multi-state data. First, we show that a baseline PAMMs model returns equal probabilities as the empirical transition probabilities stemming from the Aalen-Johannsen estimator. Second, we show that the estimated baseline model in a multi-state setting with backtransitions using PAMMs is equal to a Cox proportional hazard model. Third, we illustrate the application of PAMMs in the analysis of the non-linear effect of hemoglobin on the probabilities to transition to a plasma cell malignancy (PCM) and / or death.

## Data transformation
The data transformation required to fit PAMMs to multi-state data is similar to the transformation in the single event case (see the [data transformation vignette](https://adibender.github.io/pammtools/articles/data-transformation.html) for details). In fact, internally the standard transformation is applied to each event type using `as_ped`, however, some choices have to be made

- left-truncation becomes relevant, because, so far, settings considered only transitions from the initial status to a single event of interest 
- specify the start and end of an event
- return the data as a list (one element for each transition) or a stacked data set (with an additional column (covariate), indicating the transition)

# `prothr` Data - Time to abnormal prothrombin levels in liver cirrhosis

For illustration, we use the `prothr` data set from the **`mstate`** package. 
- 488 liver cirrhosis patients
- endpoints: `status` = 0/1 for censoring, event
- events of interest: `status` = 1 and `to` = 2/3 for abnormal prothrombin level (patient can transition between normal and abnormal prothrombin level back and forth), death, `tstart`, `tend` as time-to-event
- variable of interest  `treat`: A patient's treatment (Placebo, Prednisone)

```{r, echo = TRUE}
library(survival)
library(mgcv)
library(ggplot2)
library(pammtools)
library(purrr)
library(mstate)
library(checkmate)
library(dplyr)

data(prothr, package = "mstate")
prothr |> filter(id == 46) |> knitr::kable() # example patients
```
In general, one has to follow three steps to derive transition probabilities from multi-state survival data. 
- First, we need to transform the survival data `prothr` into piecewise exponential data. 
- Second, we need to estimate the log hazard structure using PAM objects. 
- Third, we need to post-process the data to include all relevant objects of interest in our data set. 

## Transformation to PED format

Transforming the survival data `prothr` into piecewise exponential data, we can use 
```{r lib-ms-pammtools, echo = FALSE}
# load function as it is not yet included in the package
# devtools::load_all("C:/Users/ra63liw/Documents/98_git/pammtools-multi-state/pammtools")
# devtools::install("C:/Users/ra63liw/Documents/98_git/pammtools-multi-state/pammtools")
# source("C:/Users/ra63liw/Documents/98_git/pammtools-multi-state/pammtools/R/add-functions.R")
# source("C:/Users/ra63liw/Documents/98_git/pammtools-multi-state/pammtools/tmp/add_transition_probabilities.R")
library(checkmate)
```
```{r ms-prothr-ped, echo = TRUE, dependson=c("lib-ms-pammtools")}
# not necessary, prothr already contains all possible transitions
# my.prothr <- prothr |> add_counterfactual_transitions() # add possible transitions

data("prothr", package = "mstate")
prothr <- prothr |> 
  mutate(transition = as.factor(paste0(from, "->", to))
         , treat = as.factor(treat)) |>
  rename(tstart = Tstart, tstop = Tstop) |>
  filter(tstart != tstop) |>
  select(-trans)

ped <- as_ped(
  data       = prothr,
  formula    = Surv(tstart, tstop, status)~ .,
  transition = "transition",
  id         = "id",
  timescale  = "calendar",
  tdc_specials="concurrent"
)
```
where `add_counterfactual_transitions` is a helper function, which adds all possible transitions at each point in time.

## Baseline Multi-State Model

Estimating the log hazard structure using PAM objects, we can use
```{r ms-prothr-pam, echo = TRUE, dependson=c("ms-prothr-ped")}
# pam <- pamm(ped_status ~ s(tend, by=transition) + transition * treat, data = ped)
pam <- bam(ped_status ~ s(tend, by=transition) + transition * treat
           , data = ped
           , family = poisson()
           , offset = offset
           , method = "fREML"
           , discrete = TRUE)
summary(pam)
```


## Post-processing, i.e. plotting the transition probabilities

Post-processing the data to include all relevant objects of interest in our data set, we can use 
```{r ms-prothr-ndf, echo = TRUE, warning = FALSE, dependson=c("lib-ms-pammtools", "ms-prothr-pam")}
ndf <- make_newdata(ped, tend  = unique(tend), treat  = unique(treat), transition = unique(transition)) 
ndf <- ndf  |>
  group_by(treat, transition) |>  # important!
  arrange(treat, transition, tend) |>
  add_trans_prob(pam, ci=TRUE)

ndf <- ndf  |>
  group_by(treat, transition) |>  # important!
  add_cumu_hazard(pam, overwrite = T)
```
where `make_newdata` creates a data set containing all covariates and all their combinations from the PAM object. The convenience function `add_trans_prob` crates a new column `trans_prob`, which can be visualized.

```{r prothr-viz, echo = TRUE, fig.width = 8, fig.height = 4, out.width = "600px"}
# visualization
ggplot(ndf, aes(x=tend)) + 
  geom_line(aes(y=trans_prob, col=treat)) +
  geom_ribbon(aes(ymin = trans_lower, ymax = trans_upper, fill=treat), alpha = .3) +
  scale_color_manual(values = c("firebrick2"
                                , "steelblue")
                     )+
  scale_fill_manual(values = c("firebrick2"
                                , "steelblue")
                     )+
  facet_wrap(~transition) +
  xlim(c(0, 4000)) +
  ylim(c(0,1))+
  labs(y = "Transition Probability", x = "time", color = "Treatment", fill= "Treatment")

```



## Comparison of the results with Aalen-Johannsen estimator

Comparing the ``pammtools`` results with the ``mstate`` results, we want to validate that the baselines are indeed correct. The following code shows the comparison between the ``mstate``.

First, we compare the cumulative hazards
```{r, echo = FALSE, warning = FALSE, dependson=c("lib-ms-pammtools", "ms-prothr-ndf")}
# pammtools
ndf <- ndf |>
  mutate(package = "pammtools")
```
```{r prothr-prep-mstate, eval=TRUE, echo = FALSE, warning = FALSE, dependson=c("lib-ms-pammtools")}
library(mstate)
library(msm)
library(mvna)
library(etm)

# code from mstate documentation
data(prothr, package = "mstate")
tmat <- attr(prothr, "trans")
pr0 <- subset(prothr, treat=="Placebo")
attr(pr0, "trans") <- tmat
pr1 <- subset(prothr, treat=="Prednisone")
attr(pr1, "trans") <- tmat
c0 <- coxph(Surv(Tstart, Tstop, status) ~ strata(trans), data=pr0)
c1 <- coxph(Surv(Tstart, Tstop, status) ~ strata(trans), data=pr1)


msf0 <- msfit(c0, trans=tmat)
msf1 <- msfit(c1, trans=tmat)

# plot hazards
mstate_dat0 <- msf0$Haz %>% mutate(transition = case_when(
  trans == 1 ~ "1->2",
  trans == 2 ~ "1->3",
  trans == 3 ~ "2->1",
  trans == 4 ~ "2->3",
  .default = "cens"
  )
  , treat = "Placebo")
mstate_dat1 <- msf1$Haz %>% mutate(transition = case_when(
  trans == 1 ~ "1->2",
  trans == 2 ~ "1->3",
  trans == 3 ~ "2->1",
  trans == 4 ~ "2->3",
  .default = "cens"
  )
, treat = "Prednisone")

mstate_dat <- rbind(mstate_dat0, mstate_dat1)

long_mstate <- mstate_dat %>% 
  rename(tend = time, cumu_hazard = Haz) %>% 
  mutate(package = "mstate") %>%
  select(tend, treat, transition, cumu_hazard, package)
long_msm <- ndf %>% 
  mutate(package = "pammtools") %>% 
  select(tend, treat, transition, cumu_hazard, package)

long_haz_df <- rbind(long_mstate, long_msm) %>%
  mutate(transition = case_when(transition == "1->2" ~ "0->1"
                                , transition == "1->3" ~ "0->2"
                                , transition == "2->1" ~ "1->0"
                                , transition == "2->3" ~ "1->2")
         )


# compare transition probabilities
alpha <- 0.05

pt0 <- probtrans(msf0, predt=0) # changed predt from 1000 to 0
pt1 <- probtrans(msf1, predt=0) # changed predt from 1000 to 0


pt0_df_12 <- data.frame(cbind(tend = pt0[[1]]$time
                              , trans_prob = pt0[[1]]$pstate2
                              , trans_lower = pt0[[1]]$pstate2 - pt0[[1]]$se2 * qnorm(1 - alpha / 2)
                              , trans_upper = pt0[[1]]$pstate2 + pt0[[1]]$se2 * qnorm(1 - alpha / 2))
                        ) %>%
  mutate(transition = "1->2")
pt0_df_13 <- data.frame(cbind(tend = pt0[[1]]$time
                              , trans_prob = pt0[[1]]$pstate3
                              , trans_lower = pt0[[1]]$pstate3 - pt0[[1]]$se3 * qnorm(1 - alpha / 2)
                              , trans_upper = pt0[[1]]$pstate3 + pt0[[1]]$se3 * qnorm(1 - alpha / 2)
                              )) %>%
  mutate(transition = "1->3")
pt0_df_21 <- data.frame(cbind(tend = pt0[[2]]$time
                              , trans_prob = pt0[[2]]$pstate1
                              , trans_lower = pt0[[2]]$pstate1 - pt0[[2]]$se1 * qnorm(1 - alpha / 2)
                              , trans_upper = pt0[[2]]$pstate1 + pt0[[2]]$se1 * qnorm(1 - alpha / 2)
                              )) %>%
  mutate(transition = "2->1")
pt0_df_23 <- data.frame(cbind(tend = pt0[[2]]$time
                              , trans_prob = pt0[[2]]$pstate3
                              , trans_lower = pt0[[2]]$pstate3 - pt0[[2]]$se3 * qnorm(1 - alpha / 2)
                              , trans_upper = pt0[[2]]$pstate3 + pt0[[2]]$se3 * qnorm(1 - alpha / 2)
                              )) %>%
  mutate(transition = "2->3")

pt0_df <- rbind(pt0_df_12, 
                pt0_df_13,
                pt0_df_21,
                pt0_df_23) %>%
  mutate(treat = "Placebo",
         package = "mstate")

pt1_df_12 <- data.frame(cbind(tend = pt1[[1]]$time
                              , trans_prob = pt1[[1]]$pstate2
                              , trans_lower = pt1[[1]]$pstate2 - pt1[[1]]$se2 * qnorm(1 - alpha / 2)
                              , trans_upper = pt1[[1]]$pstate2 + pt1[[1]]$se2 * qnorm(1 - alpha / 2))
                        ) %>%
  mutate(transition = "1->2")
pt1_df_13 <- data.frame(cbind(tend = pt1[[1]]$time
                              , trans_prob = pt1[[1]]$pstate3
                              , trans_lower = pt1[[1]]$pstate3 - pt1[[1]]$se3 * qnorm(1 - alpha / 2)
                              , trans_upper = pt1[[1]]$pstate3 + pt1[[1]]$se3 * qnorm(1 - alpha / 2)
                              )) %>%
  mutate(transition = "1->3")
pt1_df_21 <- data.frame(cbind(tend = pt1[[2]]$time
                              , trans_prob = pt1[[2]]$pstate1
                              , trans_lower = pt1[[2]]$pstate1 - pt1[[2]]$se1 * qnorm(1 - alpha / 2)
                              , trans_upper = pt1[[2]]$pstate1 + pt1[[2]]$se1 * qnorm(1 - alpha / 2)
                              )) %>%
  mutate(transition = "2->1")
pt1_df_23 <- data.frame(cbind(tend = pt1[[2]]$time
                              , trans_prob = pt1[[2]]$pstate3
                              , trans_lower = pt1[[2]]$pstate3 - pt1[[2]]$se3 * qnorm(1 - alpha / 2)
                              , trans_upper = pt1[[2]]$pstate3 + pt1[[2]]$se3 * qnorm(1 - alpha / 2)
                              )) %>%
  mutate(transition = "2->3")

pt1_df <- rbind(pt1_df_12, 
                pt1_df_13,
                pt1_df_21,
                pt1_df_23) %>%
  mutate(treat = "Prednisone",
         package = "mstate")

pt_df_mstate <- rbind(pt0_df, pt1_df)

pt_df_pamm <- ndf %>%
  mutate(package = "pammtools") %>%
  select(tend, treat, transition, trans_prob, trans_lower, trans_upper, package)

pt_df <- rbind(pt_df_pamm, pt_df_mstate)
```




```{r, eval=TRUE, echo = TRUE, fig.width = 8, fig.height = 4, out.width = "600px", dependson=c("prothr-prep-mstate")}
# # plot transitions
# ggplot(test_msm, aes(x=tend, y=trans_prob)) + 
#   geom_line(aes(col=as.factor(treat))) + 
#   facet_wrap(~transition, ncol = 2, labeller = label_both) +
#   # scale_color_manual(values = c("#1f78b4", "#1f78b4", "#33a02c", "#33a02c"))+
#   # scale_linetype_manual(values = c("solid", "dashed", "solid", "dashed")) +
#   ylim(c(0,0.8)) +
#   xlim(c(0, 4000)) +
#   ylab("Transition Probability") +
#   xlab("time") +
#   theme_bw() 

comparison_nelaal <- ggplot(long_haz_df, aes(x=tend, y=cumu_hazard, col=treat, linetype = package)) + 
  geom_line() +
  facet_wrap(~transition, ncol = 4, labeller = label_both) +
  scale_color_manual(values = c("firebrick2"
                                , "steelblue")
                     )+
  # scale_linetype_manual(values = c("solid", "dashed", "solid", "dashed")) +18:
  ylab("Cumulative Hazards") +
  xlab("time in days") +
  ylim(c(0,6)) +
  scale_linetype_manual(values=c("dotted", "solid")) +
  theme_bw()

comparison_nelaal
```
Second, we compare the transition probabilities, which are calculated using the cumulative hazards
```{r, eval=FALSE, echo = FALSE, fig.width = 8, fig.height = 4, out.width = "600px", dependson=c("prothr-prep-mstate")}
comparison_aaljoh <- ggplot(pt_df, aes(x=tend, y=trans_prob, col=treat, linetype = package)) + 
  geom_line() +
  # geom_ribbon(aes(ymin = trans_lower, ymax = trans_upper, fill=treat), alpha = .3) +
  facet_wrap(~transition, ncol = 4, labeller = label_both) +
  scale_color_manual(values = c("firebrick2"
                                , "steelblue")
                     )+
  # scale_linetype_manual(values = c("solid", "dashed", "solid", "dashed")) +18:
  ylab("Transition Probabilities") +
  xlab("time in days") +
  ylim(c(0,1)) +
  xlim(c(0, 4000)) +
  scale_linetype_manual(values=c("dotted", "solid")) +
  theme_bw()

comparison_aaljoh
```

Third, we compare the width of the confidence bands
```{r, eval=FALSE, echo = FALSE, fig.width = 8, fig.height = 4, out.width = "600px", dependson=c("prothr-prep-mstate")}
comparison_aaljoh_ci <- ggplot(pt_df, aes(x=tend, y=trans_prob, col=interaction(package, treat), linetype = package)) + 
  geom_line() +
  geom_ribbon(aes(ymin = trans_lower, ymax = trans_upper, fill=interaction(package, treat), alpha = package)) +
  facet_wrap(~treat+transition, ncol = 4, labeller = label_both) +
  scale_color_manual(values = c("mstate.Placebo" = "grey"
                                ,"mstate.Prednisone" = "grey"
                                , "pammtools.Placebo" = "firebrick2"
                                , "pammtools.Prednisone" = "steelblue")
                     )+
  scale_fill_manual(values = c("mstate.Placebo" = "grey"
                                ,"mstate.Prednisone" = "grey"
                                , "pammtools.Placebo" = "firebrick2"
                                , "pammtools.Prednisone" = "steelblue")
                     )+
  scale_alpha_manual(values = c("mstate" = 0.3, "pammtools" = 0.05))+
  # scale_linetype_manual(values = c("solid", "dashed", "solid", "dashed")) +18:
  ylab("Transition Probabilities") +
  xlab("time in days") +
  ylim(c(0,1)) +
  xlim(c(0, 4000)) +
  scale_linetype_manual(values=c("dotted", "solid")) +
  theme_bw()+
  guides(color = "none", fill = "none", alpha = "none", linetype = "none")

comparison_aaljoh_ci
```

```{r, eval=TRUE, echo = TRUE, fig.width = 8, fig.height = 4, out.width = "600px", dependson=c("prothr-prep-mstate")}
comparison_aaljoh_ci <- ggplot(pt_df, aes(x=tend, y=trans_prob, col=interaction(package, treat), linetype = package)) + 
  geom_line(linetype = "solid") +
  geom_line(aes(y=trans_lower), linetype = "dotted") +
  geom_line(aes(y=trans_upper), linetype = "dotted") +
  geom_ribbon(data = subset(pt_df, package == "mstate"), aes(ymin = trans_lower, ymax = trans_upper), fill = "grey", alpha = 0.3) +
  facet_wrap(~treat+transition, ncol = 4, labeller = label_both) +
  ylab("Transition Probabilities") +
  xlab("time in days") +
  ylim(c(0,1)) +
  xlim(c(0, 4000)) +
  scale_color_manual(values = c("mstate.Placebo" = "grey"
                              ,"mstate.Prednisone" = "grey"
                              , "pammtools.Placebo" = "firebrick2"
                              , "pammtools.Prednisone" = "steelblue")
                   )+
  theme_bw()+
  guides(color = "none", fill = "none", linetype = "none")

comparison_aaljoh_ci
```
