---
title: "Shape-constrained effects (scam)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Shape-constrained effects (scam)}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: Remote.bib
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  fig.align  = "center",
  cache      = FALSE,
  message    = FALSE,
  fig.height = 4.5,
  fig.width  = 4.5)
```

```{r, message = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
theme_set(theme_bw())
library(survival)
library(mgcv)
library(scam)
library(pammtools)
Set1 <- RColorBrewer::brewer.pal(9, "Set1")
```

In many applications there is good reason to believe that an effect is
monotone or otherwise shape-restricted, e.g.,

  - a baseline hazard that decreases monotonically after a successful
  operation or increases monotonically for wear-out processes,

  - a dose-response relationship that is known to be non-decreasing.

Unconstrained estimates of such effects can violate these assumptions due to
sampling variability, especially in regions with little data. *Shape-constrained
additive models* (SCAMs; @pya2015), implemented in the
[**`scam`**](https://cran.r-project.org/package=scam) package, extend
**`mgcv`**'s penalized spline approach such that individual smooth terms can be
constrained to be monotone (and/or convex/concave) by using specialized spline
basis functions (e.g., `bs = "mpd"` for *m*onotone *d*ecreasing or
`bs = "mpi"` for *m*onotone *i*ncreasing smooths).

Since PAMMs are estimated as Poisson GAMs, all of this functionality carries
over to hazard regression: simply replace the call to `mgcv::gam` with a call
to `scam::scam` (or use `pamm(..., engine = "scam")`). The entire
post-processing workflow provided by **`pammtools`** (`make_newdata`,
`add_hazard`, `add_cumu_hazard`, `add_surv_prob`, `add_term`, `add_cif`,
`tidy_smooth`, `gg_smooth`, etc.) works for `scam` fits exactly as it does for
`gam` fits, including delta-method and simulation-based confidence intervals.


## Monotone baseline hazard

We illustrate the approach with the `tumor` data, where interest could be in
a baseline hazard of dying that is assumed to *decrease* monotonically with
time since the (successful) operation. First we transform the data to the PED
format and fit both, an unconstrained PAM and a PAM with monotonically
decreasing baseline hazard (`bs = "mpd"`):

```{r, fit-baseline}
ped <- tumor %>% as_ped(
  Surv(days, status) ~ age + complications,
  cut = seq(0, 2000, by = 100))
# unconstrained fit
pam <- gam(ped_status ~ s(tend) + complications,
  data = ped, family = poisson(), offset = offset)
# fit with monotonically decreasing baseline hazard
mpam <- scam(ped_status ~ s(tend, bs = "mpd") + complications,
  data = ped, family = poisson(), offset = offset)
```

Comparison of the estimated baseline hazards shows that the unconstrained
estimate increases again towards the end of the follow-up, where events are
rare and uncertainty is large, while the constrained estimate decreases
monotonically throughout, with narrower confidence intervals in that region:

```{r, fig-baseline, fig.width = 7}
ndf <- ped %>% make_newdata(tend = unique(tend))
haz_df <- bind_rows(
  ndf %>% add_hazard(pam)  %>% mutate(model = "unconstrained (gam)"),
  ndf %>% add_hazard(mpam) %>% mutate(model = "monotone (scam)"))

ggplot(haz_df, aes(x = tend, y = hazard)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = model), alpha = .3) +
  geom_line(aes(col = model)) +
  scale_color_manual(values = Set1) +
  scale_fill_manual(values = Set1) +
  xlab("time") + ylab("hazard")
```

All other convenience functions work as usual, e.g., cumulative hazards
and survival probabilities (here stratified by `complications`):

```{r, fig-surv, fig.width = 7}
surv_df <- ped %>%
  make_newdata(tend = unique(tend), complications = unique(complications)) %>%
  group_by(complications) %>%
  add_surv_prob(mpam)

ggplot(surv_df, aes(x = tend, y = surv_prob)) +
  geom_line(aes(col = complications)) +
  geom_ribbon(
    aes(ymin = surv_lower, ymax = surv_upper, fill = complications),
    alpha = .3) +
  scale_color_manual(values = Set1) +
  scale_fill_manual(values = Set1) +
  ylim(c(0, 1)) + xlab("time") + ylab("survival probability")
```


## Monotone covariate effects

Shape constraints can be imposed on any smooth term, not just the baseline
hazard, and constrained and unconstrained terms can be combined freely within
one model. Below, the effect of `age` is constrained to be monotonically
*increasing* (`bs = "mpi"`), while the baseline hazard is constrained to be
monotonically decreasing as before:

```{r, fit-age}
mpam2 <- scam(
  ped_status ~ s(tend, bs = "mpd") + s(age, bs = "mpi") + complications,
  data = ped, family = poisson(), offset = offset)
pam2 <- gam(ped_status ~ s(tend) + s(age) + complications,
  data = ped, family = poisson(), offset = offset)
```

The estimated covariate effects can be extracted and visualized with
`add_term` (here centered around the mean age, on the log-hazard scale):

```{r, fig-age, fig.width = 7}
age_df <- ped %>% make_newdata(age = seq_range(age, n = 100))
age_df <- bind_rows(
  age_df %>% add_term(pam2, term = "age", reference = list(age = mean(.$age))) %>%
    mutate(model = "unconstrained (gam)"),
  age_df %>% add_term(mpam2, term = "age", reference = list(age = mean(.$age))) %>%
    mutate(model = "monotone (scam)"))

ggplot(age_df, aes(x = age, y = fit)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = model), alpha = .3) +
  geom_line(aes(col = model)) +
  scale_color_manual(values = Set1) +
  scale_fill_manual(values = Set1) +
  xlab("age") + ylab("log-hazard (centered)")
```

The usual tidiers and plotting functions also accept `scam` fits, e.g.:

```{r, fig-gg-smooth, fig.width = 7, fig.height = 3.5}
gg_smooth(ped, mpam2, terms = c("tend", "age"))
```


## Convenience: `pamm(..., engine = "scam")`

The `pamm` wrapper, which sets the Poisson family and the offset
automatically, also supports `scam` as fitting engine:

```{r}
mpam3 <- pamm(
  ped_status ~ s(tend, bs = "mpd") + complications,
  data = ped, engine = "scam")
summary(mpam3)
```


## Notes on confidence intervals

`scam` estimates shape-constrained smooths by re-parametrizing the
coefficients of the constrained terms (the model is non-linear in the
underlying parameters; see @pya2015 for details). All **`pammtools`**
functions are based on the re-parametrized coefficients and their covariance
matrix, i.e., on the same (approximate) distribution that `scam` itself uses
to compute standard errors and confidence intervals.

This applies to all three types of confidence intervals offered by
`add_hazard`, `add_cumu_hazard` and `add_surv_prob` (`ci_type = "default"`,
`"delta"` and `"sim"`), as well as the simulation-based intervals of
`add_cif`, `add_trans_prob` and `get_cumu_coef`. For example, the
simulation-based intervals for the survival probability:

```{r}
ndf %>%
  add_surv_prob(mpam, ci_type = "sim") %>%
  # drop the tend == 0 boundary row so the table shows the interval estimates
  filter(tend != 0) %>%
  select(tend, surv_prob, surv_lower, surv_upper) %>%
  slice(c(1:3, (n() - 2):n()))
```

Note that for terms whose estimate lies on the boundary of the constraint
(e.g., an effect that is shrunken to a constant by the monotonicity
penalty), the normal approximation underlying all three interval types is
the same one used by `scam` itself, but it can be less accurate than in the
unconstrained case.

Two further practical remarks:

  - Estimation using constraints takes somewhat longer than the respective
  unconstrained fit.

  - `scam` provides shape-constrained univariate (and bivariate) smooths;
  consult `?scam::shape.constrained.smooth.terms` for the list of available
  constraints and basis types. Unconstrained terms in a `scam` formula use
  the standard **`mgcv`** bases.


## References

<!-- Zaehlmarke VGWort -->
<img src="https://vg09.met.vgwort.de/na/f993b9c06b5249adb509a4df30d807a6" width="1" height="1" alt="">
