---
title: "Splines"
author: "Andreas Bender"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{splines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: pam.bib
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  fig.align  = "center",
  cache      = TRUE,
  message    = FALSE,
  fig.height = 5,
  fig.width  = 5)
```


```{r}
library(magrittr)
library(dplyr)
library(ggplot2)
theme_set(theme_bw())
library(pam)
library(mgcv)
library(survival)
Set1 <- RColorBrewer::brewer.pal(9, "Set1")
```

## Veteran's data 

We first illustrate the estimation on non-linear smooth effects using 
the `veteran` data from the survival package: 

```{r}
data("veteran", package="survival")
veteran %<>% 
  mutate(
    trt   = 1+(trt == 2),
    prior = 1*(prior==10)) %>% 
  filter(time < 400)
head(veteran)
```

There are two continuous covariates in the data set that could potentially 
exhibit non-linear effects, `age` and `karno`. We fit a Cox-PH model and 
a PAM of the form 

$$
\lambda(t|x) = \lambda_0(t)\exp(f(x_{karno})) 
$$

and 

$$
\lambda(t|x) = \exp(f_0(t_j) + f(x_{karno}))
$$
respectively: 

```{r}
## Cox-PH 
cph <- coxph(Surv(time, status) ~ pspline(karno, df=0), data=veteran)
## PAM 
ped <- split_data(Surv(time, status) ~ ., data=veteran, id="id")
pam <- gam(ped_status ~ s(tend) + s(karno, bs="ps"), data=ped, family=poisson(), 
  offset=offset)
```


```{r}
# crate data set with varying karno values (from min to max in 100 steps) 
# and add term contribution of karno from PAM and Cox-PH models
karno_df <- ped %>% make_newdata(expand = "karno", n = 100) %>% 
  add_term(pam, term="karno") %>% 
  mutate(
    cox = predict(cph, newdata=., type="terms")[, "pspline(karno, df = 0)"] - 
      predict(cph, newdata=data.frame(karno=mean(veteran$karno))))
ggplot(karno_df, aes(x=karno, ymin=low, ymax=high)) +
  geom_line(aes(y=fit, col="PAM")) +
  geom_ribbon(alpha=0.2) +
  geom_line(aes(y=cox, col="Cox PH"))+
  scale_colour_manual(name="Method",values=c("Cox PH"=Set1[1],"PAM"=1)) +
  xlab(expression(x[plain(karno)])) + ylab(expression(hat(f)(x[plain(karno)])))
```

Both methods are in good agreement. The higher the Karnofsky-Score the lower the
expected log-hazard. For further evaluation of the Karnofsky Score effect using
time-varying terms of the form $f(t)\cdot x_{karno}$ and $f(x_{karno}, t)$ see 
`vignette("tveffects", package="pam")`. 


## MGUS data: age example

In the following we consider another example using  using data presented in the 
respective vignette in the `survival` package 
(see `vignette("splines", package = "survival")`).


### Cox PH workflow

The example presented in the vignette goes as follows, using the standard 
base R workflow and `termplot` for visualization: 

```{r}
data("mgus", package = "survival")
mfit <- coxph(Surv(futime, death) ~ sex + pspline(age, df = 4), data = mgus)
mfit
```

```{r, figcox}
termplot(mfit, term = 2, se = TRUE, col.term = 1, col.se = 1)
```


### PAM workflow
The equivalent fit using PAMs requires the additional step of transforming the 
data into a suitable format. We then use `mgcv::gam` to fit 
the model: 

```{r}
mgus.ped <- split_data(Surv(futime, death)~sex + age, data = mgus, id = "id")
head(mgus.ped)
pamfit <- gam(ped_status ~ s(tend) + sex + s(age, bs = "ps"), data = mgus.ped, 
  method = "REML", family = poisson(), offset = offset)
summary(pamfit)
```

For visualization of the smooth effects we can use the default `mgcv::plot.gam` 
function: 

```{r, figpam1, fig.width = 7, fig.height = 3.5}
layout(matrix(1:2, nrow = 1))
termplot(mfit, term = 2, se = TRUE, col.term = 1, col.se = 1)
plot(pamfit, select = 2, se = TRUE, ylim = c(-4, 3.5))
```

In this example the PAM approach estimates a linear effect of age, which is 
consistent with the estimation using `coxph`, as there is only weak non-linearity.

### MGUS data: Hemoglobin example
Another example from the same vignette shows that estimated effects 
are very similar if the effect of the covariate is in fact strongly non-linear: 

```{r}
fit <- coxph(Surv(futime, death) ~ age + pspline(hgb, 4), mgus2)
mgus2.ped <- split_data(Surv(futime, death)~age + hgb, data = mgus2, id = "id")
pamfit2 <- gam(ped_status~s(tend) + age + s(hgb), data = mgus2.ped, 
	family = poisson(), offset = offset)
```
  
```{r, figpam, fig.width = 7, fig.height = 3.5}
layout(matrix(1:2, nrow = 1))
termplot(fit, term = 2, se = TRUE, col.term = 1, col.se = 1)
plot(pamfit2, select = 2, se = TRUE, ylim = c(-0.5, 2))
```

### Monotonicity constraints
The vignette in the `survival` package further discusses enforcing monotonicity 
contraints on the effect of `hgb`. These can be achieved here more easily 
using the functionality provided in the `scam` package (see also @pya2015). 
The usage is exactly the same as before, replacing the call to `gam` with a 
call to `scam` and specifying `bs = "mpd"`. Note that the fit using contraints 
takes considerably longer compared to the unconstrained fit. 

```{r, scammod}
library(scam)
mpam <- scam(ped_status ~ s(tend) + age + s(hgb, bs = "mpd"), data = mgus2.ped, 
  family = poisson(), offset = offset)
```

```{r, figscam, fig.height = 3.5, fig.width = 7, dependson="scammod"}
layout(matrix(1:2, nrow = 1))
plot(pamfit2, select = 2, se = TRUE, ylim = c(-0.5, 2), main="Unconstrained fit")
# 1.72 = intercept difference between pamfit2 and mpam
const <- abs(coef(pamfit2)[1] - coef(mpam)[1])
plot(mpam, select = 2, se = TRUE, shift = const, 
  ylim = c(-0.5-const, 2-const), main="Fit with monotonicity constraint")
```

## References
