---
title: "Stratified baseline"
author: "Andreas Bender"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{strata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: pam.bib
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  fig.align  = "center",
  cache      = TRUE,
  message    = FALSE,
  fig.height = 5,
  fig.width  = 5
)
```

```{r, message=FALSE}
library(magrittr)
library(dplyr)
library(ggplot2)
theme_set(theme_bw())
library(mgcv)
library(survival)
library(pam)
Set1 <- RColorBrewer::brewer.pal(9, "Set1")
```

## Stratified baselines 

In some cases the proportional hazards assumption for different groups 
(levels of a factor variable) is violated. One approach to resolve this 
problem is to fit a so called *stratified* Cox model, where each level 
$k=1,\ldots,K$ of factor variable $z$ will have its own baseline-hazard: 

$\lambda(t|z, x)=\lambda_{0k}(t, z)\exp(x'\beta)$. 

For illustration we again use the `veteran` data (censored at day 400) 
from the `survival` package, 
where the celltype of the lung cancer had four levels: 

```{r}
data("veteran")
veteran %<>% 
  mutate(
    prior = 1*(prior==10),
    status = ifelse(time > 400, 0, status),
    time = ifelse(time > 400, 400, time))
table(veteran$celltype)
```

### Stratified Cox model
Fitting a stratified Cox model using the `coxph` function from the `survival`
package is simple, including a `strata` term in the model formula. 

```{r, cph}
cph <- coxph(Surv(time, status) ~ strata(celltype), data=veteran)
```

The different baselines are visualized below: 
```{r, ggbase, fig.width=5, fig.height=4, dependson="cph", warning=FALSE}
base <- basehaz(cph, centered=FALSE)
gg.bz <- ggplot(base, aes(x=time)) +
  geom_step(aes(y=hazard, group=strata)) +
	ylab(expression(hat(Lambda)(t))) + xlab("t")
gg.bz + aes(col=strata)
```

### Stratified PAM

We can obtain similar results with PAMs by fitting one smooth baseline 
estimate for each category of `celltype` in the data. 
Technically, we specify the `by` argument in `mgcv` `s(...)` function. 
Note that in contrast to the model specification in `coxph` we need to 
include the `celltype` factor variable in the model specification:

```{r, fig.width=7, fig.height=3, warning=FALSE}
ped <- split_data(Surv(time, status)~., data=veteran, id="id")
pam <- gam(ped_status ~ celltype + s(tend, by=celltype, m=1), data=ped, 
	family=poisson(), offset=offset)
```
<!-- mit m = 1 hier bisschen weniger abgefahrenes Ergebnis als mit m=default =2
für die 'adeno'-subgruppe..... da haste ja bisher ganz schön geschummelt mit dem ylim... ;)-->

Comparing the estimated baselines we get very similar results between 
the stratified Cox and stratified PAM models. 
```{r, fig.width=6, fig.height=3, dependson="ggbase", warning=FALSE}
pinf <- ped %>% 
  group_by(celltype) %>% 
  ped_info() %>% 
  add_cumhazard(pam) %>% 
	rename(strata=celltype)
gg.bz +	aes(col="Nelson-Aalen") +
	geom_line(data=pinf, aes(x=tend, y=cumhazard, group=strata, col="PAM")) +
  geom_ribbon(data=pinf, aes(x=tend, ymin=cumlower, ymax = cumupper,
    group=strata), fill = "black", col = NA, alpha = .2) +
 	facet_wrap(~strata, nrow=1) + 
	scale_color_manual(name="Method", values=c(Set1[1], 1)) +
  theme(legend.position = "bottom")
```

### Stratified Proportional Hazards Model 

Both models can be easily extended to incorporate further covariates. 
If we assume that the covariate effects are not time-varying, this extension 
constitutes a so called stratified proportional hazards model. For example, 
including effects for treatment (`trt`), age (`age`) and the Karnofsky Score (`karno`), 
we get the model 

$$
\lambda(t|z, \mathbf{x}) =
  \lambda_{0k}(t)\exp(\beta_1 I(trt=1) + \beta_2\cdot x_{karno} + \beta_3\cdot x_{age})
$$
where $\lambda_{0k}(t)$ is the baseline for observation with celltype $k$,
$k\in \{squamos,\ smallcell,\ adeno,\ large\}$:

```{r}
cph2 <- update(cph, .~. + trt + age + karno)
pam2 <- update(pam, .~. + trt + age + karno)
cbind(pam=coef(pam2)[5:7], cox=coef(cph2))
```

Again, covariate estimates are very similar between the two models and 
can be interpreted in the same way. 
