---
title: "Stratified baseline"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stratified baseline}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: Remote.bib
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  fig.align  = "center",
  cache      = TRUE,
  message    = FALSE,
  fig.height = 5,
  fig.width  = 5
)
```

```{r, message=FALSE}
library(dplyr)
library(ggplot2)
theme_set(theme_bw())
library(mgcv)
library(survival)
library(pammtools)
Set1 <- RColorBrewer::brewer.pal(9, "Set1")
```

## Stratified baselines

In some cases the proportional hazards assumption for different groups
(levels of a factor variable) is violated. One approach to resolve this
problem is to fit a so called *stratified* Cox model, where each level
$k=1,\ldots,K$ of factor variable $z$ will have its own baseline-hazard:

$\lambda(t|z, x)=\lambda_{0k}(t, z)\exp(x'\beta)$.

For illustration we again use the `veteran` data (censored at day 400)
from the `survival` package,
where the cell type of the lung cancer had four levels:

```{r}
veteran <- survival::veteran # load veteran data
veteran <- veteran %>%
  mutate(
    trt   = 1*(trt == 2),
    prior = 1*(prior==10)) %>%
  filter(time < 400) # restriction for illustration
table(veteran$celltype)
```

### Stratified Cox model
Fitting a stratified Cox model using the `coxph` function from the `survival`
package is simple, including a `strata` term in the model formula.

```{r, cph}
cph  <- coxph(Surv(time, status) ~ strata(celltype), data=veteran)
base <- basehaz(cph)
```

The different baselines are visualized below:
```{r, ggbase, fig.width=6, fig.height=4, dependson="cph", warning=FALSE}
baseline_gg <- ggplot(base, aes(x=time)) +
  geom_step(aes(y=hazard, group=strata)) +
	ylab(expression(hat(Lambda)(t))) + xlab("t")
baseline_gg + aes(col=strata)
```

### Stratified PAM

We can obtain similar results with PAMs by fitting one smooth baseline
estimate for each category of `celltype` in the data.
Technically, we specify the `by` argument in `mgcv` `s(...)` function.

Note that in contrast to the model specification in `coxph` we need to
include the `celltype` factor variable in the model specification:

```{r, fig.width=7, fig.height=3, warning=FALSE}
ped <- veteran %>% as_ped(Surv(time, status)~., id="id")
pam <- gam(ped_status ~ celltype + s(tend, by=celltype), data=ped,
	family=poisson(), offset=offset)
```

Comparing the estimated baselines we get very similar results between
the stratified Cox and stratified PAM models.

```{r, fig.width=6, fig.height=3, dependson="ggbase", warning=FALSE}
pinf <- ped %>%
  group_by(celltype) %>%
  ped_info() %>%
  add_cumu_hazard(pam) %>%
	rename(strata=celltype) %>%
  filter(!(strata == "adeno" & tend > 200))
baseline_gg +	aes(col="Nelson-Aalen") +
	geom_line(data=pinf, aes(x=tend, y=cumu_hazard, group=strata, col="PAM")) +
  geom_ribbon(data=pinf, aes(x=tend, ymin=cumu_lower, ymax = cumu_upper,
    group=strata), fill = "black", col = NA, alpha = .2) +
 	facet_wrap(~strata, nrow=1) +
	scale_color_manual(name="Method", values=c(Set1[1], "black")) +
  theme(legend.position = "bottom")
```

### Stratified Proportional Hazards Model

Both models can be easily extended to incorporate further covariates.
If we assume that the covariate effects are not time-varying, this extension
constitutes a so called stratified proportional hazards model. For example,
including effects for treatment (`trt`), age (`age`) and the Karnofsky-Score
(`karno`), we get the model

$$
\lambda(t|z, \mathbf{x}) =
  \lambda_{0k}(t)\exp(\beta_1 I(trt=1) + \beta_2\cdot x_{karno} + \beta_3\cdot x_{age})
$$
where $\lambda_{0k}(t)$ is the baseline for observation with cell type $k$,
$k\in \{squamos,\ smallcell,\ adeno,\ large\}$:

```{r}
cph2 <- update(cph, .~. + trt + age + karno)
pam2 <- update(pam, .~. + trt + age + karno)
cbind(pam=coef(pam2)[5:7], cox=coef(cph2))
```

Again, covariate estimates are very similar between the two models and
can be interpreted in the same way.
