---
title: "Time-dependent covariates"
author: "Andreas Bender"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tdcovar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: pam.bib
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  fig.align  = "center",
  cache      = TRUE,
  message    = FALSE,
  fig.height = 5,
  fig.width  = 5
)
```

```{r, message=FALSE}
library(magrittr)
library(tidyr)
library(dplyr)
library(ggplot2)
theme_set(theme_bw())
library(survival)
library(mgcv)
library(pam)
Set1 <- RColorBrewer::brewer.pal(9, "Set1")
```

## Analysis of the recidivism data 

In the following, we demonstrate an analysis containing time-dependent covariates, 
using the well-known recidivism data discussed in detail in @Fox2011. 
The R-Code of the original analysis using the extended Cox model can be found 
[here](http://socserv.mcmaster.ca/jfox/Books/Companion/scripts/appendix-cox.R), 
the respective vignette [here](https://socserv.socsci.mcmaster.ca/jfox/Books/Companion-1E/appendix-cox-regression.pdf).

```{r}
# raw data
# http://socserv.mcmaster.ca/jfox/Books/Companion/scripts/appendix-cox.R
recidivism <- read.table(
    file   = "http://math.unm.edu/~james/Rossi.txt",
    header = TRUE) %>%
  mutate(subject=row_number())
```


### Data preprocessing 

In this example we don't need a dedicated function for transformation, 
as we basically just need to transform the data into long format (equals 
splitting at each week for which subjects are in the risk set):

```{r}
# transform into long format
recidivism_long <- recidivism %>%
  gather(calendar.week, employed, emp1:emp52) %>%
  filter(!is.na(employed)) %>% # employed unequal to NA only for intervals under risk
  group_by(subject) %>%
  mutate(
    start  = row_number()-1,
    stop   = row_number(),
    arrest = ifelse(stop == last(stop) & arrest == 1, 1, 0),
    offset = log(stop - start)) %>%
  select(subject, start, stop, offset, arrest, employed, fin:educ) %>%
  arrange(subject, stop)
recidivism_long %<>% 
  mutate(employed.lag1 = lag(employed, default=0)) %>% 
  slice(-1) %>% # exclusion of first week, as lagged information is missing
  ungroup()
```


### Fitting the models 

Below we fit a PAM and an extended Cox model. In this case the format for 
both models is the same (which is not always the case for analyses with 
time-dependent covariates, see the second example below using the `pbc` data):
The `stop` variable defines the interval endpoints and is used to model
the baseline log hazard rates.

```{r}
## Fit PAM (smooth effects of age and prio, using P-Splines)
pam <- gam(arrest ~ s(stop) + fin + s(age, bs="ps") + race + wexp + mar + paro + 
  s(prio, bs="ps") + employed.lag1,
  data=recidivism_long, family=poisson(), offset=offset)
tidy_fixed(pam)
## respective extended cox model 
cph <- coxph(
  formula = Surv(start, stop, arrest)~ fin + pspline(age) + race + wexp + mar + 
    paro + pspline(prio) + employed.lag1,
  data=recidivism_long)
# extract information on fixed coefficients
tidy_fixed(cph)[c(1, 4:7, 10), ]
```

### Graphical comparison of the two models 
The figure below summarizes the comparison between the two models.

<details> 
	<summary>Expand here for R-Code</summary>

```{r, prisonPrep}
all_eff <- purrr::map_df(
  list(
    tidy_fixed(pam), 
   	tidy_fixed(cph)[-c(2:3, 8:9), ]), 
  bind_rows, .id="Model") %>% 
  mutate(Model = factor(Model, levels=2:1, labels=c("Cox-PH", "PAM")))

## plot of fixed coefficients
coef_gg <- ggplot(all_eff, aes(x=variable, y=coef, ymin=lower, ymax=upper)) + 
  geom_hline(yintercept = 0, lty=3) +
  geom_pointrange(aes(col=factor(Model), shape=factor(Model)),
    position=position_dodge(width=0.5)) +
  scale_colour_manual(
    name   = "Method",
    values = c(1, Set1[1]),
    limits = rev(levels(all_eff$Model))) +
  scale_shape_manual(
    name   = "Method",
    values = c(19, 15),
    limits = rev(levels(all_eff$Model))) +
  coord_flip(ylim=range(-1.5, 1)) +
  ylab(expression(hat(beta)%+-% 1.96 %.% SE)) + 
  xlab("")

## to visualize smooth effect of age, create data set where all covariates are 
## fixed to mean values except for age, which varies between min and max 
## (n = 100)
age_df <- recidivism_long %>% make_newdata(expand="age", n=100)

## add information on contribution of age to linear predictor (partial effect of age)
age_df %<>% add_term(pam, term="age") %>% 
  mutate(cphfit = predict(object=cph, ., type="terms")[,"pspline(age)"])

## prep plot object for smooth effects 
smooth_gg <- ggplot(age_df, aes(y=fit)) + 
	geom_line(aes(col="PAM")) + 
	geom_ribbon(aes(ymin=low, ymax=high), alpha=0.3) + 
	geom_line(aes(y=cphfit, col="Cox-PH")) + 
	scale_colour_manual(name="Method", values=c("#E41A1C", "#000000")) + 
  ylab(expression(hat(f)(x))) + theme(legend.position="none")

## plot of the age effect 
age_gg <- smooth_gg + aes(x=age) + 
  xlab("Age")


## same as "age"" for "prio" variable 
prio_df <- recidivism_long %>% make_newdata(expand="prio", n=100)
prio_df %<>% add_term(pam, term="prio") %>% 
  mutate(cphfit = predict(object=cph, ., type="terms")[,7])

## plot of the prio effect 
prio_gg <- smooth_gg %+% prio_df + aes(x=prio) + 
  xlab("Number of prior convictions")
```

</details>

As we can see, the estimates of the fixed coefficients (left panel) are 
very similar between the two models, including the confidence intervals. 
Using the default settings in both model specifications (using P-Splines 
for smooth terms), the PAM estimates are smoother compared to the Cox 
estimates (right panel). 

```{r, fig.width=7, fig.height=5, message=FALSE, dependson="prison_prep"}
## put all plots together 
library(gridExtra)
grid.arrange(
  coef_gg +theme(legend.position="bottom"), 
  age_gg, 
  prio_gg, 
  layout_matrix=matrix(c(1, 1, 2, 3), ncol=2))
```


## Analysis of the `pbc` data

Here we show an example with continuous time-dependent covariates using the 
Primary Biliary Cirrhosis Data (`pbc`) from the `survival` package (see 
`?pbc` for documentation).

```{r}
head(pbc)[, c(1:5, 11, 12)]
head(pbcseq)[, c(1, 4:5, 7, 12, 13)]
```

### Extended Cox analysis of the `pbc` data
We first replicate the analysis from `vignette("timedep", package="survival")`:

```{r}
# below code copied from survival vignette "timedep"
temp <- subset(pbc, id <= 312, select=c(id:sex, stage)) # baseline
pbc2 <- tmerge(temp, temp, id=id, death = event(time, status)) #set range
pbc2 <- tmerge(pbc2, pbcseq, id=id, ascites = tdc(day, ascites), 
    bili = tdc(day, bili), albumin = tdc(day, albumin), 
    protime = tdc(day, protime), alk.phos = tdc(day, alk.phos))

fit1 <- coxph(Surv(time, status==2) ~ log(bili) + log(protime), pbc)
fit2 <- coxph(Surv(tstart, tstop, death==2) ~ log(bili) + log(protime), pbc2)
rbind("baseline fit" = coef(fit1), "time dependent" = coef(fit2))
```

This demonstrates that results can differ substantially if only the 
baseline values of TDCs are used for the analysis instead of their complete trajectories over time. 


### PAM analysis of the `pbc` data

First we need to transform the data to PED format. In order to do so we 
perform some minor preprocessing: 

```{r}
## data set that only contains baseline (t=0) information
event_df <- pbc %>% 
  filter(id <= 312) %>% 
  mutate(event = 1*(status==2)) %>% 
  select(id, time, event, trt, age, sex, bili, protime)

## we rename the "day" variable here for use of `split_tdc` function later
tdc_df <- pbcseq %>% 
  select(id, day, bili, protime) %>% 
  rename(time = day)
```

Now we can use the `?split_tdc` function, that needs both data sets, 
`event_df`, that contains time-to-event information and one row per subject and 
`tdc_df` that contains information on timing and value of time-dependent 
covariates. 

The `split_tdc` function performs the following tasks: 

  - extract unique event times as well as all measurement times of the TDCs
  - split the follow-up into intervals according to these times (essentially calling 
  `split_data` with above cut-points on `event_df` after removing all TDCs)
  - merge the expanded data set with the data set containing information 
  on TDCs by ID and time
  - fill in the values of TDCs for any time-points that did not occur in `tdc_df` by 
  carrying the respective previous value of the TDC forward. 

```{r}
pbc_ped <- split_tdc(Surv(time, event) ~., event_df, tdc_df, "id", "time", "status")
filter(tdc_df, id==17)
pbc_ped %>% 
  filter(id==17) %>% 
  semi_join(tdc_df, by=c("id", "tstart"="time")) %>% 
  select(id, interval, age, bili, protime)
```


Now we can fit the model as usual calling `mgcv::gam`: 

```{r}
pbc_pam <- gam(ped_status ~ s(tend) + log(bili) + log(protime), data=pbc_ped, 
  family=poisson(), offset=offset)
cbind(pam=coef(pbc_pam)[2:3], cox=coef(fit2))
```

Coefficient estimates are very similar for both models, especially for the 
effect of `bili`. A graphical comparison yields similar results: 

<details>
  <summary>Expand here for R-Code</summary>
  
```{r, fig.keep="none"}
## Effect of bilirubin
## note that we use relative = TRUE in the call to add_term, which calculates 
## the relative risk change (x - \bar{x})'\beta for comparison with predict.coxph
## (see also Details section in ?predict.coxph)
bili_df <- pbc_ped %>% ungroup() %>% 
  make_newdata(expand="bili", n = 100) %>%
  add_term(pbc_pam, term="bili", relative=TRUE) %>% 
  mutate(cox = predict(fit2, ., type="term")[, "log(bili)"])

bili_gg <- ggplot(bili_df, aes(x=bili, y=fit)) + 
    geom_line(aes(col="PAM")) + 
    geom_ribbon(aes(ymin=low, ymax=high), alpha=0.2) + 
    geom_line(aes(y=cox, col="Cox")) +
    scale_colour_manual(name="Method", values=c("#E41A1C", "#000000")) 

## Effect of protime
protime_df <- pbc_ped %>% ungroup() %>% 
  make_newdata(expand="protime", n = 100) %>% 
  add_term(pbc_pam, term="protime", relative=TRUE) %>% 
  mutate(cox = predict(fit2, ., type="term")[, "log(protime)"])

protime_gg <- ggplot(protime_df, aes(x=protime, y=fit)) + 
    geom_line(aes(col="PAM")) + 
    geom_ribbon(aes(ymin=low, ymax=high), alpha=0.2) + 
    geom_line(aes(y=cox, col="Cox")) + 
    scale_colour_manual(name="Method", values=c("#E41A1C", "#000000"))

## combine figures
library(grid)
grid.draw(
    cbind(
        ggplotGrob(bili_gg),
        ggplotGrob(protime_gg),
        size = "last"))
```
</details>

```{r, echo=FALSE, fig.width=6, fig.height=3}
library(grid)
grid.draw(
    cbind(
        ggplotGrob(bili_gg),
        ggplotGrob(protime_gg),
        size = "last"))
```


## References
