---
title: "Time-varying effects"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time-varying effects}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  fig.align  = "center",
  fig.width  = 4,
  fig.height = 4,
  crop       = TRUE)
```

```{r, message = FALSE}
library(tidyr)
library(dplyr)
library(ggplot2)
theme_set(theme_bw())
library(survival)
library(mgcv)
library(pammtools)
Set1    <- RColorBrewer::brewer.pal(9, "Set1")
Greens  <- RColorBrewer::brewer.pal(9, "Greens")
Purples <- RColorBrewer::brewer.pal(9, "Purples")
```

In this vignette we show examples of how to fit time-varying effects of
time-constant **continuous** covariates. Note that time-varying effects of time-constant
**categorical** variables are analogous to stratified proportional
hazards models, where observations from different levels of the categorical
variable have different baseline hazards. That setting is described in the
[stratification vignette](strata.html).
**Note that all time-varying effects in a PAM are still assumed to be piece-wise
constant over the intervals used to specify the PAM!**

## Possible specifications of time-variation

In the following we denote the continuous time-constant covariate with $x$ and
time with $t$.
A time-varying effect of $x$ can then be specified as an interaction term between
$x$ and $t$, where different levels of complexity and flexibility for this
interaction are possible:

- $\beta_x\cdot x+\beta_{x:t}\cdot(x\cdot g(t))$: Linear effect of $x$ with
time-variation given by $g(t)$, where $g(\cdot)$ is a known or pre-specified
transformation of time $t$, e.g. the $\log$-function.

- $f_x(x)\cdot g(t)$: Non-linear effect of $x$, (linearly) time-varying with $g(t)$,
where $g(t)$ is a known or pre-specified transformation of time

- $f_t(t)\cdot x$: A *varying coefficient model* in $x$, where time-variation
is non-linear and estimated from the data. If $x$ is a dummy variable coding for
levels of a categorical variable this constitutes a stratified model with a
different "baseline" hazard for each category,
see the [*strata* vignette](strata.html).

- $f_{x,t}(x,t)$: A non-linear effect of $x$ that varies non-linearly over time $t$.

## Veteran Data example

For illustration and comparison we use the `veteran` data presented in the
vignette of the `survival` package (`vignette("timedep", package = "survival")`).
Besides information on survival, the data set contains the
Karnofsky performance scores `karno` (the higher the better), `age` and whether
 `prior` therapy occurred, along with some additional covariates, see `help("veteran", package = "survival")` for details:

```{r}
# for some reason the prior variable is coded 0/10 instead of 0/1
veteran <- survival::veteran
veteran <- veteran %>%
  mutate(
    trt   = 1L * (trt == 2),
    prior = 1L * (prior == 10)) %>%
  filter(time < 400) # restriction for illustration
head(veteran)
```


### Extended Cox Model (with known shape of time-variation function)

To fit a time-varying effect of `karno` the authors suggest to use the function
$$
f(x_{\text{karno}},t) = \beta_{\text{karno}}\cdot x_{\text{karno}} +
  \beta_{\text{karno},t} \cdot x_{\text{karno}} \cdot \log(t+20).
$$
This is an instance of the "known time-variation function" case above with
$g(t) = \log(t+20).$

```{r}
vfit <- coxph(
  formula = Surv(time, status) ~ trt + prior + karno + tt(karno),
  data    = veteran,
  tt      = function(x, t, ...) x * log(t + 20))
coef(vfit)
```

```{r, echo=FALSE}
ttcoef <- round(coef(vfit), 3)[3:4]
```

Thus the time-varying component of the effect becomes
$\beta_{\text{karno}}+\beta_{\text{karno},t}\cdot\log(t+20) = `r ttcoef[1]` + `r ttcoef[2]`\cdot\log(t+20)$:

```{r}
t <- seq(0, 400, by = 10)
plot(x = t, y = coef(vfit)["karno"] + coef(vfit)["tt(karno)"] * log(t + 20),
  type = "l", ylab = "Beta(t) for karno", las = 1, ylim = c(-.1, .05),
  col = Set1[1])
```

### PAM (with known shape of time-variation function)

To fit a PAM with equivalent model specification (except for the baseline hazard)
we can use

```{r}
# data transformation
ped <- veteran %>% as_ped(Surv(time, status)~., id = "id") %>%
    mutate(logt20 = log(tstart + (tstart - tend) / 2 + 20))
head(ped) %>% select(interval, ped_status, trt, karno, age, prior, logt20)

# fit model
pam <- gam(ped_status ~ s(tend) + trt + prior + karno + karno:logt20,
    data = ped, offset = offset, family = poisson())
cbind(
  pam = coef(pam)[2:5],
  cox = coef(vfit))
# compare fits
plot(x = t, y = coef(vfit)["karno"] + coef(vfit)["tt(karno)"] * log(t + 20),
  type = "l", ylab = "Beta(t) for karno", ylim = c(-.1, .05), las = 1,
  col = Set1[1])
t_pem <- int_info(ped)$tend
lines(x = t_pem, y = coef(pam)["karno"] + coef(pam)["karno:logt20"] * log(t_pem + 20),
  col = Set1[2], type = "s")
```
Both methods yield very similar estimates of the time-varying effect of the Karnofsky-Score, with a reduced hazard for higher-scoring patients at the beginning of the follow-up that diminishes over time and turns into an increased hazard for higher-scoring patients after about day 150.

### PAM (penalized estimation of time-variation function)

In case we don’t want to pre-specify which shape the time-dependency should have,
we can specify the effect of `karno` as
$f(x_{\text{karno}},t) = f(t)\cdot x_{\text{karno}}$, where $f(t)$
is estimated from the data:

```{r, pam_tv_karno}
# no need to specify main effect for karno here
pam2 <- gam(ped_status ~ s(tend) + trt + prior + s(tend, by = karno),
    data = ped, offset = offset, family = poisson())
```

The figure below visualizes the estimate of the time-varying effect of `karno`
for all three models
<details>
  <summary>*Expand here to see R-Code*</summary>

```{r gg_tv_karno, dependson="pam_tv_karno"}
term.df <- ped %>% ped_info() %>% add_term(pam2, term = "karno") %>%
    mutate_at(c("fit", "ci_lower", "ci_upper"), funs(. / .data$karno)) %>%
    mutate(
        cox.fit = coef(vfit)["karno"] + coef(vfit)["tt(karno)"] * log(tend + 20),
        pam.fit = coef(pam)["karno"] + coef(pam)["karno:logt20"] * log(tend + 20))
gg_tv_karno <- ggplot(term.df, aes(x = tend, y = fit)) +
    geom_step(aes(col = "PAM with penalized spline")) +
    geom_stepribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.2) +
    geom_line(aes(y = cox.fit, col = "Cox with log-transform")) +
    geom_step(aes(y = pam.fit, col = "PAM with log-transform")) +
    scale_color_manual(name = "Method", values = c(Set1[1:2], "black")) +
    xlab("t") + ylab(expression(hat(f)(t)))
```

</details>

```{r, dependson = "gg_tv_karno", echo = FALSE, fig.width = 6.5}
gg_tv_karno
```

The semi-parametric PAM model estimate for $f(t)$ increases fairly linearly up to day 150 and flattens out at about 0 (i.e., no effect of Karnofsky-Scores on the hazard) afterwards.


### PAM with smooth, smoothly time-varying effect of the Karnofsky Score

To fit a non-linear, non-linearly time-varying effect we can specify a
two-dimensional interaction between the covariate of interest (here
the Karnofsky-Score and a variable that represents time in the respective
interval, e.g. interval end-points) using tensor product terms.

In `mgcv::gam` such two-dimensional effects can be directly used
either via `te` or `ti` terms in the model specification. The later is
especially useful for disentangling the marginal (time-constant) and interaction
(time-varying) effects of the respective covariate.

Below we first fit a model using the `te` specification. Note that we did not
include a `s(tend)` term here, as the time-variable `tend` is already present
in the `te` term, thus the effect `te(tend, karno)` also includes the *shape* of the baseline hazard
as well. The *level* of the baseline log hazard is given by the intercept of the model.

```{r}
# Non-linear, non-linearly time-varying effects
pam3 <- gam(
  formula = ped_status ~ trt + prior + s(age) + te(tend, karno),
  data   = ped,
  family = poisson(),
  offset = offset)
```

The summary of the model indicates that the estimated bivariate function
$\hat{f}(x_{\text{karno}}, t)$ is highly non-linear ($edf \approx 8.9$):

```{r}
summary(pam3)
```

The 3D perspective plot can aid interpretation, where y- and x-axes depict
the Karnofsky-Score and the time respectively and the z-axis displays the
contribution of the effect to the log-hazard for each combination of
$x_{\text{karno}}$ and $t$.^[Note that the graphical representation in the 3D wireframe plot as well as the heatmap/contour plots below are not exact -- these effects are actually step functions over time, with steps at the interval end points `tend`, since a PAM implies that *all* time-varying effects are piece-wise constant over the intervals used for the fit. In practice, this subtle difference can be neglected if the intervals are small enough, as in this case.]

```{r, fig.width=6, fig.height=6}
plot(pam3, select = 3, scheme = 1, theta = 120, ticktype = "detailed")
```
Such 3D plots are sometimes difficult to interpret, thus we also provide
a heat-/contourplot (left panel) with respective slices for fixed
values of the Karnofsky-Score (middle panel) and fixed time-points/intervals
(right panel) below.

The left panel depicts the Karnofsky-Score on the y-axis and the
time on the x-axis. The value of $\hat{f}(x_{\text{karno}}, t)$ is visualized using
a color gradient, where blue colors indicate log-hazard decrease and
red colors a log-hazard increase. The grayed out areas depict combinations
of `karno` and `tend` that were not present in the data.
Dotted horizontal and vertical lines indicate slices that are displayed
in the middle and right panel.
For fixed $t=1$, we obtain the effect of the Karnofsky-Score on the log-hazard
at the beginning of the follow-up (see also right panel for $t=1$),
which decreases strongly from low to high values of $x_{\text{karno}}$.

Holding the Karnofsky-Score constant, we can see how the log hazard
changes over time for different $x_{\text{karno}}$ (middle panel).
For larger values ($x_{\text{karno}} \in \{75, 90\}$) the log-hazard is smaller
at the beginning and increases over the course of the follow-up, while for
small values ($x_{\text{karno}} \in \{40\}$) the log-hazard is positive and decreases
toward later time points. This could indicate that the effect of
the Karnofsky-Score tends towards 0 over time as the information collected
at the beginning of the follow-up becomes outdated (but see uncertainty).

<details>
  <summary>*Expand here to see R-Code*</summary>

```{r warning=FALSE, fig.width=7, fig.height=3}
# heat map/contour plot
te_gg <- gg_tensor(pam3) +
  geom_vline(xintercept = c(1, 51, 200), lty = 3) +
  geom_hline(yintercept = c(40, 75, 95), lty = 3) +
  scale_fill_gradient2(
    name = expression(hat(f)(list(x[plain(karno)], t))),
    low  = "steelblue", high = "firebrick2") +
  geom_contour(col = "grey30") +
  xlab("t") + ylab(expression(x[plain(karno)])) +
  theme(legend.position  = "bottom")

# plot f(karno, t) for specific slices
karno_df <- ped %>%
  make_newdata(tend = unique(tend), karno = c(40, 75, 95)) %>%
  add_term(pam3, term = "karno")

# shortcut
# gg_slice(ped, pam3, "karno", tend = unique(tend), karno = c(40, 75, 95))
karno_gg <- ggplot(karno_df, aes(x = tend, y = fit)) +
  geom_step(aes(col = factor(karno)), lwd = 1.1) +
  geom_stepribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = factor(karno)),
    alpha = .2) +
  scale_color_manual(
    name   = expression(x[plain(karno)]),
    values = Greens[c(4, 7, 9)]) +
  scale_fill_manual(
    name   = expression(x[plain(karno)]),
    values = Greens[c(4, 7, 9)]) +
  ylab(expression(hat(f)(list(x[plain(karno)], t)))) +
  xlab("t") +  coord_cartesian(ylim = c(-4, 3)) +
  theme(legend.position  = "bottom")

time_df <- ped %>%
  make_newdata(tend = c(1, 51, 200), karno = seq(20, 100, by = 5)) %>%
  add_term(pam3, term = "karno")

time_gg <- ggplot(time_df, aes(x = karno)) +
  geom_line(aes(y = fit, col = factor(tend)), lwd = 1.1) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = factor(tend)),
    alpha = .2) +
  scale_color_manual(name = "t", values = Purples[c(4, 6, 8)]) +
  scale_fill_manual(name = "t", values = Purples[c(4, 6, 8)]) +
  ylab(expression(hat(f)(list(x[plain(karno)], t)))) +
  xlab(expression(x[plain(karno)])) + coord_cartesian(ylim = c(-4, 3)) +
  theme(legend.position  = "bottom")
```
</details>

```{r, fig.width = 7, fig.height = 3, warning = FALSE, echo = FALSE}
gridExtra::grid.arrange(te_gg, karno_gg, time_gg, nrow = 1)
```


The following figure shows the estimated effect (middle panel) along with a
pointwise upper (right) and lower (left) CI. Note that we have to be somewhat
cautious with interpretation, considering the large uncertainty of the effect
estimate, especially for lower Karnofsky-Scores and later time-points.
Also note that the estimate does not include the estimated average
time-constant log-hazard
(`coefficients(pam3)["(Intercept)"]=``r round(coefficients(pam3)["(Intercept)"], 3)`)
and its uncertainty.

```{r, fig.width = 7, fig.height = 3, warning = FALSE}
gg_tensor(pam3, ci = TRUE) +
  xlab("t") + ylab(expression(x[plain(karno)]))
```
