---
title: "Time-varying effects"
author: "Andreas Bender"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tveffects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  fig.align = "center",
  fig.width = 4, 
  fig.height = 4
)
```

```{r, message = FALSE}
library(ggplot2)
theme_set(theme_bw())
library(magrittr)
library(tidyr)
library(purrr)
library(survival)
library(mgcv)
library(dplyr)
library(pam)
```

In this vignette we show examples of how to fit time-varying effects of 
time-constant continuous covariates. Time-varying effects of time-constant 
categorical variables can be viewed in the context of stratified proportional 
hazards models, where all observations from one group of the categorical variable
have an individual baseline hazard. This case is detailed in `vignette("strata", pacakge="pam")`.
<!-- (see vignette on stratified baseline for details) WHICH PACKAGE IS THIS? -->.

## Possible specifications of time-variation

In the following we denote the time-constant covariate with $x$ and time with $t$. 
Essentially, a time-varying effect of $x$ then can be specified as an interaction 
term between $x$ and $t$, where different levels of complexity for this interaction 
are possible:

a) $\beta_x\cdot x+\beta_{x:t}\cdot(x\cdot g(t))$: Linear effect of $x$ with
time-variation of $g(t)$, where $g()$ is a known or pre-specified transformation of 
time $t$, e.g. the $\log$-function.

b) $f_x(x)\cdot g(t)$: Non-linear effect of $x$, (linearly) time-varying with $g(t)$,
where $g(t)$ is a known or pre-specified transformation of time

c) $f_t(t)\cdot x$: A Varying coefficient model of $x$, where time-variation is 
non-linear (if $x$ is a dummy coding for levels of a categorical variable this constitutes a stratified model with a different "baseline" hazard for each category)

d) $f_{x,t}(x,t)$: A non-linear, non-linearly time-varying effect of $x$

## Veteran Data example

For illustration (and comparison) we use the `veteran` data presented in the 
vignette of the survival package (`vignette("timedep", package = "survival")`). 
Besides information on survival, the data set contains information on the 
Karnofsky performance score `karno` (the higher the better), `age` and whether
`prior`therapy occurred:
.
```{r}
# for some reason the prior variable is coded 0/10 instead of 0/1
data("veteran", package = "survival")
# we limit the data to event times < 400 (not many events afterwards)
veteran %<>% mutate(prior = ((prior != 0) * 1)) %>% filter(time < 400)
head(veteran)
```

To fit a time-varying effect of `karno` the authors suggest to use the function 
$f(x,t) = \beta_{\text{karno}}(x_{\text{karno}})+\beta_{\text{karno},t}\cdot (x_{\text{karno}}\cdot \log(t+20))$. 
This is an instance of case a) above with $g(t) = log(t+20)$.

```{r}
vfit <- coxph(Surv(time, status) ~ trt + prior + karno + tt(karno), 
    data = veteran, 
    tt = function(x, t, ...) x * log(t + 20))
coef(vfit)
```
```{r, echo=FALSE}
ttcoef <- round(coef(vfit), 3)[3:4]
```

Thus the time-varying component of the effect becomes 
$\beta_{\text{karno}}+\beta_{\text{karno},t}\log(t+20) = `r ttcoef[1]` + `r ttcoef[2]`\cdot\log(t+20)$:

```{r}
t <- seq(0, 400, by = 10)
plot(x = t, y = coef(vfit)["karno"] + coef(vfit)["tt(karno)"]*log(t + 20), 
  type = "l", ylab = "Beta(t) for karno", las = 1)
```

To fit a PAM with equivalent model specification (except for the baseline hazard) 
we can use

```{r}
# data transformation
ped <- split_data(Surv(time, status)~., data = veteran, id = "id") %>% 
    mutate(logt20 = log(tstart+(tstart-tend)/2 + 20))
head(ped) %>% select(interval, ped_status, trt, karno, age, prior, logt20)

# fit model 
pam <- gam(ped_status ~ s(tend) + trt + prior + karno + karno:logt20, 
    data = ped, offset = offset, family = poisson())
coef(pam)[1:5]
# compare fits 
plot(x = t, y = coef(vfit)["karno"] + coef(vfit)["tt(karno)"]*log(t + 20), 
  type = "l", ylab = "Beta(t) for karno", ylim = c(-.1, .05), las = 1) 
lines(x = t, y = coef(pam)["karno"] + coef(pam)["karno:logt20"]*log(t + 20), col = 2)
```

In case we don’t want to pre-specify which shape the time-dependency should have, 
we can specify the effect of `karno` as $f(x_{\text{karno}},t) = f(t)\cdot x_{\text{karno}}$, where $f(t)$
is estimated from the data:
```{r, fig.width = 6}
# no need to specify main effect for karno here 
pam2 <- gam(ped_status ~ s(tend) + trt + prior + s(tend, by = karno), 
    data = ped, offset = offset, family = poisson())
term.df <- ped %>% ped_info() %>% add_term(pam2, term = "karno") %>% 
    mutate_at(c("fit", "low", "high"), function(z) z/.$karno) %>% 
    mutate(
        cox.fit = coef(vfit)["karno"] + coef(vfit)["tt(karno)"]*log(tend + 20), 
        pam.fit = coef(pam)["karno"] + coef(pam)["karno:logt20"]*log(tend + 20)) 
ggplot(term.df, aes(x = tend, y = fit)) + 
    geom_line() + 
    geom_ribbon(aes(ymin = low, ymax = high), alpha = 0.2) + 
    geom_line(aes(y = cox.fit, col = "Cox with log-transform")) + 
    geom_line(aes(y = pam.fit, col = "PAM with log-transform")) + 
    scale_color_brewer(name="Method", palette = "Set1") +
    xlab("t") + ylab("Beta(t) for karno")
```

The semi-parametric PAM model estimate flattens out after ca. 150 days.
