---
title: "Time-varying effects"
author: "Andreas Bender"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tveffects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  fig.align = "center",
  fig.width = 4, 
  fig.height = 4,
  crop=TRUE)
```

```{r, message = FALSE}
library(ggplot2)
theme_set(theme_bw())
library(gridExtra)
library(magrittr)
library(tidyr)
library(dplyr)
library(pam)
library(survival)
library(mgcv)
```

In this vignette we show examples of how to fit time-varying effects of 
time-constant **continuous** covariates. Note that time-varying effects of time-constant 
**categorical** variables are analogous to stratified proportional 
hazards models, where observations from different levels of the categorical 
variable have different baseline hazards. That setting is described in
[`vignette("strata", package="pam")`](strata.html).

## Possible specifications of time-variation

In the following we denote the continuous time-constant covariate with $x$ and time with $t$. 
A time-varying effect of $x$ can then be specified as an interaction 
term between $x$ and $t$, where different levels of complexity and flexibility for this interaction 
are possible:

a) $\beta_x\cdot x+\beta_{x:t}\cdot(x\cdot g(t))$: Linear effect of $x$ with
time-variation given by $g(t)$, where $g(\cdot)$ is a known or pre-specified transformation of 
time $t$, e.g. the $\log$-function.

b) $f_x(x)\cdot g(t)$: Non-linear effect of $x$, (linearly) time-varying with $g(t)$,
where $g(t)$ is a known or pre-specified transformation of time

c) $f_t(t)\cdot x$: A *varying coefficient model* in $x$, where time-variation is 
non-linear and estimated from the data. If $x$ is a dummy variable coding for levels of a categorical variable this constitutes a stratified model with a different "baseline" hazard for each category, see [`vignette("strata", package="pam")`](strata.html).

d) $f_{x,t}(x,t)$: A non-linear effect of $x$ that varies non-linearly over time $t$.

## Veteran Data example

For illustration and comparison we use the `veteran` data presented in the 
vignette of the `survival` package (`vignette("timedep", package = "survival")`). 
Besides information on survival, the data set contains the 
Karnofsky performance scores `karno` (the higher the better), `age` and whether
 `prior` therapy occurred, along with some additional covariates, see `help("veteran", package = "survival")` for details:

```{r}
# for some reason the prior variable is coded 0/10 instead of 0/1
data("veteran", package = "survival")
# we limit the data to event times < 400 (not many events afterwards)
veteran %<>% mutate(prior = ((prior != 0) * 1)) %>% filter(time < 400)
head(veteran)
```


### Extended Cox Model (with known shape of time-variation function)

To fit a time-varying effect of `karno` the authors suggest to use the function 
$$
f(x_{\text{karno}},t) = \beta_{\text{karno}}\cdot x_{\text{karno}} +
  \beta_{\text{karno},t} \cdot x_{\text{karno}} \cdot \log(t+20).
$$ 
This is an instance of case a) above with $g(t) = \log(t+20).$

```{r}
vfit <- coxph(
  formula = Surv(time, status) ~ trt + prior + karno + tt(karno),
  data    = veteran,
  tt      = function(x, t, ...) x * log(t + 20))
coef(vfit)
```

```{r, echo=FALSE}
ttcoef <- round(coef(vfit), 3)[3:4]
```

Thus the time-varying component of the effect becomes 
$\beta_{\text{karno}}+\beta_{\text{karno},t}\cdot\log(t+20) = `r ttcoef[1]` + `r ttcoef[2]`\cdot\log(t+20)$:

```{r}
t <- seq(0, 400, by = 10)
plot(x = t, y = coef(vfit)["karno"] + coef(vfit)["tt(karno)"]*log(t + 20), 
  type = "l", ylab = "Beta(t) for karno", las = 1, ylim=c(-.1, .05))
```

### PAM (with known shape of time-variation function)

To fit a PAM with equivalent model specification (except for the baseline hazard) 
we can use

```{r}
# data transformation
ped <- split_data(Surv(time, status)~., data = veteran, id = "id") %>% 
    mutate(logt20 = log(tstart+(tstart-tend)/2 + 20))
head(ped) %>% select(interval, ped_status, trt, karno, age, prior, logt20)

# fit model 
pam <- gam(ped_status ~ s(tend) + trt + prior + karno + karno:logt20, 
    data = ped, offset = offset, family = poisson())
cbind(
  pam = coef(pam)[2:5],
  cox = coef(vfit))
# compare fits 
plot(x = t, y = coef(vfit)["karno"] + coef(vfit)["tt(karno)"]*log(t + 20), 
  type = "l", ylab = "Beta(t) for karno", ylim = c(-.1, .05), las = 1) 
lines(x = t, y = coef(pam)["karno"] + coef(pam)["karno:logt20"]*log(t + 20), col = 2)
```
Both methods yield very similar estimates of the time-varying effect of the Karnofsky-Score, with a reduced hazard for higher-scoring patients at the beginning of the follow-up that diminishes over time and turns into an increased hazard for higher-scoring patients after about day 150.

### PAM (penalized estimation of time-variation function)

In case we don’t want to pre-specify which shape the time-dependency should have, 
we can specify the effect of `karno` as 
$f(x_{\text{karno}},t) = f(t)\cdot x_{\text{karno}}$, where $f(t)$
is estimated from the data:

```{r, fig.width = 6}
# no need to specify main effect for karno here 
pam2 <- gam(ped_status ~ s(tend) + trt + prior + s(tend, by = karno), 
    data = ped, offset = offset, family = poisson())
term.df <- ped %>% ped_info() %>% add_term(pam2, term = "karno") %>% 
    mutate_at(c("fit", "low", "high"), function(z) z/.$karno) %>% 
    mutate(
        cox.fit = coef(vfit)["karno"] + coef(vfit)["tt(karno)"]*log(tend + 20), 
        pam.fit = coef(pam)["karno"] + coef(pam)["karno:logt20"]*log(tend + 20)) 
ggplot(term.df, aes(x = tend, y = fit)) + 
    geom_line() + 
    geom_ribbon(aes(ymin = low, ymax = high), alpha = 0.2) + 
    geom_line(aes(y = cox.fit, col = "Cox with log-transform")) + 
    geom_line(aes(y = pam.fit, col = "PAM with log-transform")) + 
    scale_color_brewer(name="Method", palette = "Set1") +
    xlab("t") + ylab("Beta(t) for karno")
```

The semi-parametric PAM model estimate for $f(t)$ increases fairly linearly up to day 150 and flattens out at about 0 (i.e., no effect of Karnofsky scores on the hazard) afterwards.


### PAM with smooth, smoothly time-varying effect of the Karnosky Score

To fit a non-linear, non-linearly time-varying effect we can specify a 
two-dimensional interaction between the covariate of interest (here 
the Karnofsky-Score and a variable that represents time in the respective 
interval, e.g. interval end-points) using tensor product terms. 

In `mgcv::gam` such two-dimensional effects can be directly used 
either via `te` or `ti` terms in the model specification. The later is 
especially useful for disentangling the marginal (time-constant) and interaction 
(time-varying) effects of the respective covariate. 

Below we first fit a model using the `te` specification. Note that we did not 
include a `s(tend)` term here, as the time-variable `tend` is already present
in the `te` term, thus the effect `te(tend, karno)` also includes the *shape* of the baseline hazard
as well. The *level* of the baseline log hazard is given by the intercept of the model.

```{r}
## Non-linear, non-linearly time-varying effects
pam3 <- gam(
  formula = ped_status ~ trt + prior + s(age) + te(tend, karno),
  data   = ped,
  family = poisson(),
  offset = offset)
```

The summary of the model indicates that the estimated bivariate function 
$\hat{f}(x_{\text{karno}}, t)$ is highly non-linear ($edf \approx 8.9$):

```{r}
summary(pam3)
```

The 3D perspective plot can aid interpretation, where y- and x-axes depict 
the Karnosky-Score and the time respectively and the z-axis displays the 
contribution of the effect to the log-hazard for each combination of 
$x_{\text{karno}}$ and $t$.

```{r, fig.width=6, fig.height=6}
plot(pam3, select = 2, pers = T, theta = 150, ticktype = "detailed")
```
Such 3D plots are sometimes difficult to interpret, thus we also provide 
a heat-/contourplot (left panel) with respective slices for fixed 
values of the Karnofsky-Score (middle panel) and fixed time-points/intervals 
(right panel) below. 

The left panel again depicts the Karnosky Score on the y-axis and the 
time on the x-axis. The value of $\hat{f}(x_{\text{karno}}, t)$ is visualized using 
a color gradient, where blue colors indicate log-hazard decrease and 
red colors a log-hazard increase. The grayed out areas depict combinations
of \code{karno} and \code{tend} that were not present in the data. 
Dotted horizontal and vertical lines indicate slices that are displayed 
in the middle and right panel. 

For fixed $t=0$, we obtain the effect of the Karnofsky-Score on the log-hazard 
at the beginning of the follow-up (red line left panel), which decreases strongly from low 
to high values of $x_{\text{karno}}$. 
<!-- This result is in line with the estimation 
one would obtain fitting a time-constant, non-linear effect of 
$f(x_{\text{karno}})$. ABER DAS IST ZUFALL HIER; DAS MUSS JA NICHT SO SEIN...  -->

Holding the Karnofsky Score constant, we can see how the log hazard
changes over time for different given values of $x_{\text{karno}}$ (middle panel). 
For higher values $x_{\text{karno}} >60$ the log-hazard is smaller
at the beginning and increases over the course of the follow-up 
(orange and blue lines middle panel). This 
is consistent with the time-varying effect of the Karnofsky-Score 
estimated in the previous section. However, the bivariate function indicates
that the log hazard may have a parabolic shape for low $x_{\text{karno}}$ values, 
decreasing at the beginning and increasing again for later 
time-points (green line middle panel). 

<details>
  <summary>Expand here for detailed R-Code</summary>
```{r}
te_plot <- gg_tensor(pam3) +
  geom_vline(xintercept=c(0, 50, 200, 300), lty=3) +
  geom_hline(yintercept=c(40, 75, 95), lty=3) +
  scale_fill_gradient2(
    name=expression(hat(f)(list(x[plain(karno)],t))),
    low="steelblue", high="firebrick2") +
  geom_contour(col="grey30") +
  xlab("t") + ylab(expression(x[plain(karno)])) +
  theme(legend.position="bottom")

## plot f(karno, t) for specific slices
df.karno <- combine_df(
  int_info(ped),
  select(sample_info(ped), -karno),
  data.frame(karno = c(40, 75, 95)))

df.karno$fte <- predict(pam3, newdata=df.karno, type="terms")[,"te(tend,karno)"]
karno_plot <- ggplot(df.karno, aes(x=tend, y=fte)) +
  geom_line(aes(col=factor(karno))) +
  scale_color_brewer(
    name=expression(x[plain(karno)]),
    palette = "Set2") +
  ylab(expression(f(list(x[plain(karno)],t)))) +
  xlab("t")+
  theme(legend.position="bottom")

df.time  <- combine_df(
  data.frame(sample_info(ped)%>% select(-karno)),
#  data.frame(karno = seq_range(c(20, 40, 75, 95))),
  data.frame(karno = seq(20, 100, by=5)),
  data.frame(tend  = c(0, 50, 200, 300)))
df.time$fte <- predict(pam3, newdata=df.time, type="terms")[,"te(tend,karno)"]
time_plot <- ggplot(df.time, aes(x=karno, y=fte)) +
  geom_line(aes(col=factor(tend))) +
  scale_color_brewer(name="t", palette="Set1") +
  ylab(expression(f(list(x[plain(karno)],t)))) +
  xlab(expression(x[plain(karno)])) +
  theme(legend.position="bottom")
```
</details>

```{r, fig.width=7, fig.height=3, warning=FALSE}
grid.arrange(te_plot, karno_plot, time_plot, nrow=1)
```

The following figure shows the estimated effect (middle panel) along with a pointwise upper (right) and lower (left) CI. Note that we have to be somewhat cautious with interpretation, considering the large uncertainty of the effect estimate, especially for lower Karnofsky-Scores
and later time-points. Also note that the estimate does not include the estimated average 
time-constant log-hazard (`coefficients(pam3)["(Intercept)"]=``r round(coefficients(pam3)["(Intercept)"], 3)`) and its uncertainty: 

```{r, fig.width=7, fig.height=3, warning=FALSE}
gg_tensor(pam3, ci=TRUE) + 
  xlab("t") + ylab(expression(x[plain(karno)]))
```

