---
title: "Time-varying effects"
author: "Andreas Bender"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tveffects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  fig.align = "center",
  fig.width = 4, 
  fig.height = 4,
  crop=TRUE)
```

```{r, message = FALSE}
library(ggplot2)
theme_set(theme_bw())
library(gridExtra)
library(magrittr)
library(tidyr)
library(dplyr)
library(pam)
library(survival)
library(mgcv)
```

In this vignette we show examples of how to fit time-varying effects of 
time-constant continuous covariates. Time-varying effects of time-constant 
categorical variables can be viewed in the context of stratified proportional 
hazards models, where all observations from one group of the categorical 
variable have an individual baseline hazard. This case is detailed in
`vignette("strata", pacakge="pam")`.

## Possible specifications of time-variation

In the following we denote the time-constant covariate with $x$ and time with $t$. 
Essentially, a time-varying effect of $x$ then can be specified as an interaction 
term between $x$ and $t$, where different levels of complexity for this interaction 
are possible:

a) $\beta_x\cdot x+\beta_{x:t}\cdot(x\cdot g(t))$: Linear effect of $x$ with
time-variation of $g(t)$, where $g()$ is a known or pre-specified transformation of 
time $t$, e.g. the $\log$-function.

b) $f_x(x)\cdot g(t)$: Non-linear effect of $x$, (linearly) time-varying with $g(t)$,
where $g(t)$ is a known or pre-specified transformation of time

c) $f_t(t)\cdot x$: A Varying coefficient model of $x$, where time-variation is 
non-linear (if $x$ is a dummy coding for levels of a categorical variable this constitutes a stratified model with a different "baseline" hazard for each category)

d) $f_{x,t}(x,t)$: A non-linear, non-linearly time-varying effect of $x$

## Veteran Data example

For illustration (and comparison) we use the `veteran` data presented in the 
vignette of the survival package (`vignette("timedep", package = "survival")`). 
Besides information on survival, the data set contains information on the 
Karnofsky performance score `karno` (the higher the better), `age` and whether
`prior`therapy occurred:
.
```{r}
# for some reason the prior variable is coded 0/10 instead of 0/1
data("veteran", package = "survival")
# we limit the data to event times < 400 (not many events afterwards)
veteran %<>% mutate(prior = ((prior != 0) * 1)) %>% filter(time < 400)
head(veteran)
```


### Extended Cox Model (with known shape of time-variation function)

To fit a time-varying effect of `karno` the authors suggest to use the function 
$$
f(x_{text{karno}},t) = \beta_{\text{karno}}(x_{\text{karno}}) +
  \beta_{\text{karno},t} \cdot (x_{\text{karno}} \cdot \log(t+20))
$$. 
This is an instance of case a) above with $g(t) = log(t+20)$.

```{r}
vfit <- coxph(
  formula = Surv(time, status) ~ trt + prior + karno + tt(karno),
  data    = veteran,
  tt      = function(x, t, ...) x * log(t + 20))
coef(vfit)
```

```{r, echo=FALSE}
ttcoef <- round(coef(vfit), 3)[3:4]
```

Thus the time-varying component of the effect becomes 
$\beta_{\text{karno}}+\beta_{\text{karno},t}\log(t+20) = `r ttcoef[1]` + `r ttcoef[2]`\cdot\log(t+20)$:

```{r}
t <- seq(0, 400, by = 10)
plot(x = t, y = coef(vfit)["karno"] + coef(vfit)["tt(karno)"]*log(t + 20), 
  type = "l", ylab = "Beta(t) for karno", las = 1, ylim=c(-.1, .05))
```

### PAM (with known shape of time-variation function)

To fit a PAM with equivalent model specification (except for the baseline hazard) 
we can use

```{r}
# data transformation
ped <- split_data(Surv(time, status)~., data = veteran, id = "id") %>% 
    mutate(logt20 = log(tstart+(tstart-tend)/2 + 20))
head(ped) %>% select(interval, ped_status, trt, karno, age, prior, logt20)

# fit model 
pam <- gam(ped_status ~ s(tend) + trt + prior + karno + karno:logt20, 
    data = ped, offset = offset, family = poisson())
cbind(
  pam = coef(pam)[2:5],
  cox = coef(vfit))
# compare fits 
plot(x = t, y = coef(vfit)["karno"] + coef(vfit)["tt(karno)"]*log(t + 20), 
  type = "l", ylab = "Beta(t) for karno", ylim = c(-.1, .05), las = 1) 
lines(x = t, y = coef(pam)["karno"] + coef(pam)["karno:logt20"]*log(t + 20), col = 2)
```

### PAM (penalized estimation of time-variation function)

In case we don’t want to pre-specify which shape the time-dependency should have, 
we can specify the effect of `karno` as 
$f(x_{\text{karno}},t) = f(t)\cdot x_{\text{karno}}$, where $f(t)$
is estimated from the data:

```{r, fig.width = 6}
# no need to specify main effect for karno here 
pam2 <- gam(ped_status ~ s(tend) + trt + prior + s(tend, by = karno), 
    data = ped, offset = offset, family = poisson())
term.df <- ped %>% ped_info() %>% add_term(pam2, term = "karno") %>% 
    mutate_at(c("fit", "low", "high"), function(z) z/.$karno) %>% 
    mutate(
        cox.fit = coef(vfit)["karno"] + coef(vfit)["tt(karno)"]*log(tend + 20), 
        pam.fit = coef(pam)["karno"] + coef(pam)["karno:logt20"]*log(tend + 20)) 
ggplot(term.df, aes(x = tend, y = fit)) + 
    geom_line() + 
    geom_ribbon(aes(ymin = low, ymax = high), alpha = 0.2) + 
    geom_line(aes(y = cox.fit, col = "Cox with log-transform")) + 
    geom_line(aes(y = pam.fit, col = "PAM with log-transform")) + 
    scale_color_brewer(name="Method", palette = "Set1") +
    xlab("t") + ylab("Beta(t) for karno")
```

The semi-parametric PAM model estimate flattens out after ca. 150 days.


### PAM with smooth, smoothly time-varying effect of the Karnosky Score

To fit a non-linear, non-linearly time-varying effect we can specify a 
two-dimensional interaction between the covariate of interest (here 
the Karnofsky-Score and a variable that represents time in the respective 
interval, e.g. interval end-points) using tensor product terms. 

In `mgcv::gam` such two-dimensional effects can be directly used 
either via `te` or `ti` terms in the model specification. The later is 
especially useful, when we want to disentangle the marginal and interaction 
effects of the respective covariates. 

Below we first fit a model using the `te` specification. Note that we did not 
include a `s(tend)` term here, as the time-variable `tend` is already present
in the `te` term, thus we could still obtain a "baseline hazard" by setting 
all other covariates to 0. 

```{r}
## Non-linear, non-linearly time-varying effects
pam3 <- gam(
  formula = ped_status ~ trt + prior + s(age) + te(tend, karno),
  data   = ped,
  family = poisson(),
  offset = offset)
```

The summary of the model indicates that the estimated bivariate function 
$\hat{f}(x_{karno}, t)$ is highly non-linear (edf \approx 8.9):
```{r}
summary(pam3)
```

The 3D perspective plot can aid interpretation, where y- and x-axes depict 
the Karnosky-Score and the time respectively and the z-axis displays the 
contribution of the effect to the log-hazard for each combination of 
$x_{karno}$ and $t$.
```{r, fig.width=6, fig.height=6}
plot(pam3, select=2, pers=T, theta=150, ticktype="detailed")
```
Such 3D plots are sometimes difficult to interpret, thus we also provide 
a heat-/contourplot (left panel) with respective slices for fixed 
values of the Karnofsky-Score (middle panel) and fixed time-points/intervals 
(right panel) below. 

The left panel again depicts the Karnosky Score on the y-axis and the 
time on the x-axis. The value of $\hat{f}(x_{karno}, t)$ is visualized using 
a color gradient, where blue colors indicate log-hazard decrease and 
red colors a log-hazard increase. The grayed out areas depict combinations
of \code{karno} and \code{tend} that were not present in the data. 
Dotted horizontal and vertical lines indicate slices that are displayed 
in the middle and right panel. 

Fixing $t=0$, we obtain the effect of the Karnofsky-Score at the beginning 
of the follow-up (red line left panel), which decreases strongly from low 
to high values of $x_{karno}$. This result is in line with the estimation 
one would obtain fitting a time-constant, non-linear effect of the 
$f(x_{karno}).

Fixing the Karnofsky Score, we can see how the effect of the Karnofsky-Score
changes over time for different values of $x_{karno}$ (middle panel). 
For higher values $x_{karno} >60$ the effect is negative (decreased log-hazard)
at the beginning and increases over the course of the follow-up 
(orange and blue lines middle panel). This 
again is consistent with the time-varying effect of the Karnofsky-Score 
estimated in the previous section. However, the bivariate function indicates
that for lower $x_{karno}$ values the effect may have a parabolic shape, 
decreasing log-hazard at the beginning and increasing again for later 
time-points (green line middle panel). 

<details>
  <summary>Expand here for detailed R-Code</summary>
```{r}
te_plot <- gg_tensor(pam3) +
  geom_vline(xintercept=c(0, 50, 200, 300), lty=3) +
  geom_hline(yintercept=c(40, 75, 95), lty=3) +
  scale_fill_gradient2(
    name=expression(hat(f)(list(x[plain(karno)],t))),
    low="steelblue", high="firebrick2") +
  geom_contour(col="grey30") +
  xlab("t") + ylab(expression(x[plain(karno)])) +
  theme(legend.position="bottom")

## plot f(karno, t) for specific slices
df.karno <- combine_df(
  int_info(ped),
  select(sample_info(ped), -karno),
  data.frame(karno = c(40, 75, 95)))

df.karno$fte <- predict(pam3, newdata=df.karno, type="terms")[,"te(tend,karno)"]
karno_plot <- ggplot(df.karno, aes(x=tend, y=fte)) +
  geom_line(aes(col=factor(karno))) +
  scale_color_brewer(
    name=expression(x[plain(karno)]),
    palette = "Set2") +
  ylab(expression(f(list(x[plain(karno)],t)))) +
  xlab("t")+
  theme(legend.position="bottom")

df.time  <- combine_df(
  data.frame(sample_info(ped)%>% select(-karno)),
#  data.frame(karno = seq_range(c(20, 40, 75, 95))),
  data.frame(karno = seq(20, 100, by=5)),
  data.frame(tend  = c(0, 50, 200, 300)))
df.time$fte <- predict(pam3, newdata=df.time, type="terms")[,"te(tend,karno)"]
time_plot <- ggplot(df.time, aes(x=karno, y=fte)) +
  geom_line(aes(col=factor(tend))) +
  scale_color_brewer(name="t", palette="Set1") +
  ylab(expression(f(list(x[plain(karno)],t)))) +
  xlab(expression(x[plain(karno)])) +
  theme(legend.position="bottom")
```
</details>

```{r, fig.width=7, fig.height=3, warning=FALSE}
grid.arrange(te_plot, karno_plot, time_plot, nrow=1)
```

Note that we have to be somewhat cautious with interpretation, considering 
the uncertainty of the estimation, especially for lower Karnofsky-Scores
and later time-points: 

```{r, fig.width=7, fig.height=3, warning=FALSE}
gg_tensor(pam3, ci=TRUE)
```

