---
title: "Defining a new backend: gradient boosting with xgboost"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Defining a new backend: gradient boosting with xgboost}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: Remote.bib
link-citations: yes
---

```{r, echo = FALSE}
library(knitr)
has_xgb <- requireNamespace("xgboost", quietly = TRUE)
opts_chunk$set(
  fig.align  = "center",
  cache      = FALSE,
  message    = FALSE,
  warning    = FALSE,
  fig.height = 3.6,
  fig.width  = 6,
  eval       = has_xgb)
```

```{r, message = FALSE}
library(dplyr)
library(ggplot2)
theme_set(theme_bw())
library(survival)
library(mgcv)
library(xgboost)
library(pammtools)
Set1 <- RColorBrewer::brewer.pal(9, "Set1")
```

## Why a new backend is cheap

The key idea behind PAMMs [@Bender2018a] is a *data transformation*: a
time-to-event data set is split into the **piece-wise exponential data (PED)**
format, and the hazard is then estimated by a **Poisson regression with a
log-interval-length offset**. The default estimation engine in **`pammtools`**
is `mgcv::gam`, but the reduction to Poisson regression means that *any* learner
that can do count/Poisson regression with an offset can serve as a PAMM
backend.

To make a backend a first-class citizen, **`pammtools`** abstracts every model
behind a small internal interface and builds *all* of its post-processing on top
of it:

* `make_X()`, `get_coefs()`, `get_Vp()` — design matrix, coefficients and their
  covariance, such that `make_X %*% get_coefs` is the linear predictor;
* `sample_coefs()` and `sim_hazard()` — drawing from the model's sampling
  distribution, used for simulation-based confidence intervals.

Everything else — `add_hazard()`, `add_cumu_hazard()`, `add_surv_prob()`
(point estimates **and** intervals), and the `geom_*` plotting layers — is
written against this interface. So to teach `xgboost` to act as a PAMM backend
we only need to express it in these terms; the entire workflow then works
unchanged.

As the example learner we use
[**`xgboost`**](https://CRAN.R-project.org/package=xgboost) with its built-in
`objective = "count:poisson"`. Since a boosted tree ensemble has no coefficient
covariance, we obtain uncertainty from a **subject-level bootstrap ensemble**
(bagging): the point estimate is the bagged mean hazard, and confidence bands
are percentiles across the ensemble members.

## The backend

**1. A `ped` becomes an `xgboost` data set.** A `ped` already carries the
bookkeeping we need in its attributes: `intvars` lists the internal interval
columns, `ped_status` is the Poisson response, and `offset` is the
log-interval-length. We keep `tend` as the time *feature*, drop the remaining
internal columns and — the only trick — pass the PED `offset` as `xgboost`'s
`base_margin`, so it plays the role of the Poisson offset.

```{r}
as_xgb_data <- function(ped) {
  features <- c("tend", setdiff(colnames(ped), attr(ped, "intvars")))
  x <- model.matrix(~ . - 1, data = as.data.frame(ped)[, features, drop = FALSE])
  d <- xgb.DMatrix(x, label = ped[["ped_status"]])
  setinfo(d, "base_margin", ped[["offset"]]) # PEM offset = log(interval length)
  d
}
```

**2. Fitting = a bootstrap ensemble of boosters.** We resample subjects (the
`id_var`) with replacement and fit one booster per replicate. Setting
`base_score = 1` together with the offset means that at *prediction* time (where
we supply no `base_margin`) the default margin is `log(base_score) = 0`, so
predictions are the bare hazard `exp(trees)`. We store the ensemble plus the
features it expects; the object is a plain list of class `pam_xgb`.

```{r}
xgb_pam <- function(ped, params = list(), nrounds = 200, nboot = 40, ...) {
  features <- c("tend", setdiff(colnames(ped), attr(ped, "intvars")))
  id_var   <- attr(ped, "id_var")
  X   <- model.matrix(~ . - 1, data = as.data.frame(ped)[, features, drop = FALSE])
  y   <- ped[["ped_status"]]
  off <- ped[["offset"]]
  rows_by_id <- split(seq_len(nrow(ped)), ped[[id_var]])
  params <- c(params, list(objective = "count:poisson", base_score = 1, nthread = 2))
  fit_rows <- function(rows) {
    d <- xgb.DMatrix(X[rows, , drop = FALSE], label = y[rows])
    setinfo(d, "base_margin", off[rows])
    xgb.train(params = params, data = d, nrounds = nrounds, verbose = 0, ...)
  }
  models <- lapply(seq_len(nboot), function(b) {
    fit_rows(unlist(rows_by_id[sample(names(rows_by_id), replace = TRUE)],
                    use.names = FALSE))
  })
  structure(
    list(models = models, orig_features = features,
         trafo_args = attr(ped, "trafo_args")),
    class = "pam_xgb")
}
```

**3. Plugging into the `pammtools` interface.** `pammtools` builds *every*
derived quantity and its simulation-based confidence intervals from two
per-model primitives, so all we implement are these two S3 methods:

* `get_hazard()` returns the **point hazard** as a plain vector (here the bagged
  mean over the ensemble);
* `sim_hazard()` returns a **matrix of hazard draws** (one column per draw). For
  our ensemble a "draw" is a bootstrap member, so we predict the feature matrix
  through every booster.

```{r}
get_hazard.pam_xgb <- function(object, newdata, ...) {
  rowMeans(sim_hazard.pam_xgb(object, newdata))      # bagged point hazard
}

sim_hazard.pam_xgb <- function(object, newdata, nsim = NULL, ...) {
  X  <- model.matrix(~ . - 1,
    data = as.data.frame(newdata)[, object$orig_features, drop = FALSE])
  dm <- xgb.DMatrix(X)
  vapply(object$models, function(m) predict(m, dm), numeric(nrow(X))) # rows x members
}
```

That is the whole backend. `add_hazard()`, `add_cumu_hazard(ci_type = "sim")`
and `add_surv_prob(ci_type = "sim")` now work for `pam_xgb` objects, returning
point estimates and bootstrap intervals, and the `geom_*` layers plot them, all
for free. The very same engine also drives `add_cif()` (competing risks) and
`add_trans_prob()` (multistate) from these two primitives — a backend that keeps
the cause/transition indicator among its features gets those for free too.

## Example 1: recovering a known hazard, with uncertainty

We simulate from a known hazard with `sim_pexp` and transform to PED with
`as_ped`, exactly as for a standard PAMM. Passing `cut` to `as_ped` fixes a
modest interval grid (here 20 intervals), which keeps the boosters fast to fit.

```{r}
set.seed(2024)
n  <- 2000
df <- data.frame(x1 = runif(n, -2, 2), x2 = runif(n, 0, 4))
# true log-hazard: smooth baseline + linear x1 + linear x2
f_true <- function(t, x1, x2) -2.2 + 0.5 * sin(0.8 * t) - 0.4 * x1 + 0.25 * x2
sim_df <- sim_pexp(~ -2.2 + 0.5 * sin(0.8 * t) - 0.4 * x1 + 0.25 * x2, df,
                   cut = seq(0, 6, by = 0.1))
ped <- as_ped(sim_df, Surv(time, status) ~ x1 + x2, cut = seq(0, 6, by = 0.3))
```

We fit the boosted PAMM with `xgb_pam()` and, for comparison, a standard
GAM-based PAMM.

```{r}
xgb_fit <- xgb_pam(ped,
  params  = list(eta = 0.1, max_depth = 3, min_child_weight = 20,
                 subsample = 0.7, colsample_bytree = 0.8),
  nrounds = 200, nboot = 40)

pam <- gam(ped_status ~ s(tend) + x1 + s(x2), data = ped,
  family = poisson(), offset = offset)
```

For prediction we build a grid with `make_newdata` and add the hazard with the
*same* `add_hazard()` call for both models — the boosted one is served by our
`get_hazard.pam_xgb` method, returning bootstrap confidence bands.

```{r}
nd <- ped %>% make_newdata(tend = unique(tend), x1 = c(-1, 1), x2 = c(2))

haz <- bind_rows(
  nd %>% group_by(x1, x2) %>% add_hazard(xgb_fit)         %>% mutate(method = "xgboost"),
  nd %>% group_by(x1, x2) %>% add_hazard(pam, ci = TRUE)  %>% mutate(method = "gam (PAMM)")
) %>%
  mutate(truth = exp(f_true(tend, x1, x2)),
         profile = paste0("x1 = ", x1))

ggplot(haz, aes(x = tend)) +
  geom_stepribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = method), alpha = 0.2) +
  geom_stephazard(aes(y = hazard, col = method)) +
  geom_line(aes(y = truth), lty = 2) +
  facet_wrap(~ profile) +
  scale_color_manual(values = c("xgboost" = Set1[2], "gam (PAMM)" = Set1[1])) +
  scale_fill_manual(values  = c("xgboost" = Set1[2], "gam (PAMM)" = Set1[1])) +
  labs(y = "hazard", col = "", fill = "",
       title = "Estimated hazard (95% CI) vs. truth (dashed)")
```

Both backends recover the baseline shape and the covariate effects; the GAM is
smooth by construction, while the boosted hazard is piece-wise constant. The
bootstrap bands of the boosted fit are comparable to the GAM's Bayesian
intervals.

Survival probabilities come from the same machinery via `add_surv_prob()` with
`ci_type = "sim"` — no manual cumulative sums:

```{r}
surv <- nd %>%
  group_by(x1, x2) %>%
  add_surv_prob(xgb_fit, ci_type = "sim") %>%
  ungroup() %>%
  group_by(x1, x2) %>%
  mutate(truth = exp(-cumsum(exp(f_true(tend, x1, x2)) * (tend - lag(tend, default = 0))))) %>%
  ungroup() %>%
  mutate(profile = paste0("x1 = ", x1))

ggplot(surv, aes(x = tend)) +
  geom_ribbon(aes(ymin = surv_lower, ymax = surv_upper), alpha = 0.2, fill = Set1[2]) +
  geom_line(aes(y = surv_prob), col = Set1[2]) +
  geom_line(aes(y = truth), lty = 2) +
  facet_wrap(~ profile) +
  labs(y = "survival probability",
       title = "Boosted survival probability (95% CI) vs. truth (dashed)")
```

## Example 2: a time-varying effect, recovered automatically

Because `tend` is just another feature, the trees can model `tend`-by-covariate
interactions, i.e. **time-varying effects**, without us specifying them. Here
the effect of `x1` *grows over time*, `β(t) = 0.1 + 0.25 t` — a clear violation
of proportional hazards.

```{r}
set.seed(7)
n   <- 4000
df2 <- data.frame(x1 = runif(n, -1.5, 1.5), x2 = runif(n, -1, 1))
beta_true <- function(t) 0.1 + 0.25 * t
f2_true   <- function(t, x1, x2) -2.0 + 0.3 * sin(0.8 * t) + beta_true(t) * x1 + 0.3 * x2
sim2 <- sim_pexp(~ -2.0 + 0.3 * sin(0.8 * t) + (0.1 + 0.25 * t) * x1 + 0.3 * x2, df2,
                 cut = seq(0, 6, by = 0.1))
ped2 <- as_ped(sim2, Surv(time, status) ~ x1 + x2, cut = seq(0, 6, by = 0.3))

xgb2 <- xgb_pam(ped2,
  params  = list(eta = 0.1, max_depth = 3, min_child_weight = 20,
                 subsample = 0.7, colsample_bytree = 0.8),
  nrounds = 200, nboot = 40)
```

The estimated hazard for `x1 ∈ {-1, 0, 1}` shows the gap between the curves
*widening over time*, tracking the truth:

```{r, fig.width = 6.5}
nd2 <- ped2 %>% make_newdata(tend = unique(tend), x1 = c(-1, 0, 1), x2 = c(0))
haz2 <- nd2 %>%
  group_by(x1, x2) %>%
  add_hazard(xgb2) %>%
  ungroup() %>%
  mutate(truth = exp(f2_true(tend, x1, x2)), x1 = factor(x1))

ggplot(haz2, aes(x = tend, col = x1, fill = x1)) +
  geom_stepribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.2, col = NA) +
  geom_stephazard(aes(y = hazard)) +
  geom_line(aes(y = truth), lty = 2) +
  scale_color_manual(values = Set1[c(1, 3, 2)]) +
  scale_fill_manual(values  = Set1[c(1, 3, 2)]) +
  labs(y = "hazard", col = "x1", fill = "x1",
       title = "Time-varying effect: hazard (95% CI) vs. truth (dashed)")
```

We can read off the estimated time-varying log-hazard ratio
`β(t) = log h(x1 = 1) / h(x1 = 0)` directly from the ensemble (one ratio
trajectory per member), and it recovers the true linear-in-time effect with
honest uncertainty:

```{r}
feat_mat <- function(d) model.matrix(~ . - 1,
  as.data.frame(d)[, xgb2$orig_features, drop = FALSE])
grid <- ped2 %>% make_newdata(tend = unique(tend), x1 = c(0, 1), x2 = c(0))
H1 <- vapply(xgb2$models, function(m) predict(m, xgb.DMatrix(feat_mat(filter(grid, x1 == 1)))),
             numeric(length(unique(grid$tend))))
H0 <- vapply(xgb2$models, function(m) predict(m, xgb.DMatrix(feat_mat(filter(grid, x1 == 0)))),
             numeric(length(unique(grid$tend))))
beta_hat <- data.frame(
  tend  = sort(unique(grid$tend)),
  est   = rowMeans(log(H1 / H0)),
  lower = apply(log(H1 / H0), 1, quantile, 0.025, type = 6),
  upper = apply(log(H1 / H0), 1, quantile, 0.975, type = 6))

ggplot(beta_hat, aes(x = tend)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, fill = Set1[2]) +
  geom_line(aes(y = est), col = Set1[2]) +
  geom_line(aes(y = beta_true(tend)), lty = 2) +
  labs(y = expression(beta(t)),
       title = expression("Estimated time-varying log-hazard ratio " * beta(t) * " (95% CI) vs. truth"))
```

## Wrap-up

A usable `xgboost` backend for PAMMs took four short functions: two helpers
(`as_xgb_data`, `xgb_pam`) and two S3 methods (`get_hazard.pam_xgb`,
`sim_hazard.pam_xgb`) expressed against
**`pammtools`**' model interface. Everything else — `make_newdata`,
`add_hazard`, `add_cumu_hazard`, `add_surv_prob`, the `geom_*` layers and the
bootstrap confidence intervals — came for free.

A caveat on inference: the bootstrap bands capture the **sampling variability**
of the ensemble. Like for any flexible learner, they do not reflect the
approximation **bias** of the trees, so they can be optimistic in regions where
the ensemble is systematically off — visible here at the end of follow-up, where
data are sparse and the estimated time-varying effect `β(t)` lags behind the
truth. The same recipe extends to tuning/cross-validation and to competing risks
(stack cause-specific PEDs and use the same offset trick), with no changes to
**`pammtools`** itself.
