---
title: "Defining a new backend: gradient boosting with xgboost"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Defining a new backend: gradient boosting with xgboost}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: Remote.bib
link-citations: yes
---

```{r, echo = FALSE}
library(knitr)
has_xgb <- requireNamespace("xgboost", quietly = TRUE)
opts_chunk$set(
  fig.align  = "center",
  cache      = FALSE,
  message    = FALSE,
  warning    = FALSE,
  fig.height = 4,
  fig.width  = 5,
  eval       = has_xgb)
```

```{r, message = FALSE}
library(dplyr)
library(ggplot2)
theme_set(theme_bw())
library(survival)
library(mgcv)
library(xgboost)
library(pammtools)
Set1 <- RColorBrewer::brewer.pal(9, "Set1")
```

## Why a new backend is cheap

The key idea behind PAMMs [@Bender2018a] is a *data transformation*: a
time-to-event data set is split into the **piece-wise exponential data (PED)**
format, and the hazard is then estimated by a **Poisson regression with a
log-interval-length offset**. In **`pammtools`** the standard estimation
engine is `mgcv::gam`, but the reduction to Poisson regression means that
*any* learner that can do count/Poisson regression with an offset can serve as
a PAMM backend.

This vignette shows that wiring up such an alternative backend takes only a
handful of lines, because all of the heavy lifting — data simulation
(`sim_pexp`), the survival-to-PED transformation (`as_ped`), prediction-grid
construction (`make_newdata`), and the post-processing/plotting machinery
(`add_hazard`, `geom_stephazard`, `geom_surv`, ...) — is reused unchanged. As
the example backend we use [**`xgboost`**](https://CRAN.R-project.org/package=xgboost)
with its built-in `objective = "count:poisson"`.

## The glue code

Everything needed to turn `xgboost` into a PAMM backend fits in three small
functions.

**1. Transform a `ped` object into `xgboost`'s data structure.** A `ped`
already carries all the bookkeeping we need in its attributes: `intvars` lists
the internal interval columns (`id`, `tstart`, `tend`, `interval`, `offset`,
`ped_status`), `ped_status` is the (Poisson) response, and `offset` is the
log-interval-length. We keep `tend` as the time *feature*, drop the remaining
internal columns, and — this is the only trick — pass the PED `offset` as
`xgboost`'s `base_margin` so it plays the role of the Poisson offset.

```{r}
as_xgb_data <- function(ped) {
  # features = interval end time + all original covariates
  features <- c("tend", setdiff(colnames(ped), attr(ped, "intvars")))
  x <- model.matrix(~ . - 1, data = as.data.frame(ped)[, features, drop = FALSE])
  d <- xgb.DMatrix(x, label = ped[["ped_status"]])
  setinfo(d, "base_margin", ped[["offset"]]) # the PEM offset = log(interval length)
  d
}
```

**2. Fit the model.** We simply call `xgb.train()` with the Poisson objective.
Setting `base_score = 1` together with the offset above means that at *prediction*
time (where we will *not* supply a `base_margin`) the default margin is
`log(base_score) = 0`, so predictions are the bare hazard `exp(trees)`. We
remember which features the model expects and the transformation arguments of
the training data, then tag the object with the class `pam_xgb`.

```{r}
xgb_pam <- function(ped, params = list(), nrounds = 500, ...) {
  features <- c("tend", setdiff(colnames(ped), attr(ped, "intvars")))
  fit <- xgb.train(
    params  = c(params, list(objective = "count:poisson", base_score = 1)),
    data    = as_xgb_data(ped),
    nrounds = nrounds,
    ...)
  attr(fit, "orig_features") <- features
  attr(fit, "trafo_args")    <- attr(ped, "trafo_args")
  class(fit) <- c("pam_xgb", class(fit))
  fit
}
```

**3. Hook into the post-processing machinery.** `pammtools` computes hazards
through the S3 generic `get_hazard()`, and `add_hazard()` dispatches to it.
Providing a single `get_hazard` method for our new class is therefore enough to
make `add_hazard()` work for boosted fits — no changes to the package required.

```{r}
get_hazard.pam_xgb <- function(object, newdata, ...) {
  X <- model.matrix(~ . - 1,
    data = newdata[, attr(object, "orig_features"), drop = FALSE])
  newdata[["hazard"]] <- predict(object, xgb.DMatrix(X))
  newdata
}
```

That's the whole backend. Cumulative hazard and survival probability are not
estimated by a separate model but follow directly from the hazard via the
piece-wise-exponential identities
$\Lambda(t) = \sum_j \lambda_j\,\Delta_j$ and $S(t) = \exp(-\Lambda(t))$,
i.e. a grouped cumulative sum over the interval lengths — we use that below.

## Example 1: recovering a known hazard

We simulate survival times from a known hazard with `sim_pexp` and transform to
PED with `as_ped`, exactly as for a standard PAMM.

```{r}
set.seed(2024)
n  <- 750
df <- data.frame(x1 = runif(n, -3, 3), x2 = runif(n, 0, 6))
f0 <- function(t) dgamma(t, 8, 2) * 6
# true log-hazard: smooth baseline + linear x1 + sqrt(x2)
sim_df <- sim_pexp(~ -3.5 + f0(t) - 0.5 * x1 + sqrt(x2), df, cut = seq(0, 10, by = 0.2))
ped    <- as_ped(sim_df, Surv(time, status) ~ x1 + x2)
```

We fit the boosted PAMM with our new `xgb_pam()` and, for comparison, a standard
GAM-based PAMM with `mgcv::gam`.

```{r}
xgb_fit <- xgb_pam(ped,
  params  = list(eta = 0.05, max_depth = 2, min_child_weight = 10,
                 subsample = 0.8, colsample_bytree = 0.8),
  nrounds = 300, verbose = 0)

pam <- gam(ped_status ~ s(tend) + x1 + s(x2), data = ped,
  family = poisson(), offset = offset)
```

For prediction we build a grid with `make_newdata` (a single covariate profile,
all interval end points) and add the hazard with the *same* `add_hazard()` call
for both models — the boosted one is served by our `get_hazard.pam_xgb` method.

```{r}
nd <- ped %>% make_newdata(tend = unique(tend), x1 = c(0), x2 = c(3))

haz <- bind_rows(
  nd %>% add_hazard(xgb_fit)          %>% transmute(tend, hazard, method = "xgboost"),
  nd %>% add_hazard(pam, ci = FALSE)  %>% transmute(tend, hazard, method = "gam (PAMM)"),
  nd %>% transmute(tend, method = "truth",
                   hazard = exp(-3.5 + f0(tend) - 0.5 * x1 + sqrt(x2)))
)

ggplot(haz, aes(x = tend, y = hazard, col = method)) +
  geom_stephazard() +
  scale_color_manual(values = c("xgboost" = Set1[2], "gam (PAMM)" = Set1[1],
                                "truth" = "black")) +
  labs(title = "Estimated vs. true hazard", col = "")
```

Survival probabilities follow from the hazard by the cumulative-sum identity
above (`geom_surv` automatically anchors each curve at $S(0) = 1$):

```{r}
surv <- haz %>%
  group_by(method) %>%
  arrange(tend) %>%
  mutate(
    intlen    = tend - lag(tend, default = 0),
    surv_prob = exp(-cumsum(hazard * intlen))) %>%
  ungroup()

ggplot(surv, aes(x = tend, y = surv_prob, col = method)) +
  geom_surv() +
  scale_color_manual(values = c("xgboost" = Set1[2], "gam (PAMM)" = Set1[1],
                                "truth" = "black")) +
  labs(title = "Estimated vs. true survival probability", col = "")
```

Both backends recover the baseline shape and the covariate effects; the GAM is
smoother by construction, while the tree ensemble produces a piece-wise constant
hazard.

## Example 2: interactions for free

A practical advantage of a tree-based backend is that it picks up
interactions automatically, without us having to specify them. Here the true
log-hazard contains a strong `x1 * x2` interaction. We fit the boosted PAMM
(features `x1`, `x2`) against a *purely additive* GAM-PAMM `s(x1) + s(x2)` that
cannot represent it.

```{r}
set.seed(7)
n   <- 1000
df2 <- data.frame(x1 = runif(n, -3, 3), x2 = runif(n, -3, 3))
lhaz <- function(t, x1, x2) -2.5 + dgamma(t, 8, 2) * 4 + 0.8 * x1 * x2
sim2 <- sim_pexp(~ -2.5 + dgamma(t, 8, 2) * 4 + 0.8 * x1 * x2, df2,
                 cut = seq(0, 10, by = 0.25))
ped2 <- as_ped(sim2, Surv(time, status) ~ x1 + x2)

xgb2 <- xgb_pam(ped2,
  params  = list(eta = 0.05, max_depth = 3, min_child_weight = 10,
                 subsample = 0.8, colsample_bytree = 0.8),
  nrounds = 400, verbose = 0)
gam_add <- gam(ped_status ~ s(tend) + s(x1) + s(x2), data = ped2,
  family = poisson(), offset = offset) # additive: no interaction
```

We compare predicted survival for two covariate profiles that differ only in the
sign of `x1` (and hence in the sign of the interaction):

```{r, fig.width = 6.5, fig.height = 3.6}
add_surv <- function(d) {
  d %>% group_by(x1, x2) %>% arrange(tend) %>%
    mutate(intlen = tend - lag(tend, default = 0),
           surv_prob = exp(-cumsum(hazard * intlen))) %>%
    ungroup()
}
nd2 <- ped2 %>% make_newdata(tend = unique(tend), x1 = c(-2, 2), x2 = c(2))

surv2 <- bind_rows(
  nd2 %>% add_hazard(xgb2)             %>% add_surv() %>% mutate(method = "xgboost"),
  nd2 %>% add_hazard(gam_add, ci = FALSE) %>% add_surv() %>% mutate(method = "gam (additive)"),
  nd2 %>% mutate(hazard = exp(lhaz(tend, x1, x2)))      %>% add_surv() %>% mutate(method = "truth"))

ggplot(surv2, aes(x = tend, y = surv_prob, col = method)) +
  geom_surv() +
  facet_wrap(~ paste0("x1 = ", x1, ", x2 = ", x2)) +
  scale_color_manual(values = c("xgboost" = Set1[2], "gam (additive)" = Set1[1],
                                "truth" = "black")) +
  labs(title = "Survival under an x1:x2 interaction", col = "")
```

The boosted PAMM tracks the truth across both profiles, whereas the additive
GAM-PAMM — lacking the interaction term — predicts essentially the same survival
regardless of the sign of `x1` and is badly off in both panels.

## Wrap-up

Defining a usable `xgboost` backend for PAMMs took three short functions
(`as_xgb_data`, `xgb_pam`, `get_hazard.pam_xgb`) on top of the existing
**`pammtools`** infrastructure. The same recipe — *transform to PED, fit a
Poisson learner with the interval-length offset, predict the hazard, derive
$\Lambda(t)$/$S(t)$ by cumulative sums* — extends naturally to:

  - **tuning / cross-validation**, using `xgboost`'s own `xgb.cv` or any resampling
    on the original (subject-level) data before transformation;
  - **competing risks**, by stacking cause-specific PEDs (`as_ped(...,
    Surv(time, status) ~ ., combine = ...)`) and using the same offset trick per
    cause;
  - **other learners** (random forests, neural networks, ...) — anything that can
    fit a Poisson/count model with an offset.

These extensions are left out here to keep the example short, but require no
changes to **`pammtools`** itself.
